### 2014-05-23 PAPER/rough_draft.tex

\documentclass[11pt]{article} % use larger type; default would be 10pt
\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% BEGIN Article customizations

%%% PAGE LAYOUT
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=1in} % for example, change the margins to 2 inches all round
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% FIGURES
\usepackage{graphicx} % support the \includegraphics command and options
\graphicspath{{../FIGURES/FIGURE_PDFS/}}

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Double Sequencing Analysis (this still needs a title)}
\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite}

\begin{document}
\maketitle

\section{Introduction}

Intro goes here.

\section{Results}

\subsection{Unfiltered data: WGA v. WGS}

As expected, we found that

% \begin{figure}
\includegraphics[scale=0.5]{unfiltered_overlap_WGS_WGA_together_densities.png}
% \end{figure}

\subsection{Unfiltered data: Time points}

\subsection{Filtering}

\section{Methods}

\subsection{Data and back-end processing}

All sequence data came from The Cancer Genome Atlas (TCGA) Gliobastoma multiforme (GBM) data set. We downloaded raw reads in (XXX format) using (CGHub?) on (date) for 68 patients. For each patient, data consisted of one set of rads taken from blood DNA, and two sets of reads taken from tumor DNA. In 55 cases, the two sets of reads from tumor DNA were one set of reads from whole genome sequencing (WGS) and one set of reads from whole genome sequencing with amplification (WGA). In 13 cases, the two sets of reads from tumor DNA were one set of reads pre-readiation treatment and one set post-radiation treatment. We developed a pipeline to align all reads to HG19. This pipeline used (bowtie?) for alignment and (bedtools? whatever -- quality control...) We used SomaticSniper to align each tumor sequence with it's corresponding blood sequence. We then filtered the SomaticSniper data using 8 custom filters. These removed (i) , (ii) , (iii) , (iv) , (v) , (vi) , (vii) , and (viii) . All custom code is available in a github repository, located at (webaddress). In addition to sequence data, filter 6 uses dbSNP, version 137, to find common SNPs among the putative mutaitons called by SomaticSniper.

\subsection{Analysis}

We used custom python scrips, available in the above github repository, to perform simple calculations and data operations. We used R to do statistics and generate figures, and this code is also availabel in the git repository.

\section{Discussion}

What do I think of this?

\end{document}

### 2014-05-23 PAPER/rough_draft.tex
more writing on paper...

@@
 
-As expected, we found that
+For 55 samples, TCGA has two identical data sets except for the sample preparation: one is whole genome sequencing (WGS) and the other is whole genome amplified sequencing (WGA).
+We calculated the number of mutations in each of these 110 samples (2 each from 55 patients).
+The number of mutations per sample ranged considerably, from XX to XX in the WGS samples and from XX to XX in the WGA samples.
+That the range is higher among the WGA samples is expected, because the amplification process may introduces mutations thereby artificially inflating the number of mutations in the dataset.
+To test this, we plotted the number of putative SNPs in the WGS and WGA sample preparation for each individual patient.
+As expected, for most patients, there were more mutations in the WGA sample than in the WGS sample (Figure 1).
+Surprisingly, not all exceptions had night numbers of mutations.
+One WGA sample with fewer than 50 mutaions had nearly 500 in the corresponding WGS sample; and one WGS sample with more than 200 mutations had only 200 in the corresponding WGA sample.
 
-% \begin{figure}
+\begin{figure}
+\includegraphics[scale=0.5]{C282_v_C484.png}
+\caption{This figure plots the number of putative SNPs called by SomaticSniper (unfiltered) using the WGS v.
+WGA.
+Each point is a patient.
+The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\end{figure}
+
+We next looked at the overlap between the called putative SNPs in the WGS and WGA samples.
+Although amplification may introduce new mutations, the mutations found in the WGS sample for each patient should be (mostly) a subset of those found in the WGA sample for that same patient.
+Although WGS may itself introduce some new mutations, the percent of WGS mutations that overlap with those found in the WGA sample should be significantly higher than the percent of WGA mutations found in the overlap with WGS.
+To test this, we calculated the overlap between the WGS and WGA samples for each patient.
+As expected, WGS samples have fewer non-overlapping putative SNPs than WGA samples (Figure 2).
+
+\begin{figure}
 \includegraphics[scale=0.5]{unfiltered_overlap_WGS_WGA_together_densities.png}
-% \end{figure}
+\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other from teh same patient.
+The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
+}
+\end{figure}
+
+Although most samples agree with the hypotheses in figures one and two, there are outlyers:
+
+\begin{figure}
+\includegraphics[scale=0.5]{unfiltered_total_muts_v_percent_overlap.png}
+\caption{This is figure 3}
+\end{figure}
 
@@
 
-We used custom python scrips, available in the above github repository, to perform simple calculations and data operations.
+We used custom python scripts, available in the above github repository, to perform simple calculations and data operations.
 We used R to do statistics and generate figures, and this code is also availabel in the git repository.

### 2014-05-23 FIGURES/FIGURE_SCRIPTS/histograms_of_mut_frequency.r

@@ -19,8 +19,10 @@ dev.off()
 
 # plot of (paired) C282 against C484, with a line at 1 to see how things fall
 # anything below the line is in line with the hypothesis, above is bad
-pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
-plot(data$C282_count, data$C484_count, log='xy')
+png( "../FIGURE_PDFS/C282_v_C484.png" )
+#pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
+par(bty='n')
+plot(data$C282_count, data$C484_count, log='xy', pch=20, xlab="No. of putative SNPs, WGA", ylab="No. of putative SNPs, WGS" )
 abline(0,1)
 dev.off()
 

### 2014-05-23 FIGURES/FIGURE_SCRIPTS/histograms_of_unfiltered_overlap.R

@@ -17,9 +17,11 @@ plot( difference, main="Density of [WGS/WGA difference]/[Total WGA mutations]", 
 dev.off()
 
 # plot of the size of the overlap v. total # unfiltered mutations
-pdf( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.pdf" )
+png( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.png" )
+#pdf( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.pdf" )
+# when necessary, use the following title: No. of putative SNPs does not correlate with % WGS/WGA overlap
 par(bty = 'n')
-plot( data$TOTAL, data$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="No. of putative SNPs does not correlate with % WGS/WGA overlap", xlab="log of number of putative SNPs (unfiltered)", ylab="percent overlap between WGS and WGA (unfiltered)" )
+plot( data$TOTAL, data$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="log of no. of putative SNPs (unfiltered)", ylab="percent overlap between WGS and WGA (unfiltered)" )
 abline( 0,1 )
 dev.off()
 

### 2014-05-23 PAPER/rough_draft.tex
more writing

@@
 
-\title{Double Sequencing Analysis (this still needs a title)}
+\title{Needs a title...}
 \author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite}
@@
 \includegraphics[scale=0.5]{unfiltered_overlap_WGS_WGA_together_densities.png}
-\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other from teh same patient.
+\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
@@
 
-Although most samples agree with the hypotheses in figures one and two, there are outlyers:
+As seen in figure one, most WGS samples have around 500 putative mutaitons, and most WGA samples have around 1000, but there are a number of exceptions.
+We hypothesized that these samples, some containing as many as 10,000 mutations, were mostly corrupted data.
+To test this, we plotted the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample (Figure 3).
+We expected a negative correlation, which would indicate that as the quality of the sample worsened, the number of mutations increased.
+Instead, we got a completely random distribution.
+This could indicate...
+(still trying to decide what I think this indicates)
 
@@
 \includegraphics[scale=0.5]{unfiltered_total_muts_v_percent_overlap.png}
-\caption{This is figure 3}
+\caption{Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
 \end{figure}

### 2014-06-01 PAPER/rough_draft.tex
added title and moved figures to the end

@@
 
-\title{Needs a title...}
+\title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
 \author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite}
@@
 
-\begin{figure}
-\includegraphics[scale=0.5]{C282_v_C484.png}
-\caption{This figure plots the number of putative SNPs called by SomaticSniper (unfiltered) using the WGS v.
-WGA.
-Each point is a patient.
-The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
-\end{figure}
-
 We next looked at the overlap between the called putative SNPs in the WGS and WGA samples.
@@
 As expected, WGS samples have fewer non-overlapping putative SNPs than WGA samples (Figure 2).
-
-\begin{figure}
-\includegraphics[scale=0.5]{unfiltered_overlap_WGS_WGA_together_densities.png}
-\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
-The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
-}
-\end{figure}
 
@@
 (still trying to decide what I think this indicates)
-
-\begin{figure}
-\includegraphics[scale=0.5]{unfiltered_total_muts_v_percent_overlap.png}
-\caption{Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
-\end{figure}
 
@@
 
+\section{Figures}
+
+\begin{figure}
+\includegraphics[scale=1.0]{C282_v_C484.png}
+\caption{This figure plots the number of putative SNPs called by SomaticSniper (unfiltered) using the WGS v.
+WGA.
+Each point is a patient.
+The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=1.0]{unfiltered_overlap_WGS_WGA_together_densities.png}
+\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
+The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
+}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=1.0]{unfiltered_total_muts_v_percent_overlap.png}
+\caption{Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
+\end{figure}
+
 \end{document}

### 2014-06-06 PAPER/rough_draft.tex
added a bibliography to this paper

@@
 
-All sequence data came from The Cancer Genome Atlas (TCGA) Gliobastoma multiforme (GBM) data set.
-We downloaded raw reads in (XXX format) using (CGHub?) on (date) for 68 patients.
-For each patient, data consisted of one set of rads taken from blood DNA, and two sets of reads taken from tumor DNA.
+All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Gliobastoma multiforme (GBM) data set \cite{TCGA-GBM}.
+We downloaded raw reads in .bam files from 68 patients using CGHub \cite{CGHub}.
+For each patient, data consisted of one set of reads taken from blood DNA, and two sets of reads taken from tumor DNA.
 In 55 cases, the two sets of reads from tumor DNA were one set of reads from whole genome sequencing (WGS) and one set of reads from whole genome sequencing with amplification (WGA).
@@
 We developed a pipeline to align all reads to HG19.
-This pipeline used (bowtie?) for alignment and (bedtools?
-whatever -- quality control...) We used SomaticSniper to align each tumor sequence with it's corresponding blood sequence.
-We then filtered the SomaticSniper data using 8 custom filters.
+This pipeline used picard, BWA for alignment, SamTools, GATK, and custom python code.
+We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 8 custom filters.
 These removed (i) , (ii) , (iii) , (iv) , (v) , (vi) , (vii) , and (viii) .
@@
 We used custom python scripts, available in the above github repository, to perform simple calculations and data operations.
-We used R to do statistics and generate figures, and this code is also availabel in the git repository.
+We used R \cite{Rsoftware} to do statistics and generate figures, and this code is also available in the git repository.
 
@@
 
+\bibliography{TCGA_bib}
+\bibliographystyle{plain}
+
 \end{document}

### 2014-06-07 PAPER/rough_draft.tex
still adding citations

@@
 We developed a pipeline to align all reads to HG19.
-This pipeline used picard, BWA for alignment, SamTools, GATK, and custom python code.
-We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 8 custom filters.
-These removed (i) , (ii) , (iii) , (iv) , (v) , (vi) , (vii) , and (viii) .
-All custom code is available in a github repository, located at (webaddress).
-In addition to sequence data, filter 6 uses dbSNP, version 137, to find common SNPs among the putative mutaitons called by SomaticSniper.
+This pipeline used picard \cite{picard}, BWA \cite{BWA} for alignment, SamTools \cite{SAMtools}, GATK \cite{GATK}, and custom python code.
+We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 6 custom filters.
+These removed putative SNPs (i) with a SomaticScore less than 40, (ii) within 10 bp from a high-quality indel, (iii) within 10 bp of another putative SNP, (iv) with a variant allele quality less than 20, (v) with a dbSNP \cite{dbSNP} identifier and fewer than 10x coverage, and (vi) that were loss of heterozygosity.
+All custom code is available in a github repository, and is available upon request.
 
@@
 We used custom python scripts, available in the above github repository, to perform simple calculations and data operations.
-We used R \cite{Rsoftware} to do statistics and generate figures, and this code is also available in the git repository.
+We used R \cite{Rsoftware} to do statistics and generate figures, and this code is also stored in a github repository and available upon request.
 

### 2014-06-20 PAPER/rough_draft.tex
more paper work

@@
 
+\section{Abstract}
+
+A) What system are you studying?
+B) Why is it important?
+C) What is your research approach?
+D) (optional) How is your research different from previous work?
+E) What are the results of your research?
+F) What conclusions can we draw from these results?
+
 \section{Introduction}
 
-Intro goes here.
+A) What is the general field of study?
+B) What is the specific question of the research?
+C) Why is (a) and/or (b) important?
+D) What has previously been done in this area?
+E) How does your research differ?
+F) (end with) a summary of the research described in this paper
+
+Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor, with a median survival of 13.9 months and a 5-year survival of ~5\% \cite{TCGA-GBM-13}.
+Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemoteraputics.
+Like all cancers, GBM is an evolutionary disease caused by mutations and other alterations to normal glial cell lines.
+However, following substantial research efforts dedicated to identifying all of the mutations in an individual patient's tumor to try and identify causal variants, few precise causal elationships have been identified.
+There are two theories behind this descrepancy: First is the view that "GBM growth is driven by a signaling network with functional redundancy that permits adaptation" \cite{TCGA-GBM-13}.
+Second is the notion that batch effects and sequencing protocols, and thus the data on which genomic analyses are based, is imperfect.
+In reality, both of these views are probably correct.
+In this paper, we use data from doubly-sequnced GBM patients to investigate key attributes of the data and some of the most common computational processing pipelines.
+
+ The largest effort behind this research is that of The Cancer Genome Atlas' Research Network, who first sequenced 206 GBMs in 2008 \cite{TCGA-GBM}.
+In the intervening years, this data set has grown to contain over 500 tumors with multiple types of molecular data, all made publicly available \cite{TCGA-GBM-13}.
+Each sample in the database contains data from a single patient's tumor and blood samples, allowing researchers to directly compare the two, and then compare sets of differences between tumor and "normal" tissue between patients.
+While most of the ~600 sequencing samples in teh TCGA are from different patients, some 63 patients appear in the database twice, 55 having been seuqenced once each with two different DNA preparation methods, and 13 having been sequenced at two different times.
+In this paper,
+
+Since it's publication in 2008, TCGA data has been used by researchers to investigate a variety of phenomena in GBM.
+(subtyping, drivres, pathways, targets) These successes aside, other results have caused some researchers to question what exactly is measured by the TCGA sequencing data.
+(the apper that showed all four subtypes in one tumor) (the more recent over time paper) (maybe another about heterogeneity) Sample preparation and sequencing methods, as well as computational analytical methods, indubitably influence sequencing results.
+In this paper, we begin to ask how.
+
+WGS and WGA are different... We have 55 samples of each... Expectation is X, and we find as follows...
+
+Same thing for the 13 samples...
+
+Everyone must filter, (describe why)... We begin with the one SNP-caller SomaticSniper, and compare the filtered to the unfiltered date... We find as follows...
 
@@
 
-What do I think of this?
+A) Start with a brief (one paragraph or less) summary fo your research, what you did and what you found
+B) Put your results into perspective, discussing (i) what they imply and (ii) how they compare to existing literature
+C) Address potential shortcomings. What could have gone wrong, which parts of the research might be misleading, are there any other caveats?
+D) (optional) Describe implications for future research; are there obvious extensions?
+E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
 

### 2014-06-20 ANALYSIS/filtered_by_filter.py
working on comparing the filters

@@ -1,0 +1,24 @@
+import path
+import sys
+
+# the purpose of this script is to look at what the individual filters are taking out and make comparisons
+
+GBM_genomics = /share/WilkeLab/work/dzd58/TCGA_Reanalysis/GBM_genomics
+data_dir = /share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA
+
+## make sure you're using the correct data
+if sys.argv[ 1 ] = "":
+	data_level = "MUTATION_CALLS_STD40"
+else:
+	data_level = sys.argv[ 1 ] 
+
+data = os.path.join( data_dir, data_level )
+
+##
+
+
+
+## MAIN FUNCTION ##
+##
+## which filtered data set to run this on is sys.argv[ 1 ]
+## but the default is MUTATION_CALLS_STD40

### 2014-06-20 ANALYSIS/overlap_by_chrom.py

@@ -1,1 +1,45 @@
 # the purpose of this script is to calculate the overlap by chromosome of a sample
+# it is possible that I am wasting my time... but if one chromosome is bigger than the rest in terms ov NEVER OVERLAPPING that would be cool
+
+## COMMAND LINE ARGUMENTS ##
+## sys.argv[1] = filtered or not; it is entered as "unfiltered.txt" or "filtered.txt"
+
+import sys
+import os
+
+all_overlaps = {}
+
+# first load the data...
+overlap_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/DOUBLES/WGA-S/MUTATION_CALLS_STD40"
+sample_dirs = [ x[ 0 ] for x in os.walk( overlap_dir ) ]
+for dir in sample_dirs:
+	overlap_file = os.path.join( dir, "overlap_" + sys.argv[ 1 ] )
+	
+	# make sure the file exists...
+	if not os.path.exists( overlap_file ):
+		print "This one, real is not: " + overlap_file
+		continue
+	else:
+		#print "GOOD ONE"
+		overlap_fh = open( overlap_file, 'r' )
+		for line in overlap_fh:
+			line = line.strip( '\r\n' )
+			fields = line.split( '\t' )
+			chrom = fields[ 0 ]
+			loc = fields[ 1 ]
+			if chrom not in all_overlaps:
+				all_overlaps[ chrom ] = [ loc ]
+			else:
+				all_overlaps[ chrom ].append( loc )
+				
+# now start to do some calculations...
+chroms = []
+counts = []
+for chrom in all_overlaps:
+	chroms.append( chrom )
+	counts.append( len( all_overlaps[ chrom ] ) )
+
+print chroms
+print counts
+
+## for a rough estimate, I do still need to normalize this by chromosome length...

### 2014-07-02 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r
changing computers, so updating repo

@@ -2,4 +2,28 @@
 
 # first look at the STD_40 data
 # first, just everything removed, each file individually, how many by each filter 3-8
-data = read.table( "../FIGURE_DATA/histograms_of_filter_removals.r", header=T )
+data = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+
+# make density of each filter removal
+d3 = density( data$FILTER_3 )
+d4 = density( data$FILTER_4 )
+d5 = density( data$FILTER_5 )
+d6 = density( data$FILTER_6 )
+d7 = density( data$FILTER_7 )
+d8 = density( data$FILTER_8 )
+
+# make the box plot that will show everything
+par(bty='n')
+boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n' )
+axis( 1, at=1:6, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
+
+### still working on two problems:
+### how to make the figure horizontally longer so that all labels fit, and 
+### how to move the x-axis lables down relative to the x-axis, again so they fit
+
+
+#axis( 1, at=1:6 )
+#mtext( side=1, text="Variant Allele\nQuality (VAQ)\n>20", line=4, cex=-20 )
+#, "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%"), line=4 )
+
+## this second part of the script makes the same figure, but for only the putative SNPs that are in the overlaps in the unfiltered data

### 2014-07-02 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r
not sure what is going on, apparently have competing changes on two computers

@@ -2,4 +2,4 @@
 
 # first look at the STD_40 data
 # first, just everything removed, each file individually, how many by each filter 3-8
-data = read.table( "../FIGURE_DATA/histograms_of_filter_removals.r", header=T )
+data = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )

### 2014-07-02 ANALYSIS/overlap_only_filter_by_filter.py
added filter_by_filter for the overlap files

@@ -1,0 +1,3 @@
+## this script does the same thing as filter_by_filter.py,
+## which is to see which filters are removing which mutations,
+## but it does so for only the putative mutations in the 55 double samples that are in the unfiltered overlap

### 2014-07-03 ANALYSIS/overlap_only_filter_by_filter.py
finished overlap only which filters script

@@ -1,3 +1,194 @@
 ## this script does the same thing as filter_by_filter.py,
 ## which is to see which filters are removing which mutations,
 ## but it does so for only the putative mutations in the 55 double samples that are in the unfiltered overlap
+
+import os
+import sys
+from sets import Set
+import VCF_list
+
+
+data_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/MUTATION_CALLS"
+overlap_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/OVERLAP"
+fig_data_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/GBM_genomics/FIGURES/FIGURE_DATA"
+
+# this function is taken directly from overlap.py, in this file
+# it checks to make sure the files I want to analyze actually exist
+def check_file_existence( filter_level, double_samples ):
+
+    useable_double_samples = {}
+    num_doub = 0
+
+    for sample in double_samples:
+        existence = []
+        # print double_samples
+        for dir in double_samples[ sample ]:
+            file = os.path.join( dir, filter_level )
+            if os.path.exists( file ):
+                existence.append( "TRUE" )
+        if existence == [ "TRUE", "TRUE" ]:
+            num_doub += 1
+            useable_double_samples[ sample ] = double_samples[ sample ]
+
+    print "there are %g useable samples..." % num_doub
+    
+    # sort so all the 282s (A) come before all the 484s (B)
+    # and in the next set all the XXX (A) come before the XXX (B) 
+    for sample in useable_double_samples:
+        useable_double_samples[ sample ] = sorted( useable_double_samples[ sample ] )
+
+    return useable_double_samples
+
+def make_SNP_dict( dir, filter_level ):
+
+	SNP_dict = {}
+	#SNP_set = Set([])
+
+	# read each line in the file
+	file = os.path.join( dir, filter_level )
+	file_fh = open( file, 'r' )
+	for line in file_fh:
+		line = line.strip( '\r\n' )
+		fields = line.split( '\t' )
+		chrom = fields[ 0 ]
+		loc = fields[ 1 ]
+		
[…]

### 2014-07-03 ANALYSIS/overlap_only_filter_by_filter.py
changed data set used in overlap filter by filter

@@ -8,7 +8,7 @@ from sets import Set
 import VCF_list
 
 
-data_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/MUTATION_CALLS"
+data_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/MUTATION_CALLS_STD40"
 overlap_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/DATA/OVERLAP"
 fig_data_dir = "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/GBM_genomics/FIGURES/FIGURE_DATA"
 

### 2014-07-03 PAPER/rough_draft.tex
finished the box plots for the filter analysis and added them to the paper draft

@@
 
+\begin{figure}
+\includegraphics[scale=1.0]{boxplot_number_filtered.png}
+\caption{Boxplot of the total number of putative SNPs removed by each of six filters, in each of 379 sequencing samples, from 311 patients.
+The x-axis gives the name of those filters that removed putative SNPs, and the y-axis gives, on a log scale, the number of mutations removed by a given sample.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=1.0]{boxplot_percent_overlap_filtered.png}
+\caption{Boxplot of the percentage of putative SNPs making up the overlap between the WGS and WGA sequences of a single sample that was removed by a given filter.
+The x-axis describes the filters, and the y-axis gives the percentage of the overlap removed.}
+\end{figure}
+
 \bibliography{TCGA_bib}

### 2014-07-03 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r

@@ -13,9 +13,17 @@ d7 = density( data$FILTER_7 )
 d8 = density( data$FILTER_8 )
 
 # make the box plot that will show everything
+png( "../FIGURE_PDFS/boxplot_number_filtered.png" )
 par(bty='n')
-boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n' )
-axis( 1, at=1:6, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
+boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
+axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
+dev.off()
+
+pdf( "../FIGURE_PDFS/boxplot_number_filtered.pdf" )
+par(bty='n')
+boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
+axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
+dev.off()
 
 ### still working on two problems:
 ### how to make the figure horizontally longer so that all labels fit, and 
@@ -27,3 +35,19 @@ axis( 1, at=1:6, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH",
[…]

### 2014-08-17 JOB_SCRIPTS/test_bam_to_fastq.sh
updating things...

@@ -3,7 +3,7 @@
 #SBATCH -J t1_3.1
 #SBATCH -o t1_3.1.out
 #SBATCH -e t1_3.1.err
-#SBATCH -n 16
+#SBATCH -n 96
 #SBATCH -p normal
 #SBATCH -A mattcowp
 #SBATCH -t 24:00:00
@@ -14,4 +14,4 @@ echo `hostname`
 
 # Run the executable
 cd /scratch/00001/mattcowp/dakota
-bash $HOME/GBM_genomics/MAKE_VCF/pt_2_only_batchFastExon.sh C484.TCGA-02-2486-01A-01D-1494-08.5 2000 32 4
+bash $HOME/GBM_genomics/MAKE_VCF/12_batchFastExon.sh C484.TCGA-06-0190-01A-01D-1491-08.11 2000000 32 4

### 2014-08-17 MAKE_VCF/12_batchFastExon.sh

@@ -18,15 +18,15 @@ queue="normal"
 
 # 1. .bam to .fastq conversion
 cd /scratch/00001/mattcowp/dakota
-echo $pfx
+# echo $pfx
 cd $pfx
 echo "$pfx:" `date` > ../$pfx.log
 
 # we have to convert a bam into a fastq
 # might have to change path to picard; check this in idev
-echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
-(java -d64 -Xmx4g -jar /opt/apps/picard/1.98/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
-echo "SamToFastQ done: " `date` >> ../$pfx.log
+# echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
+# (java -d64 -Xmx4g -jar /opt/apps/picard/1.98/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
+# echo "SamToFastQ done: " `date` >> ../$pfx.log
 
 # exit: this is the test point for bam to fastq conversion
 # I tested it successfully on sample TCGA-02-2468-01A-01D-1494-08.5
@@ -138,7 +138,7 @@ python /home1/01839/dakotaz/GBM_genomics/MAKE_VCF/launcher.py merge.script $core
 rm -rf pylauncher_tmp*
 echo "Chromosome merge Finished: " `date` >> ../$pfx.log
 
-exit 
+# exit, not testing
 
 # HAVEN'T YET LOOKED AT CODE BEYOND THIS POINT!!!!!!!!! (Keep looking...)
 

### 2014-08-17 FIGURES/FIGURE_SCRIPTS/LOH_VAQ_analysis.r
figure update

@@ -1,0 +1,2 @@
+# this is the same as the "histograms_of_unfiltered_data"
+# but for the LOH/VAQ datasets

### 2014-08-18 MAKE_VCF/PARALLEL/pt_2_only_batchFastExon.sh
moving things around

@@ -1,0 +1,167 @@
+# this script begins only after the .fastq file (step 1 of the original batchFastExon) has been made.
+
+#!/bin/bash
+
+module load bwa/0.7.7
+module load samtools
+module load picard
+module load python
+module load pylauncher
+module load gatk/2.5.2
+
+# these are command line arguments; pfx is the sample name (the TCGA barcode; everything before .bam or .fastq)
+pfx=$1
+splitSize=$2
+batchSize=$3
+coresPerRun=$4
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+queue="normal"
+
+# 2. split and clean up the fastq files
+# which is done by the script split_fastq_threaded.py
+rm split.*
+echo "" >> ../$pfx.log
+echo "python $HOME/GBM_genomics/MAKE_VCF/split_fastq_threaded.py $pfx $splitSize" > split.script
+echo "Waiting while splitting and cleaning the fastq file."  `date` >> ../$pfx.log
+python $HOME/GBM_genomics/MAKE_VCF/launcher.py split.script 2
+echo "Fastq split finished :" `date` >> ../$pfx.log
+rm -rf pylauncher_tmp*
+
+# I guess we're making a note to show that we finished running pylauncher here? 
+echo "-----"
+echo "$pfx:" `date` > $pfx.log
+
+# we need to make sure we are in the right working directory, just cause it's super important
+if [[ "$PWD" != *"$pfx"* ]];
+then
+    echo "Changing directory to /scratch/00001/mattcowp/dakota/$pfx"  `date` >> $pfx.log
+    cd /scratch/00001/mattcowp/dakota/$pfx
+fi;
+
+# exit: working on debugging. Ran the job on this file: TCGA-02-2468-01A-01D-1494-08.6
+# log says good, checked in idev, and for this one, it all checks out
+
+#  check to ensure that we go an equal number of splits, if there
+# are no r1 or r2 files, then try to re-run split script
+numR1Files=`ls r1.* | wc -l`
+echo "Number of r1 files: " $numR1Files  `date` >> ../$pfx.log
+numR2Files=`ls r2.* | wc -l`
[…]

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/SNPs_SomaticSniper.sh
breaking make_VCF into non-parallel steps

@@ -1,0 +1,1 @@
+# this takes a single sample's normal and tumor hg19-aligned bams, and returns a VCF, through the software SomaticSniper

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/align_fastq.sh

@@ -1,0 +1,1 @@
+# the purpose of this script is to align the two fastq files of a TCGA sample (paired ends) to hg19

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/bam_to_fastq.sh

@@ -1,0 +1,1 @@
+# this script takes a TCGA .bam (aligned to hg18) and converts it to a pair of unaligned .fastq files

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/bam_to_fastq.sh
sam to fastq script complete

@@ -1,1 +1,19 @@
 # this script takes a TCGA .bam (aligned to hg18) and converts it to a pair of unaligned .fastq files
+
+module load samtools
+module load picard
+
+# these are command line arguments; pfx is the sample name (the TCGA barcode)
+pfx=$1
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+queue="normal"
+
+# 1. pfx.bam to pfx.fastq conversion
+cd /scratch/00001/mattcowp/dakota
+cd $pfx
+echo "$pfx:" `date` > ../$pfx.log
+
+echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
+(java -d64 -Xmx4g -jar /opt/apps/picard/1.98/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
+echo "SamToFastQ done: " `date` >> ../$pfx.log

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/align_fastq.sh
stopping now; back to finish small scripts later

@@ -1,1 +1,32 @@
+#!/bin/bash
+
 # the purpose of this script is to align the two fastq files of a TCGA sample (paired ends) to hg19
+
+module load bwa
+module load samtools
+
+# pfx is the sample name
+pfx=$1
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+queue="normal"
+
+# move to working directory
+cd /scratch/00001/mattcowp/dakota/$pfx
+
+# bwa aln first fastq
+bwa aln -q 30 -t 4 $refindex $f3fastqfile > $f3fastqfile_prefix.sai 2>$f3fastqfile_prefix.bwa.log
+# bwa aln second fastq
+bwa aln -q 30 -t 4 $refindex $f5fastqfile > $f5fastqfile_prefix.sai 2>$f5fastqfile_prefix.bwa.log
+echo "bwa aln for $f5fastqfile done"
+
+# bwa sampe | samtools view
+(bwa sampe -a 600 -P -r "$RG" $refindex $f3fastqfile_prefix.sai $f5fastqfile_prefix.sai $f3fastqfile $f5fastqfile | samtools view -bSh -o $outprefix.bam -) 2>$outprefix.sampe.log
+# samtools sort
+samtools sort $outprefix.bam $outprefix.sorted >>$outprefix.sampe.log
+# samtools index
+samtools index $outprefix.sorted.bam >>$outprefix.sampe.log
+
+
+
+* rename bams so aligned to hg19

### 2014-08-18 MAKE_VCF/IN_SEQUENCE/bam_to_fastq.sh

@@ -1,3 +1,5 @@
+#!/bin/bash
+
 # this script takes a TCGA .bam (aligned to hg18) and converts it to a pair of unaligned .fastq files
 
 module load samtools
@@ -5,8 +7,6 @@ module load picard
 
 # these are command line arguments; pfx is the sample name (the TCGA barcode)
 pfx=$1
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
 queue="normal"
 
 # 1. pfx.bam to pfx.fastq conversion

### 2014-10-24 PAPER/rough_draft.tex
figure update
paper writing...

--- 2014-06-07 PAPER/rough_draft.tex
@@
 \title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
-\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite}
+\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite} %**ask them about author order, I'm out of it...
 
@@
 
+\section{Abstract}
+
+%A) What system are you studying?
+We examined 55 technical sequencing replicates of Glioblatoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA).
+%Each replicate included the tumor sequenced once using standard whole exome sequencing with amplification, and once without amplification prior to creating a library.
+%B) Why is it important?
+In recent years, TCGA \cite{TCGA-GBM} and others in cancer genomics \cite{Parsons} have used large scale sequencing results of many tumors to identify drivers and essential cancer genes using a variety of computational analyses.
+However, very little is known about the reproducibility of this genomic data and, given the heterogeneity found in tumors, it could be quite low.
+%C) What is your research approach?
+Wemeasured the extent of the overlap between two technical sequencing replicates from the same tumor (n=55), to find what proportion of mutations were reliably found between them, using strictly computational means.
+We further examined whether the raw data was improved or not, as measured by the extent of the overlap, using the SNP calling program SomaticSniper.
+%D) (optional) How is your research different from previous work?
+%E) What are the results of your research?
+We found that in the raw data, about half of the mutations found in one sequencing run can be found in the second.
+Interestingly, the size of the overlap varies by two orders of magnitude, between about 50 and 7000.
+Using a SNP caller, rather than improving the data, removes the overlap completely.
+This is because overlap is almost entirely composed of LOH, removed automatically by SomaticSniper, and SNPs with only a few reads cvering the alternate allele.
+%F) What conclusions can we draw from these results?
+We interpret these results to mean that, as expected, sequencing runs of tumors detect many false positives, and the number of these is increased by an amplification step prior to building a library.
+This tendency aside, the actual number of muations in different tumors of the same type (GBM) may vary by orders of magnitude.
+Finally, while verificaion of sequencing using PCR remains the gold standard, when SNP-calling using strictly computational means, SNP-callers may actually decrease the quality of results.
+
 \section{Introduction}
 
-Intro goes here.
+A) What is the general field of study?
+Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor, with a median survival of 13.9 months and a 5-year survival of ~5\% \cite{TCGA-GBM-13}.
+Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemoteraputics.
+Like all cancers, GBM is an evolutionary disease caused by mutations and other alterations to normal glial cell lines.
+However, following substantial research efforts dedicated to identifying all of the mutations in an individual patient's tumor to try and identify causal variants, few precise causal elationships have been identified.
+There are likely many related causes for this failure, including but not limited to genomic heterogeneity in tumors, \cite{ValentineD}, the view that "GBM growth is driven by a signaling network with functional redundancy that permits adaptation" \cite{TCGA-GBM-13}, and imperfect computational methods.
+%B) What is the specific question of the research?
+In this paper, we are interested in the third.
+We used technical sequencing replicates from 55 GBM tumors to look at the reproducbility of the raw data using a standard computational pipeline.
+Assuming that those mutations that appear in each of the technical replicates are true positives, and all others are false positives (which in fact may or may not be true), we want to see what proportion of mutations recovered are ture positives, and whether or not the proportion is increased by SNP-calling methods designed to remove false positives.
+
+%C) Why is (a) and/or (b) important?
+The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors.
+TCGA's glioblastoma dataset alone includes some 580 (and constantly growing) tumor-normal pairs.
+A varity of researchers in cancer genomics have used this data to discover which genes and pathways are mutated in GBM \cite{pathways}, discover different GBM subtypes \cite{subtypes}, and develop a variety of computational models to find GBM driver mutations \cite{drivers}.
+Despite this apparent success, most of the relationships discovered, though real, are small effects, and few of these innovations have been successfully translated into clinically useful results and techniques.
+Several possible factors likely contribute to this effect.
+%D) What has previously been done in this area?
+[NAME] showed that batch effects have a significant effect on large scale sequencing projects such as TCGA \cite{batch-effects}, and that eliminating this bias without collecting thousands more samples is virtually impossible.
+While TCGA may eventually do this, eliminating these effects at this time is impossible.
+Secondly, multiple groups have shown that tumors are highly genetically heterogeneous, and sampling a tumor only once cannot show the entirely of the mutational profile of the tumor \cite{ValentineD} \cite{hetero1} \cite{hetero2}.
+Third, several groups have shown that epigenetic factors, including methylation \cite{methylation-new}, chromatin remodeling \cite{chromatin}, and micro-RNAs \cite{microRNA}, may significantly effect tumor development and growth.
+%E) How does your research differ?
+For all of these reasons, and likley more besides, we know that tumor sequencing is imperfect, and results based on it have to account for a significant amount of noisy data.
+In this paper, we attempt to look at how much noise (without regard to the theoretical source of it).
+We take 55 TCGA samples with technical replicates, and look at the similarity of the replicates.
+%F) (end with) a summary of the research described in this paper
+We took data from 55 TCGA samples that were sequenced twice, once with the standard WGS protocol, and once with an additional amplification step before library prep, which we call the WGA protocol.
+We aligned the data and used the SNP-calling program SomaticSniper.
+Comparing the WGS and WGA results of these 55 samples, we find significant overlap (around 50\%), but as many differences, between technical sequencing replicates of the same sample.
+As expected, we found that the additional amplification step in the WGA protocol versus the WGS protocol added some mutations to the sample, so that on average these replicates had (i) more mutations, and (ii) a smaller percentage overlap.
+The number of mutations per sample varied by orders of magnitude, from under 100 to almost 10,000.
+We expected to see the quality of results (measured by the percentage overlap between technical replicates) decrease with increasing numbers of mutations, but found no correlation.
+We next looked at standard computational filters meant to increase the quality of the data.
+We found that to the contrary, employing these filters completely removed the overlap between replicates, and that two filters in particular, one for quality of the alternate allele read and one for loss of heterozygosity, removed most of this data.
+We conclude that the number of mutations in GBM tumors may be more variable tthan is generally supposed.
+Further, while nothing will ever beat the biolgical goldstandard of PCR for mutation verification, computational methods need to improve.
 
@@
 
-\subsection{Unfiltered data: WGA v. WGS}
+\subsection{Raw data}
 
-For 55 samples, TCGA has two identical data sets except for the sample preparation: one is whole genome sequencing (WGS) and the other is whole genome amplified sequencing (WGA).
-We calculated the number of mutations in each of these 110 samples (2 each from 55 patients).
-The number of mutations per sample ranged considerably, from XX to XX in the WGS samples and from XX to XX in the WGA samples.
-That the range is higher among the WGA samples is expected, because the amplification process may introduces mutations thereby artificially inflating the number of mutations in the dataset.
-To test this, we plotted the number of putative SNPs in the WGS and WGA sample preparation for each individual patient.
-As expected, for most patients, there were more mutations in the WGA sample than in the WGS sample (Figure 1).
-Surprisingly, not all exceptions had night numbers of mutations.
-One WGA sample with fewer than 50 mutaions had nearly 500 in the corresponding WGS sample; and one WGS sample with more than 200 mutations had only 200 in the corresponding WGA sample.
+% Experiment: size of overlap (figures 1 and 2)
+%(a) a brief (one sentence to one paragraph) summary of what you did and why
+%(b) a description of what you found carrying out this experiment
+For 55 samples, TCGA provides two sets of raw sequence data, one set captured with tthe standard whole exome sequencing methods (WGA) including an amplification of the DNA extracted from the tumor prior to library pred, and one using a protocol for large quantities of DNA that omits this amplification (WGS).
+For each sample, we analyzed each technical replicate using the same pipeline (see Methods), and then compared the WGA replicate to the WGS replicate, to see the extent of the similarity between the two.
+We first looked at the similarity between the two replicates, by number of mutations (Figure 1) and percent overlap partcular mutations (Figure 2) between replicates.
+We found that overall, the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 1).
+We found that the overlap between the two samples, caclulated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was mostly between 20-40\% in WGA replicates and mostly between 40-60\% in WGS replicates.
+As expected the distribution is narrower and taller in the WGS replicates, as a greater percentage of those samples are likley "real" mutations and so present in the overlap (Figure 2).
 
-We next looked at the overlap between the called putative SNPs in the WGS and WGA samples.
-Although amplification may introduce new mutations, the mutations found in the WGS sample for each patient should be (mostly) a subset of those found in the WGA sample for that same patient.
-Although WGS may itself introduce some new mutations, the percent of WGS mutations that overlap with those found in the WGA sample should be significantly higher than the percent of WGA mutations found in the overlap with WGS.
-To test this, we calculated the overlap between the WGS and WGA samples for each patient.
-As expected, WGS samples have fewer non-overlapping putative SNPs than WGA samples (Figure 2).
-
-As seen in figure one, most WGS samples have around 500 putative mutaitons, and most WGA samples have around 1000, but there are a number of exceptions.
-We hypothesized that these samples, some containing as many as 10,000 mutations, were mostly corrupted data.
-To test this, we plotted the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample (Figure 3).
-We expected a negative correlation, which would indicate that as the quality of the sample worsened, the number of mutations increased.
-Instead, we got a completely random distribution.
-This could indicate...
-(still trying to decide what I think this indicates)
-
-\subsection{Unfiltered data: Time points}
+% Experiment: number of mutations (figure 3)
+%(a) a brief (one sentence to one paragraph) summary of what you did and why
+%(b) a description of what you found carrying out this experiment
+Different cancers mutate at different rates; from some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, to tumors with a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}.
+GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this thoery, several samples also had mutation frequencies an order of magnitude greater.
+One possible explanation is a degraded DNA sample, or bad data.
+If this were the case, we would expect the percentage of the overlap between replicates to decrease with the overall number of putative mutations.
+We found no correlation between the number of putatve mutations and the percentage of the sample that was overlapping (Figure 3).
 
 \subsection{Filtering}
+
+% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
+%(a) a brief (one sentence to one paragraph) summary of what you did and why
+%(b) a description of what you found carrying out this experiment
+Large-scale DNA sequencing produces a significant number of errors, both predictable and not \cite{sequence_error}.
+The best way to validate errors is and remains PCR, but in many cases this is not possible, and having a faster, computational method to "validate" mutations signficantly speeds up genomics research.
+To this end, one can employ SNP-calling software, and can filter out mutations with specific characteristics.
+We used the software SomaticSniper \cite{SomaticSniper} and several custom filters to remove from our raw sample putative mutations of with the following traits: (i) a variant allele quality (VAQ), as calculated by SomaticSniper, less than 20, (ii) Loss of heterozygosity (LOH), (iii) within 10 bp of another putative SNP, (iv) within 10 bp of a putative indel, (v) that overlaped wth common human SNPs in dbSNP \cite{dbSNP}, (vi) when the number of reads covering the alternate allele in the tumor sample was less than 10% of the total reads in the tumor sample, (vii) when the Somatic Score (as calculated by SomaticSniper) was less than 40, and (viii) when the quality score (as calculated by GATK) was less than 40.
+Each of these filters is a common way to computationally clean up data.
+We first looked at the number of putative mutations per sample in the filtered data and the overlap between replicates of the filtered data.
+We found only 0-14 putative mutations per sample (average 3) in the filtered data, and a filtered overlap between replicates of 0-2 (average 0).
+(I don't have a figure for this.
+Should I?)
+
+Since using all filters destoryed our signal completely, we next looked at the effects of individual filters.
+Two filters removed nothing.
+Figure 4 shows the number of putative SNPs removed by each of the remaining 6 filters in all 110 samples (55 WGS and 55 WGA) (Figure 4).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations are removed by the VAQ filter and the LOH filter.
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50-100 mutatioins each per sample (Figure 4).
+We next asked how many of the mutations removed by each filter are mutations present in both technical replicates?
+To answer this, we looked at the percent of the overla that was filtered by each of the six filters.
+We found that the three filters removeing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50-100 putative mutations on average (Figure 5).
+In contrast, the VAQ filter (secific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 5).
+Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different filters.
+We
+
+%% this is figures 6 and 7, which need to be added to the document and explained now...
 
@@
 All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Gliobastoma multiforme (GBM) data set \cite{TCGA-GBM}.
-We downloaded raw reads in .bam files from 68 patients using CGHub \cite{CGHub}.
+We downloaded raw reads in .bam files from 55 patients using CGHub \cite{CGHub}.
 For each patient, data consisted of one set of reads taken from blood DNA, and two sets of reads taken from tumor DNA.
-In 55 cases, the two sets of reads from tumor DNA were one set of reads from whole genome sequencing (WGS) and one set of reads from whole genome sequencing with amplification (WGA).
-In 13 cases, the two sets of reads from tumor DNA were one set of reads pre-readiation treatment and one set post-radiation treatment.
+In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library.
+On samples (WGS) did not include this step, and on (WGA) did).
 We developed a pipeline to align all reads to HG19.
-This pipeline used picard \cite{picard}, BWA \cite{BWA} for alignment, SamTools \cite{SAMtools}, GATK \cite{GATK}, and custom python code.
+This pipeline used picard \cite{picard} to generate fastq files from bams aligned to HG18, BWA \cite{BWA} for alignment, SamTools \cite{SAMtools} to clean the data according to accepted standards \cite{seq-standards}, and GATK \cite{GATK} for initial SNP-calling and indel processing.
+These processes were strung together using custom python code.
 We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 6 custom filters.
@@
 
-What do I think of this?
+%A) Start with a brief (one paragraph or less) summary fo your research, what you did and what you found
+We looked at the similarity between technical GBM genomic sequencing replicates for 55 samples in TCGA.
+We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates, and that this was anywhere between 50 and 5,000 putative mutations.
+The number of putative mutations in the WGS and WGA samples did not correlate with the size of the overlap.
+Filtering the raw computational data using both custom filters and those in the software SomaticSniper eliminated more than half of the total number of putative mutations in all 110 samples, as well as the entirely or near entirely of the overlaps between replicates.
+While filters removing overlap with the dbSNP database and mutations within a 10 bp window of other putatitive mutations seemed to preferentially remove mutaitons not in the overlap between replicates, those filters removing putative mutations with low VAQ (calculated by SomaticSniper) and LOH removed between them the majority of the overlap.
+Despite each removing about half of the overlap between replicates, the VAQ and LOH filters appear to be removing different things.
+
+%B) Put your results into perspective, discussing (i) what they imply and (ii) how they compare to existing literature
+GBM, primary brain cancer, is an evolutionary process by which mutaions arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole.
+Treatment for GBM is difficult because complete surgical resection is impossible, and recurrence is a near certainty.
+Even the best treatment, surgical resection followed by readiation and chemoterapy, extends life expectency by less than a year and results in decreases in quality of life for pateints \cite{GBM-treatments}.
+Creation of better therapies is essential, but will require a more complete understanding of the mutational and physiological landscape of GBM.
+
+%C) Address potential shortcomings. What could have gone wrong, which parts of the research might be misleading, are there any other caveats?
+-- using repatability is also problematic 1) cancer is heterogenous; just becuase a mutaiton doesn't show up in the replicate doesn't mean it's not real
+-- using repatability is also problematic 2) could be degraded DNA, so large number of mutations could still be false positive
+-- we may have shown that filtering doesn't do very much good, but we haven't really given alternatives
+-- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
+--
+
+%D) (optional) Describe implications for future research; are there obvious extensions?
+-- other snp callers
+-- biological replicates
+-- our own data?
+
+%E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
 

### 2014-10-24 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r

@@ -46,8 +46,8 @@ axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (V
 dev.off()
 
 # make the same plot for the overlaps
-pdf( "../FIGURE_PDFS/boxplot_percent_overlap_filtered.pdf" )
-par(bty='n')
+pdf( "../FIGURE_PDFS/boxplot_percent_overlap_filtered.pdf", width=6, height=5 )
+par(bty='n', mar=c(5.1,4.1,1.1,0))
 boxplot( overlap_data[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="% of unfiltered overlap filtered, per sample", pch=20 )
 axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
 dev.off()

### 2014-10-28 PAPER/paper_d1.tex
more paper drafts

--- 2014-10-24 PAPER/rough_draft.tex
@@
 
-A) What is the general field of study?
+%A) What is the general field of study?
 Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor, with a median survival of 13.9 months and a 5-year survival of ~5\% \cite{TCGA-GBM-13}.
@@
 GBM, primary brain cancer, is an evolutionary process by which mutaions arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole.
-Treatment for GBM is difficult because complete surgical resection is impossible, and recurrence is a near certainty.
-Even the best treatment, surgical resection followed by readiation and chemoterapy, extends life expectency by less than a year and results in decreases in quality of life for pateints \cite{GBM-treatments}.
-Creation of better therapies is essential, but will require a more complete understanding of the mutational and physiological landscape of GBM.
+Treatment for GBM is difficult and has poor outcomes \cite{GBM-treatments}, but could hopefully be improved by a more complete understanding of the disease.
+Large-scale genomics platforms and big data may be one route towards improved understanding of this devestating condition, but there are significant limitations.
+First (does the data look like the tumor)?
+Second, (PCR is long and expensive to do for every single potential point mutation).
+(We make a first pass at evalutating these limitations, with an eye towards maybe fixing them in the future).
+%Discuss 2-4 concrete elements of the data:
+% - high numer of mutations in unfiltered case (which seems to be real, or at least repeatable)
+The high number of mutations in some, but not all, of the TCGA GBM samples is, at a minimum, repeatable across technical replicates.
+Further, in samples with higher frequency of mutations, technical replicates have as great a percent overlap of individual putative mutations as found in less frequently-mutated samples.
+At a minimum, this suggests the possibility that a higher mutation rate could be a feature of a subset of GBM tumors, and not only a data artifact.
+
+%- effects of various filters
+%- discuss figure of what was removed from one filter vs. the other
+%-- maybe better off with raw data? minus the filters that work
+When comparing the effects of the various standard filters we employed to examine the data, only five filters of eight removed an appreciable number of putative SNPs form the sample.
+Of those five, two (LOH and VAQ) removed primarily mutations that found in both technical replicates, and three preferentially removed putative mutaions found in only one of two replicates.
+Two of the three filters that removed mostly non-reproducible data removed putative mutations that were within 10 bp of other putative mutations; essentially, they recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration.
+A third found dbGAP overlap; essentially, found evidence that while a polymorphism may not be somatic.
+Our analysis suggests that these filters are useful, and do clean up the purely computaitonal data in a meaningful way.
+%- LOH
+By contrast, in may be more useful to discard the two filters that removed principally data from the overlap of the two replicates.
+Of these, one, VAQ is built into SomaticSniper.
+The second removes loss of heterozygosity (LOH) mutations.
+Multiple sources suggest that LOH mutations may be essential to cancer \cite{LOH}.
+This, along with the data presented here, makes a strong arugment for retaining these mutaitons in functional analyses, rather than excluding them.
 
 %C) Address potential shortcomings. What could have gone wrong, which parts of the research might be misleading, are there any other caveats?
--- using repatability is also problematic 1) cancer is heterogenous; just becuase a mutaiton doesn't show up in the replicate doesn't mean it's not real
--- using repatability is also problematic 2) could be degraded DNA, so large number of mutations could still be false positive
--- we may have shown that filtering doesn't do very much good, but we haven't really given alternatives
--- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
---
-
-%D) (optional) Describe implications for future research; are there obvious extensions?
--- other snp callers
--- biological replicates
--- our own data?
+Although the data is suggestive, it is far from conclusive.
+In this analysis, we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
+This metric is potentially problematic for two reasons:
+%-- using repatability is also problematic 1) cancer is heterogenous; just becuase a mutaiton doesn't show up in the replicate doesn't mean it's not real
+(i) First, cancer is highly heterogeneous \cite{hetero}, and so a legitimate somatic SNP might show up in one replicate and not another.
+%-- using repatability is also problematic 2) could be degraded DNA, so large number of mutations could still be false positive
+(ii) Second, if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates.
+Although repeatability does not and could not perfectly measure confidence, it should correlate.
+Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gague the accuracy of other measures (filtering).
+%-- we may have shown that filtering doesn't do very much good, but we haven't really given alternatives
+%-- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
+In this sample, we have shown that while some filters function 
+%-- other snp callers
 
@@
 \includegraphics[scale=1.0]{C282_v_C484.png}
-\caption{This figure plots the number of putative SNPs called by SomaticSniper (unfiltered) using the WGS v.
-WGA.
-Each point is a patient.
+\caption{\textbf{Number of putative SNPs in WGS v.
+WGA, as called by SomaticSniper.} Each point is a patient.
 The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
@@
 \includegraphics[scale=1.0]{unfiltered_overlap_WGS_WGA_together_densities.png}
-\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
+\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
@@
 \includegraphics[scale=1.0]{unfiltered_total_muts_v_percent_overlap.png}
-\caption{Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
+\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
 \end{figure}
@@
 \includegraphics[scale=1.0]{boxplot_number_filtered.png}
-\caption{Boxplot of the total number of putative SNPs removed by each of six filters, in each of 379 sequencing samples, from 311 patients.
-The x-axis gives the name of those filters that removed putative SNPs, and the y-axis gives, on a log scale, the number of mutations removed by a given sample.}
+\caption{\textbf{Number of putative SNPs removed by each of six filters.} The x-axis gives the name of those filters that removed putative SNPs, and the y-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
+Filters removing LOH and low VAQ putative SNPs removed around 100-200 putative SNPs on average.
+Three other filters, removing overlap with dbSNP and putative SNPs close to other putaive mutations, removed on average ~50 mutations per sample.}
 \end{figure}
@@
 \begin{figure}
+% 379 samples, 311 pateints
 \includegraphics[scale=1.0]{boxplot_percent_overlap_filtered.png}
-\caption{Boxplot of the percentage of putative SNPs making up the overlap between the WGS and WGA sequences of a single sample that was removed by a given filter.
-The x-axis describes the filters, and the y-axis gives the percentage of the overlap removed.}
+\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The x-axis describes the filters, and the y-axis gives the percentage of the WGS/WGA overlap removed by the filter, per sample.
+Filters removing LOH and putative SNPs with low VAQ removed a significant ercentage of the overlap; nothing else did.}
+\end{figure}
+
+% this is not the final path name, because this is not the final figure
+\begin{figure}
+\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/LOH_all.png}
+\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by LOH increases.
+This is the opposite of the pattern observed in VAQ mutaitons (figure 6).}
+\end{figure}
+
+% this is not the final path name, because this is not the final figure
+\begin{figure}
+\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/VAQ_all.png}
+\caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by VAQ decreases.
+This is the opposite of the pattern observed in LOH mutaitons (figure 5).}
 \end{figure}

### 2014-10-28 PAPER/paper_d1.tex
fixed figures

@@
 \includegraphics[scale=1.0]{unfiltered_total_muts_v_percent_overlap.png}
-\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
+\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample.}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/LOH_all.png}
+\includegraphics[scale=1.0]{./LOH_VAQ/LOH_all.png}
 \caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by LOH increases.
@@
 \begin{figure}
-\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/VAQ_all.png}
+\includegraphics[scale=1.0]{./LOH_VAQ/VAQ_all.png}
 \caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by VAQ decreases.

### 2014-10-28 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r

@@ -13,16 +13,16 @@ d7 = density( data$FILTER_7 )
 d8 = density( data$FILTER_8 )
 
 # make the box plot that will show everything
-png( "../FIGURE_PDFS/boxplot_number_filtered.png" )
+png( "../FIGURE_PDFS/boxplot_number_filtered.png", width = 480, height = 360, res=55 )
 par(bty='n')
 boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
-axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
+axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Removes Variant\nAllele Quality\n(VAQ)<20", "Removes\nLOH", "Removes putative\nSNPs within\n10 bp window\nof other SNPs", "Removes putative\nSNPs within\n10 bp window\nof indels", "Removes\noverlap\nwith dbSNP\ncoverage", "Removes when\nalt. allele\ncoverage<=10%" ) )
 dev.off()
 
-pdf( "../FIGURE_PDFS/boxplot_number_filtered.pdf" )
-par(bty='n')
+pdf( "../FIGURE_PDFS/boxplot_number_filtered.pdf", width=6, height=5 )
+par(bty='n', mar=c(5.1,4.1,1.1,0))
 boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
-axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Variant Allele\nQuality (VAQ)\n>20", "Removes\nLOH", "No putative\nSNPs within\n10 bp window\nof other SNPs", "No putative\nSNPs within\n10 bp window\nof indels", "No overlap\nwith dbSNP\ncoverage", "Alternate\nallele\ncoverage\n>=10%") )
[…]

### 2014-10-28 FIGURES/FIGURE_SCRIPTS/histograms_of_unfiltered_overlap.R

@@ -34,6 +34,7 @@ amp_frac = density( data$OVERLAP/data$TOTAL_WGA.C282., from=0 )
 seq_frac = density( data$OVERLAP/data$TOTAL_WGS.C484., from=0 )
 par(bty='n')
 plot( seq_frac$x, seq_frac$y, type='l', col='blue', main="", xlab="% of sample obtained in overlap", ylab="Density")
+legend( "topright", c( "WGS", "WGA" ), col=c( "blue", "red" ), lty=1 )
 lines( amp_frac$x, amp_frac$y, col='red' )
 dev.off()
 

### 2014-10-28 PAPER/paper_d1.tex
flowcharts

@@
 %-- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
-In this sample, we have shown that while some filters function 
+In this sample, we have shown that while some filters remove primarily putative SNPs in only one of two replicates (more likely to be false positives), two filters, VAQ and LOH, remove most of the overlap (more likely to be true positives).
+While our results suggest that VAQ and LOH filters are probably removing some true positives, that does not preclude them removing false positives.
+Therefore, if we remove these filters, we may retain more true positives, but we also retain more false positives.
 %-- other snp callers
+Finally, we looked at results from one SNP caller. There are many more equally widely used SNP callers available, which might do better or worse than the one we have chosen here.
 
 %E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
+We have shown that
 
 \section{Figures}
+
+\begin{figure}
+\includegraphics[scale=1.0]{data_processing_flowchart_2.png}
+\caption{\textbf{Number of putative SNPs in WGS v.
+WGA, as called by SomaticSniper.} Each point is a patient.
+The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=1.0]{overlap_explained.png}
+\caption{\textbf{Number of putative SNPs in WGS v.
+WGA, as called by SomaticSniper.} Each point is a patient.
+The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\end{figure}
 

### 2014-10-29 PAPER/paper_d1.tex
working on results

@@
 
-\subsection{Raw data}
+\subsection{Unfiltered data}
 
@@
 %(b) a description of what you found carrying out this experiment
-For 55 samples, TCGA provides two sets of raw sequence data, one set captured with tthe standard whole exome sequencing methods (WGA) including an amplification of the DNA extracted from the tumor prior to library pred, and one using a protocol for large quantities of DNA that omits this amplification (WGS).
+Next-generation sequencing is far from perfect; it produces errors at a rate of around XX per Mbp, and those errors are both predictable and not.
+Cancer sequencing, with highly heterogenous DNA targets from impure tissue samples, is possibly even mroe unreliable.
+Thus, we expect a high number of false positives in tumor sequencing.
+The gold-standard for determining a true somatic mutation from a germ line mutation or a sequencing error is to double check for the variant with PCR, but this is expensive, time consuming with so many samples, and often not available.
+To try and estimate an approximate rate of false positives, we first looked at the degree of overlap between technical replicates in GBM sequencing.
+For 55 GBM tumors, TCGA provides two sets of raw sequence data, one set captured with tthe standard whole exome sequencing methods (WGA) including an amplification of the DNA extracted from the tumor prior to library prep, and one using a protocol for large quantities of DNA that omits this amplification (WGS).
+Our assumption is that putative SNPs appearing in both technical replicates are more likely to be true positives than those appearing only in one replicate.
+
 For each sample, we analyzed each technical replicate using the same pipeline (see Methods), and then compared the WGA replicate to the WGS replicate, to see the extent of the similarity between the two.
@@
 \subsection{Filtering}
+
+\begin{itemize}
+	\item Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration)
+	\item Removes ... I forget
+	\item Removes putative SNPs with SomaticSniper VAQ scores less than or equal to 20 (VAQ)
+	\item Removes putative SNPs that are identified as loss of heterozygosity (LOH)
+	\item Removes putative SNPs located within a 10 bp window of any other putative SNP
+	\item Removes putative SNPs located within a 10 bp window of indels
+	\item Removes putative SNPs that overlap with dbSNP coverage
+	\item Removes putative SNPs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10%
+\end{itemize}
 
@@
 %E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
-We have shown that
+We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50% of putative SNPs in whole exome sequencing and about 30% in whole genome sequencing.
+This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on teh putative mutaions in each technical replicate and the overlap between them, for all 55 samples.
+We found that some filters remove principally those mutaions found in one sample or the other, while other filers remove primarily those in the overlap.
+We suggest using just the first set in the future, when working with strictly computational data.
 
@@
 \includegraphics[scale=1.0]{data_processing_flowchart_2.png}
-\caption{\textbf{Number of putative SNPs in WGS v.
-WGA, as called by SomaticSniper.} Each point is a patient.
-The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\caption{\textbf{Data processing flowchart.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18.
+For each bamfile, we used picard to regerate fastq files, bwa to realign the fastq files to hg19, GATK to recalibrate bases and indels, and SomaticSniper to call somatic mutations.}
 \end{figure}
@@
 \includegraphics[scale=1.0]{overlap_explained.png}
-\caption{\textbf{Number of putative SNPs in WGS v.
-WGA, as called by SomaticSniper.} Each point is a patient.
-The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
+\caption{\textbf{Overlap between WGS and WGA.} For each sample, we generated a SomaticSniper VCF with putative somatic mutations for each of two replicates.
+We took the overlap between the two lists for the most likely true positives.}
 \end{figure}

### 2014-10-29 JOB_SCRIPTS/align_fastq.sh
edited stray punctuation

@@ -16,8 +16,8 @@ data_dir="/scratch/00001/mattcowp/dakota/"
 
 # Run the executable (tumor)
 cd $data_dir/$pfx
-bash $HOME/GBM_genomics/MAKE_VCF/IN_SEQUENCE/align_fastq.sh C484.TCGA-06-0125-02A-11D-2280-08.1
+bash $HOME/GBM_genomics/MAKE_VCF/IN_SEQUENCE/align_fastq.sh C484.TCGA-06-0125-01A-01D-1490-08.6 tumor
 
 # run the executable (normal)
-cd $data_dir/$pfx/normal
-bash $HOME/GBM_genomics/MAKE_VCF/IN_SEQUENCE/align_fastq.sh C484.TCGA-06-0125-10A-01D-1490-08.6
+# cd $data_dir/$pfx/normal
+# bash $HOME/GBM_genomics/MAKE_VCF/IN_SEQUENCE/align_fastq.sh C484.TCGA-06-0125-10A-01D-1490-08.6 blood

### 2014-10-29 JOB_SCRIPTS/dedup_realn_recal.sh

@@ -16,4 +16,4 @@ echo `hostname`
 cd /scratch/00001/mattcowp/dakota
 bash /home1/01839/dakotaz/GBM_genomics/MAKE_VCF/IN_SEQUENCE/SNPs_GATK.sh C484.TCGA-06-0125-02A-11D-2280-08.1 C484.TCGA-06-0125-10A-01D-1490-08.6
 
-bash /home1/01839/dakotaz/GBM_genomics/MAKE_VCF/IN_SEQUENCE/SNPs_SomaticSniper.sh C484.TCGA-06-0125-02A-11D-2280-08.1 C484.TCGA-06-0125-10A-01D-1490-08.6
+#bash /home1/01839/dakotaz/GBM_genomics/MAKE_VCF/IN_SEQUENCE/SNPs_SomaticSniper.sh C484.TCGA-06-0125-02A-11D-2280-08.1 C484.TCGA-06-0125-10A-01D-1490-08.6

### 2014-10-29 MAKE_VCF/IN_SEQUENCE/SNPs_SomaticSniper.sh

@@ -3,27 +3,20 @@
 ## command line variables (this is taken from the previous script, GATK_SNPs.sh)
 tumor_pfx=$1
 blood_pfx=$2
-tumor_sample=${tumor_pfx:0:17}
-blood_sample=${blood_pfx:0:17}
-echo $tumor_pfx $tumor_sample
-echo $blood_pfx $blood_sample
-if [ $tumor_sample != $blood_sample ]; then
-	echo "Non-matching sample names:"
-	echo "	$tumor_sample"
-	echo "	$blood_sample"
-	exit 1;
-fi
 
-tumor_dir="$tumor_pfx\_working"
+## working directory
+tumor_dir="$tumor_pfx.WORK"
 cd /scratch/00001/mattcowp/dakota/$tumor_dir
 
 ## files you may need...
 ref_dir="/work/00001/mattcowp/Hs_reference_datasets"
 hgReference="$ref_dir/Homo_sapiens.GRCh37.72.dna.fa"
+tumor_bam="/scratch/00001/mattcowp/dakota/$tumor_dir/$tumor_pfx.realn.recal.bam"
+normal_bam="/scratch/00001/mattcowp/dakota/$tumor_dir/$blood_pfx.realn.recal.bam"
 
 ## run SomaticSniper on a set of T-N pairs
-echo "SomaticSniper: " 'date' >> ../$tumor_pfx.ss.log
-/scratch/00001/mattcowp/dakota/bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f $hgReference $tumor_pfx.realn.recal.bam $blood_pfx.realn.recal.bam $tumor_pfx.SS.vcf
-echo "SomaticSniper finished: " 'date' >> ../$tumor_pfx.ss.log
+echo "SomaticSniper: " `date` >> ./$tumor_pfx.ss.log
+/scratch/00001/mattcowp/dakota/bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f $hgReference $tumor_bam $normal_bam $tumor_pfx.SS.vcf &>> ./$tumor_pfx.ss.log
+echo "SomaticSniper finished: " `date` >> ./$tumor_pfx.ss.log
 
 exit;

### 2014-10-29 MAKE_VCF/IN_SEQUENCE/align_fastq.sh

@@ -7,27 +7,34 @@ module load samtools
 
 # pfx is the sample name
 pfx=$1
+tissue=$2
 outprefix="$pfx.out"
 queue="normal"
 # reference hg19
 refDir="/work/00001/mattcowp/Hs_reference_datasets"
 hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
 # what data we're running on...
-data_dir="/scratch/00001/mattcowp/dakota/"
+data_dir="/scratch/00001/mattcowp/dakota"
+if [ $tissue = "blood" ]
+then
+	data_dir="/scratch/00001/mattcowp/dakota/normal"
+fi
 fastq1="$data_dir/$pfx/$pfx.1.fastq"
 fastq2="$data_dir/$pfx/$pfx.2.fastq"
 
 # move to working directory
-cd $pfx
+cd $data_dir/$pfx
 
 # the bwa manual says this is important
-bwa index -p RefSeqbwaidx -a bwtsw $hgReference
+# but Matt says take it out...
+# bwa index -p RefSeqbwaidx -a bwtsw $hgReference
 
 # bwa aln first fastq
-bwa aln -q 30 -t 4 $hgReference $fastq1 > $fastq1_prefix.sai 2>$fastq1_prefix.bwa.log
+bwa aln -q 30 -t 8 $hgReference $fastq1 > $fastq1.aln.sai 2>$fastq1.aln.bwa.log &
 # bwa aln second fastq
-bwa aln -q 30 -t 4 $hgReference $fastq2 > $fastq2_prefix.sai 2>$fastq2_prefix.bwa.log
-echo "bwa aln for $f5fastqfile done"
+bwa aln -q 30 -t 8 $hgReference $fastq2 > $fastq2.aln.sai 2>$fastq2.aln.bwa.log &
+wait
+echo "bwa aln for $pfx done"
 
 # need to specify read groups (RG), which is the argument of  -r in bwa sempe
 # first get the library information from the TCGA bam:
@@ -40,8 +47,8 @@ fi
 RG="@RG\tID:1\tPL:ILLUMINA\tSM:$pfx\tLB:$LB\tDS:ref=hg19"
 
 # bwa sampe | samtools view
-(bwa sampe -a 600 -P -r "$RG" $hgReference $fastq1_prefix.sai $fastq2_prefix.sai $fastq1 $fastq2 | samtools view -bSh -o $outprefix.bam -) 2>$outprefix.sampe.log
[…]

### 2014-10-29 PAPER/paper_d1.tex
revised methods, added table

@@
 \usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
+\usepackage{lscape}
 % These packages are all incorporated in the memoir class to one degree or another...
@@
 
-For each sample, we analyzed each technical replicate using the same pipeline (see Methods), and then compared the WGA replicate to the WGS replicate, to see the extent of the similarity between the two.
-We first looked at the similarity between the two replicates, by number of mutations (Figure 1) and percent overlap partcular mutations (Figure 2) between replicates.
-We found that overall, the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 1).
-We found that the overlap between the two samples, caclulated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was mostly between 20-40\% in WGA replicates and mostly between 40-60\% in WGS replicates.
-As expected the distribution is narrower and taller in the WGS replicates, as a greater percentage of those samples are likley "real" mutations and so present in the overlap (Figure 2).
+For each sample, we called mutations in both technical replicates and the pateint's blood sample using the same computational pipeline (see Figure 1 and Methods): we used CGHub to download TCGA bamfiles aligned to hg18, re-generated teh fastq files with picard, re-aligned the fastq files to hg19 with bwa, performed indel re-alignment and base recalibration with GATK, and finally called somatic mutaitons with SomaticSniper.
+We then analyzed the list of putative SNPs generated for each replicate by SomaticSniper, as well as the intersection, or overlap, of the two lists.
+We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3).
+We further found that the overlap between the two samples, caclulated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20-40\% in WGA replicates and mostly between 40-60\% in WGS replicates.
+As expected the distribution is narrower and taller in the WGS replicates, as a greater percentage of those samples are likley true positives and so more likely to be present in the overlap (Figure 4).
 
@@
 If this were the case, we would expect the percentage of the overlap between replicates to decrease with the overall number of putative mutations.
-We found no correlation between the number of putatve mutations and the percentage of the sample that was overlapping (Figure 3).
+We found no correlation between the number of putatve mutations and the percentage of the sample that was overlapping (Figure 5).
+That the the percentage overlap doesn not vary with the number of putative mutaions in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.
 
@@
 
+% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
+%(a) a brief (one sentence to one paragraph) summary of what you did and why
+%(b) a description of what you found carrying out this experiment
+In the normal case lacking sequencing replicates, scientists need other ways to distinguish true and false positives.
+While the generally accepted standard is PCR, several attempts have been made at computational algorithms that can distinguish somatic mutations from germ line mutations and sequencing error.
+These platforms typically identify false positives in two ways, (i) first by probablalistic caclulations based on features of the data, and (ii) by features of the mtuations themselves.
+In this paper, we used GATK for base calibration and indel realignment, and SomaticSniper for somatic mutation calling.
+Each of these platforms has it's own filters of the first type:
+
 \begin{itemize}
 	\item Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration)
-	\item Removes ... I forget
-	\item Removes putative SNPs with SomaticSniper VAQ scores less than or equal to 20 (VAQ)
+	\item Removes putative SNPs with a SomaticScore less than 40 (SS)
+	\item Removes putative SNPs with SomaticSniper Varaint Allele Quality scores less than or equal to 20 (VAQ)
+\end{itemize}
+
+We additionally employed five filters of the second type, all of which have some support in the computational literature:
+
+\begin{itemize}
 	\item Removes putative SNPs that are identified as loss of heterozygosity (LOH)
@@
 
-% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-Large-scale DNA sequencing produces a significant number of errors, both predictable and not \cite{sequence_error}.
-The best way to validate errors is and remains PCR, but in many cases this is not possible, and having a faster, computational method to "validate" mutations signficantly speeds up genomics research.
-To this end, one can employ SNP-calling software, and can filter out mutations with specific characteristics.
-We used the software SomaticSniper \cite{SomaticSniper} and several custom filters to remove from our raw sample putative mutations of with the following traits: (i) a variant allele quality (VAQ), as calculated by SomaticSniper, less than 20, (ii) Loss of heterozygosity (LOH), (iii) within 10 bp of another putative SNP, (iv) within 10 bp of a putative indel, (v) that overlaped wth common human SNPs in dbSNP \cite{dbSNP}, (vi) when the number of reads covering the alternate allele in the tumor sample was less than 10% of the total reads in the tumor sample, (vii) when the Somatic Score (as calculated by SomaticSniper) was less than 40, and (viii) when the quality score (as calculated by GATK) was less than 40.
-Each of these filters is a common way to computationally clean up data.
-We first looked at the number of putative mutations per sample in the filtered data and the overlap between replicates of the filtered data.
-We found only 0-14 putative mutations per sample (average 3) in the filtered data, and a filtered overlap between replicates of 0-2 (average 0).
-(I don't have a figure for this.
-Should I?)
+Our initial question was, does filtering the data in this way increase of decrease the percentage of the sample that is overlapping between the two technical replicates?
+Alternatively stated, does filtering seem to increase the percentage of probable true positives, or not?
+We found that after running all eight filters on the data, the number of putative mutaions per preplicate decreases from 50-10,000 to 0-14 (average 3).
+The size of the overlap drops to 0-2 per sample, with 0 as the mode, so the overall percentage also decreases.
+We conclude that running all the filters on the data, in the absence of any other verification method, is counterproductive, becuase it seems to remove all of the signal (as well as the noise).
 
-Since using all filters destoryed our signal completely, we next looked at the effects of individual filters.
-Two filters removed nothing.
-Figure 4 shows the number of putative SNPs removed by each of the remaining 6 filters in all 110 samples (55 WGS and 55 WGA) (Figure 4).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations are removed by the VAQ filter and the LOH filter.
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50-100 mutatioins each per sample (Figure 4).
-We next asked how many of the mutations removed by each filter are mutations present in both technical replicates?
-To answer this, we looked at the percent of the overla that was filtered by each of the six filters.
-We found that the three filters removeing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50-100 putative mutations on average (Figure 5).
-In contrast, the VAQ filter (secific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 5).
-Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different filters.
-We
+Together, the eight filters we ran drew equally from those putative SNPs found in either replicate and those found in both.
+We next asked if this was true individually as well, or if different filters acted differently.
+We did not analyze data on the first two filers, those removing putative SNPs with GATK quality scores below 40 and Somatic Scores below 40, because those filters are part of the Somatic Sniper platform that calls somatic variants, and so none of these putative SNPs were in the unfiltered sample above.
+We did look at individual data of the other six filters.
+We first looked at the total number of putative SNPs removed by each filter (figure 6).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations are removed by the VAQ and LOH filters, each of which remove 200 or more putative SNPs on average.
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50-100 mutatioins each per sample (Figure 6).
+The final filter, which removed putative SNPs with less than 10% coverage of the alternate allele, removed 0-1 putative SNPs on average, but did on occasion remove as many as 50.
+
+We next asked how many of the mutations removed by each filter are mutations present in the overlap between technical replicates, and how many are in just one sample?
+Put differently, what percentage of putative SNPs removed by a given filter are in the category more likley to be true positives (overlap), versus the category less likely to be true positives (only one replicate).
+To answer this, we graphed the percent of the overlap only (per sample) that was filtered by each of the six filters (Figure 7).
+We found that the three filters removeing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50-100 putative mutations on average.
+In contrast, the VAQ filter (secific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 7).
+Thus, our evidence suggests that the filters removeing overlap with dbSNP and putative mutaitons near other putative mutaions are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
 
 %% this is figures 6 and 7, which need to be added to the document and explained now...
+Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample.
+We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (figure 8), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (figure 9).
+We found opposite trends: the percentage filtered out by VAQ seems to decrease, and the percentage filtered out by LOH seems to increase, with increasing length of overlap.
+We conclude that the LOH and VAQ filters are not removing the same putative mutations.
 
@@
 All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Gliobastoma multiforme (GBM) data set \cite{TCGA-GBM}.
-We downloaded raw reads in .bam files from 55 patients using CGHub \cite{CGHub}.
-For each patient, data consisted of one set of reads taken from blood DNA, and two sets of reads taken from tumor DNA.
+We downloaded .bam files for 55 patients using CGHub \cite{CGHub}.
+For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate.
 In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library.
-On samples (WGS) did not include this step, and on (WGA) did).
 We developed a pipeline to align all reads to HG19.
-This pipeline used picard \cite{picard} to generate fastq files from bams aligned to HG18, BWA \cite{BWA} for alignment, SamTools \cite{SAMtools} to clean the data according to accepted standards \cite{seq-standards}, and GATK \cite{GATK} for initial SNP-calling and indel processing.
-These processes were strung together using custom python code.
-We used SomaticSniper \cite{SomaticSniper} to align each tumor sequence with it's corresponding blood sequence, then filtered the SomaticSniper data using 6 custom filters.
-These removed putative SNPs (i) with a SomaticScore less than 40, (ii) within 10 bp from a high-quality indel, (iii) within 10 bp of another putative SNP, (iv) with a variant allele quality less than 20, (v) with a dbSNP \cite{dbSNP} identifier and fewer than 10x coverage, and (vi) that were loss of heterozygosity.
-All custom code is available in a github repository, and is available upon request.
+This connected the pipeline using custom python code (available in a git repository upon request), to include the steps outlined in Figure 1, and given in more detail in Table 1.
 
-\subsection{Analysis}
+\subsection{Filtering and data analysis}
 
-We used custom python scripts, available in the above github repository, to perform simple calculations and data operations.
-We used R \cite{Rsoftware} to do statistics and generate figures, and this code is also stored in a github repository and available upon request.
+After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we filtered the putative SNPs according to the eight filters described avove using a combination of command line options and custom python code.
+The GATK quality score and SomaticScore filters were both done with flags when we called SomaticSniper, as shown above (-q 40 -Q 40), so that SomaticSniper returned no putative SNPs that did not pass these tests.
+SomaticSniper also identified, but did not exclude, LOH mutations; we wrote a custom script to remove them later.
+We coded custom python scripts for the other five filters, which are in a git repository and available upon request.
+We also wrote custom scripts to determine the overlap between samples, also available in the git repository upon request.
+
+We plotted all data and did all statistics with standard plotting and statictics functions in R \cite{Rsoftware}. This code is also available upon request.
 
@@
 %E) (optional) End with a paragraph that again summarizes your research and highlights the most important findings
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50% of putative SNPs in whole exome sequencing and about 30% in whole genome sequencing.
+We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in whole exome sequencing and about 30\% in whole genome sequencing.
 This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
-While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on teh putative mutaions in each technical replicate and the overlap between them, for all 55 samples.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutaions in each technical replicate and the overlap between them, for all 55 samples.
 We found that some filters remove principally those mutaions found in one sample or the other, while other filers remove primarily those in the overlap.
@@
 
-\section{Figures}
+\section{Tables and Figures}
+
+\begin{landscape}
+\begin{table}
+\begin{tabular}{ p{2.5cm} p{4cm} p{14cm} }
+	software & purpose & command\\
+	\hline
+	picard & regenerate fastq files from bamfile aligned to hg18 & java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1 \\
+	bwa & align fastq files to hg19 & bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai \\
+	bwa, samtools & convert aligned fastq files into new bamfile & bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam - \\
+	samtools & sort and index new bamfile & samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2 \\
+	samtools & remove duplicate reads from bamfiles & samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam \\
+	GATK & indel realignment & java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam \\
+	GATK & base recalibration & java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp \\
+	samtools & index recalibrated bamfile & samtools index \$tumorpfx.realn.recal.bam \\
+	SomaticSniper & call somatic mutations, generate VCF & bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf \\
+\end{tabular}
+\caption{\textbf{Data processing pipeline.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
+The rows are in order of use.}
+\end{table}
+\end{landscape}
 

### 2014-10-30 FIGURES/FIGURE_SCRIPTS/more_filter_figures.r
added the script for final figures

@@ -1,0 +1,57 @@
+# the purpose of this script is to make some figures...
+
+## 1. For LOH, VAQ, plot size of overlap (x-axis) against percent of overlap removed by relevant filter (y-axis)
+
+overlap_data = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
+filter_data = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+
+all_data <- merge( overlap_data, filter_data, by="SAMPLE" )
+
+# LOH (filter 7)
+LOH_vars = c( "OVERLAP", "FILT_7_COUNT" )
+LOH_data = all_data[ LOH_vars ]
+FILT_7_PERCENT_new = LOH_data$FILT_7_COUNT / LOH_data$OVERLAP
+LOH_data$FILT_7_PERCENT_new = FILT_7_PERCENT_new
+
+png( "/Users/dakota/Desktop/FIG_RD/LOH_all.png" )
+par(bty='n')
+plot( LOH_data$OVERLAP, LOH_data$FILT_7_PERCENT_new, xlab="Length of overlap", ylab="% of overlap filtered out by LOH", pch=20 )
+dev.off()
+
+png( "/Users/dakota/Desktop/FIG_RD/LOH_partial.png" )
+par(bty='n')
+plot( LOH_data$OVERLAP, LOH_data$FILT_7_PERCENT_new, xlim=c( 0, 500 ), xlab="Length of overlap", ylab="% of overlap filtered out by LOH", pch=20 )
+dev.off()
+
+# VAQ (filter 5)
+VAQ_vars = c( "OVERLAP", "FILT_5_PERCENT" )
+VAQ_data = all_data[ VAQ_vars ]
+
+png( "/Users/dakota/Desktop/FIG_RD/VAQ_all.png" )
+par(bty='n')
+plot( VAQ_data$OVERLAP, VAQ_data$FILT_5_PERCENT, xlab="Length of overlap", ylab="% of overlap filtered out by VAQ", pch=20 )
+dev.off()
+
+png( "/Users/dakota/Desktop/FIG_RD/VAQ_partial.png" )
+par(bty='n')
+plot( VAQ_data$OVERLAP, VAQ_data$FILT_5_PERCENT, xlim=c( 0, 1000 ), xlab="Length of overlap", ylab="% of overlap filtered out by VAQ", pch=20 )
[…]

### 2015-02-18 PAPER/paper_d3.tex
paper edits made

\documentclass[11pt]{article} % use larger type; default would be 10pt
\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% BEGIN Article customizations

%%% PAGE LAYOUT
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=1in} % for example, change the margins to 2 inches all round
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% FIGURES
\usepackage{graphicx} % support the \includegraphics command and options
\graphicspath{{../FIGURES/FIGURE_PDFS/}}

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{lscape}
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite} %**ask them about author order, I'm out of it...

\begin{document}
\maketitle

\section*{Abstract}

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability between sequencing runs of a single tumor, and to determine if the similarity between the two is increased or decreased by SNP-callers like SomaticSniper. To investigate this, we measured the extent of the overlap between two technical sequencing replicates from the same tumor (n=55), that is how many specific point mutations were found by both sequencing runs, using strictly computational means. We further examined whether the degree of overlap was higher or lower when using the SNP calling program SomaticSniper.  We found that about half of the putative mutations found in one sequencing run of a given sample are also found in the second, and that this percentage remains steady throughout orders of magnitude of variation in the total number of mutations called (50 to 7000). We further found that using a SNP caller removes the overlap completely, probably because the SNP caller weeds out such a high percentage of the total putative mutations (between 95\% and 100\%). We conclude that there is variation in the frequency of mutations in GBMs, and that some parts of the SNP caller function more accurately than others.

\section*{Introduction}

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \cite{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including 580 and counting samples alone of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \cite{GBM-stats}. In an effort to improve patient outcomes, TCGA and other groups \cite{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \cite{pathways}, discover different GBM subtypes \cite{subtypes}, and develop a variety of computational models to find GBM driver mutations \cite{drivers}.

% How the hell do I cite anything in this paragraph? There is NOTHING TO CITE. (Except that power point intro???)
Despite widespread use of TCGA data in these and other research efforts, and significant monetary investment in collecting the data, there has been very little investigation of the best way to sequence the 580+ paired tumor/normal samples in the TCGA GBM database. Most samples in the database include, at a minimum, whole exam sequencing of both the tumor and a matched blood sample at a coverage depth of between 8X and 80X, usually around 10--15X in the blood and 20--50X in the tumor. This is low, but similar to the 30X recommended for variant calling by the Broad Institute Introduction to NGS. However, we can find no research looking at the accuracy of variant calling at this or any sampling depth, for common or rare variants, in normal or cancerous tissue. Given that cancer is thought to be genetically heterogenous \cite{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \cite{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.

In this paper, we begin this inquiry with two simple questions: Given two technical sequencing replicates of a single tumor, how similar are they? Does SNP-calling software increase or decrease the degree similarity? We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol. For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \cite{SomaticSniper}. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. Still, the number of mutations in the WGS replicates only varied by orders of magnitude, from under 100 to almost 10,000. Contrary to expectations, the quality of results (measured by the percentage overlap between technical replicates) did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples. This may be in part due to know mutational hotspots in some tumors \cite{Karen}. Using SomaticSniper and some custom protocols to analyze the likelihood that individual variants were true positives (actual somatic mutations in the tumor, and not sequencing or analysis errors or various sorts) was unsuccessful, because almost no putative SNPs were identified as probably SNPs.

\section*{Results}

\subsection{Just how similar are two iterations of sequencing the same tumor?}

% Experiment: size of overlap (figures 3 and 4)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis. Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a sample that is sequenced, so that a partially penetrant mutation may appear fixed or not at all. In general, we would like to know how often these error occur. One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between the iterations. Theoretically, any mutation fixed throughout the tumor should appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors would not. (Incompletely penetrant mutations would also not be fixed in all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.

TCGA's GBM data set includes 2 technical replicates each of 55 tumors. In this case, the technical replicates are not identical; one includes an additional amplification of DNA before preparing a library; we refer to this replicate as the WGA replicate, and the other as the WGS replicate. Despite this difference, we can still assume that any mutation fixed throughout the tumor should appear in both replicates, and so those putative mutations appearing in both replicates are more likely to be real, fixed somatic variants than those found in only one replicate. We further hypothesized that the WGA samples should have a greater number of amplification errors, and thus there should be more putative SNPs per sample, than the corresponding WGS samples.

For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub, re-generated fastq files with picard, re-aligned the fastq files to hg19 with bwa, performed indel re-alignment and base recalibration with GATK, and finally called somatic mutations with SomaticSniper. We then compared the lists of putative SNPs generated by the WGS and WGA bamfiles for each patient, and calculated the length of each list and the length of the overlap, or intersection, between the two lists (Figure 2). We further calculated the percent of the WGA list that overlapped with the WGS list and the percent of the WGS list that overlapped with the WGA list. We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3). We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20--40\% in WGA replicates and mostly between 40--60\% in WGS replicates. As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).

% Experiment: number of mutations (figure 5)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 100 to over 10,000). Different cancers mutate at different rates: some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, while some tumors might have a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}. GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater. One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations. We found no correlation between the number of putative mutations and the percentage of the sample that was overlapping (Figure 5). That the percentage overlap does not vary with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.

\subsection{Does SNP-calling software increase or decrease the degree similarity between replicates?}

% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
The aim of sequencing tumors is generally to find somatic mutations, those mutations that arise in a tumor and are not present in the rest of the organism. To do this, researchers sequence a patient's tumor and blood and align the two to each other; Differences between the two sequences are putative somatic mutations. Because sequencing is not error free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed. If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and it's purity, this is often not possible. In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed.

Software platforms to distinguish somatic mutations from germ line mutations and sequencing error are plentiful, and each one typically employs multiple methods to identify true positive mutations. SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones. Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads. In this paper, we used eight different methods, or filters, to identify somatic mutations, summarized in Table 1. The first three are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds). The last five we coded ourselves in Python. Each of these five filters represents a quality of the data or putative mutation that is generally thought to indicate that it is probably a false positive.

Our initial question was, does filtering the data increase or decrease the percentage of the sample that is overlapping between the two technical replicates? Put another way, does filtering out putative somatic mutations with these features increase or decrease the proportion of true somatic mutations in the remaining dataset? We found that after removing putative mutations tagged by any of the eight filters, the number of putative mutations per replicate decreased from 50--10,000 to 0--14. The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, so the overall percentage also decreases. We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as the noise).

We next looked at the individual effects of each of the five filters that we coded, and one of the SomaticSniper filters (Variant Allele Quality, or VAQ). We first looked at the total number of putative SNPs removed by each filter (Figure 6). We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, each of which removed 200 or more putative SNPs on average. Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50--100 mutations each per sample (Figure 6). The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 0--1 putative SNPs on average, but did on occasion remove as many as 50.

We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample? Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category less likely to be true positives (only one replicate). To answer this, we graphed the percent of the overlap only (per sample) that was filtered by each of the six filters (Figure 7). We found that the three filters removing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50--100 putative mutations on average. In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 7). Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.

%% this is figures 8 and 9, which need to be added to the document and explained now...
Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample. We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure 8), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure 9). We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap. We concluded that the LOH and VAQ filters were not removing the same putative mutations.

\section*{Discussion}

GBM, primary brain cancer, is an evolutionary process by which mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole. Treatment for GBM is difficult and has poor outcomes \cite{GBM-treatments}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM. Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers. This data is enormously valuable to the research community, but the accuracy and reproducibility of these data are unknown, and knowing them could only increase the utility and value of the data. We make a first pass at evaluating the accuracy and repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.

We looked at the similarity between technical GBM genomic sequencing replicates for 55 samples in TCGA. We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates. The number of mutations present in both replicates was anywhere between 50 and 5,000 putative mutations. We found that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates. Further, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations. This suggests the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.

Filtering the raw computational data using both custom filters and those in the software SomaticSniper eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.  Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates. Combined, the two removed most or all of those mutations present in both replicates. Despite this, two still appeared to remove different sets of mutations. Of the remaining four, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate. Two of these three filters, those removing putative mutations within a 10 bp window of other putative mutations or within a 10 bp window of a putative indel, recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration. The third removed overlap with dbGaP.  Our analysis suggests that these three filters are useful, and do clean up the purely computational data in a meaningful way. By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates. Of these, one, VAQ is built into SomaticSniper. The second removes loss of heterozygosity (LOH) mutations. Multiple sources suggest that LOH mutations may be essential to cancer \cite{LOH}. This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.

Several factors limit the conclusions we may draw from this analysis. First, in this analysis we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP. This metric is potentially problematic for two reasons: (i) cancer is highly heterogeneous \cite{hetero}, and so a legitimate somatic SNP might show up in one replicate and not another; and (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates. Although repeatability does not and could not perfectly measure confidence, it should correlate. Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).  In this sample, we have shown that while some filters remove primarily putative SNPs in only one of two replicates (more likely to be false positives), two filters, VAQ and LOH, remove most of the overlap (more likely to be true positives). While our results suggest that VAQ and LOH filters are probably removing some true positives, that does not preclude them removing false positives. Therefore, if we remove these filters, we may retain more true positives, but we also retain more false positives.

Second, we looked at results from one SNP caller. There are many more equally widely used SNP callers available, which might do better or worse than the one we have chosen here. Additionally, we have shown that the VAQ filter does not avoid putative mutations in both replicates; however, we did not look at the SomaticScore, also a feature of SomaticSniper, or the GATK quality score.

We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in whole exome sequencing and about 30\% in whole genome sequencing. This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others. While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutations in each technical replicate and the overlap between them, for all 55 samples. We found that some filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap. We suggest using just the first set in the future, when mutation validation by PCR is not an option.

\section*{Methods}

\subsection{Data and back-end processing}

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \cite{TCGA-GBM}. We downloaded three bamfiles for each of 55 patients using CGHub \cite{CGHub}. For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library.

We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 2. The custom python code to connect the pipeline is available in a public 
git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}). our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard. Next, we used BWA to align the fastq files to hg19, and samtools to sort, index, clean, and de-duplicate the new bamfile. We used GATK to do indel realignment and base recalibration, according to the standards best practices for genomics data \cite{best-practices},. Finally, we predicted somatic variants with SomaticSniper \cite{SomaticSniper}, with the output given in VCF format.

\subsection{Filtering and data analysis}

After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates. We then filtered the lists of putative SNPs according to the eight filters described in Table 1, using a combination of command line options and custom python code (available in a public git repository). The eight filters were enacted as follows:

\begin{enumerate}
\item (GATK) The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item (SS) The SomaticScore is a similar metric calculated by SomaticSniper. As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item (VAQ) The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type. SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (LOH) It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale cancer genomics, because it to easily results from sequencing errors. We were able to exclude LOH variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (10bp-SNP) It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because this often indicates an error in reads or sequence alignment. We were able to exclude 10bp-SNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (10bp-INDEL) It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because this often indicates an error in reads or sequence alignment. We were able to exclude 10bp-INDEL variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (dbSNP) It is universal or near universal practice to exclude variants that overlap with dbSNP, because this often indicates an amplification error and not a true somatic variant. We were able to exclude dbSNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item ($<10\%$) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because this often indicates sequencing error. We were able to exclude $<10\%$ variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\end{enumerate}

We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering. These scripts are also available in the public git repository. We plotted all data and did all statistics with standard plotting and statistics functions in R \cite{Rsoftware}. This code is also available in the public git repository.

\section*{}

\bibliography{TCGA_bib}
\bibliographystyle{plain}

\section*{Tables and Figures}

%% Filters table
\begin{landscape}
\begin{table}
\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor bamfiles relative to blood bamfiles are real somatic variants, as opposed to sequencing error or other variants.}
\begin{tabular}{ p{2.25cm} p{4.25cm} p{14cm} }
	filter & software & purpose \\
	\hline
	GATK & GATK & Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration) \\
	SS & SomaticSniper & Removes putative SNPs with a SomaticScore less than 40 \\
	VAQ & SomaticSniper & Removes putative SNPs with SomaticSniper Varaint Allele Quality scores less than 20 \\
	LOH & SomaticSniper, python & Removes putative SNPs that are identified as loss of heterozygosity \\
	10bp-SNP & python & Removes putative SNPs located within a 10 bp window of any other putative SNP \\
	10bp-INDEL & python & Removes putative SNPs located within a 10 bp window of indels \\
	dbSNP & python & Removes putative SNPs that overlap with dbSNP coverage \\
	$<$10\% & python & Removes putative SNPs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10\% \\
\end{tabular}
\end{table}
\end{landscape}

%% Data processing commands table
\begin{landscape}
\begin{table}
\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it. The rows are in order of use.}
\begin{tabular}{ p{2.5cm} p{6cm} p{12cm} }
	software & purpose & command\\
	\hline
	picard & regenerate fastq files from bamfile aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
	bwa & align fastq files to hg19 & \texttt{bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai} \\
	bwa, samtools & convert aligned fastq files into new bamfile & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
	samtools & sort and index new bamfile & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
	samtools & remove duplicate reads from bamfiles & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
	GATK & indel realignment & \texttt{java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam} \\
	GATK & base recalibration & \texttt{java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp} \\
	samtools & index recalibrated bamfile & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
	SomaticSniper & call somatic mutations, generate VCF & \texttt{bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf} \\
\end{tabular}
\end{table}
\end{landscape}

%% Figure 1
\begin{figure}
\centerline{
\includegraphics[width=6in]{figure1.pdf} }
\caption{\textbf{Back-end processing pipeline.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18. For each bamfile, we used picard to regerate fastq files, bwa to realign the fastq files to hg19, GATK to recalibrate bases and indels, and SomaticSniper to call somatic mutations.}
\end{figure}

%% Figure 2
\begin{figure}
\centerline{
\includegraphics[width=6in]{figure2.pdf} }
\caption{\textbf{Definition of overlap between replicates.} For each sample, we generated a SomaticSniper VCF with putative somatic mutations for each of two replicates. We took the overlap between the two lists for the most likely true positives.}
\end{figure}

%% Figure 3
\begin{figure}
\centerline{
\includegraphics[width=5in]{C282_v_C484.pdf} }
\caption{\textbf{Number of putative SNPs in WGS v. WGA, as called by SomaticSniper.} Each point is a patient. The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.}
\end{figure}

%% Figure 4
\begin{figure}
\centerline{
\includegraphics[width=5in]{unfiltered_overlap_WGS_WGA_together_densities.pdf} }
\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient. The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter. }
\end{figure}

%% Figure 5
\begin{figure}
\centerline{
\includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample.}
\end{figure}

%% Figure 6
\begin{figure}
\centerline{
\includegraphics[width=6in]{boxplot_number_filtered.pdf} }
\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x-axis$ gives the name of those filters that removed putative SNPs, and the $y-axis$ gives, on a log scale, the number of mutations removed by a given filter in a given sample. Filters removing LOH and low VAQ putative SNPs removed around 100--200 putative SNPs on average. Three other filters, removing overlap with dbSNP and putative SNPs close to other putative mutations, removed on average about 50 mutations per sample.}
\end{figure}

%%Figure 7
\begin{figure}
% 379 samples, 311 patients
\centerline{
\includegraphics[width=6in]{boxplot_percent_overlap_filtered.pdf} }
\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x-axis$ describes the filters, and the $y-axis$ gives the percentage of the WGS/WGA overlap removed by the filter, per sample. Filters removing LOH and putative SNPs with low VAQ removed a significant percentage of the overlap; nothing else did.}
\end{figure}

%%Figure 8
\begin{figure}
\centerline{
\includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the number of mutations filtered out by LOH increases. This is the opposite of the pattern observed in VAQ mutations (Figure 6).}
\end{figure}

%%Figure 9
\begin{figure}
\centerline{
\includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} }
\caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by VAQ decreases. This is the opposite of the pattern observed in LOH mutations (Figure 5).}
\end{figure}

\end{document}

### 2015-03-20 MAKE_VCF/IN_SEQUENCE/SNPs_GATK.sh
worked on GATK processing scripts

@@ -1,100 +1,96 @@
 #!/bin/bash
 
 # load modules (gatk current default is 2.7.2)
-module load gatk/2.7.2
+module load gatk
 module load samtools
 module load picard 
 
 ## this script does the GATK processing and variant calibration on all of the sorted .bam and .bam.bai files
 
-cd /scratch/00001/mattcowp/dakota
+dataDir="/scratch/00001/mattcowp/dakota"
+cd $dataDir
 
 ## get sample names and file names, and make sure the two files have the same sample name
-tumor_pfx=$1
-blood_pfx=$2
-tumor_sample=${tumor_pfx:0:17}
-blood_sample=${blood_pfx:0:17}
-tumor_dir="$tumor_pfx.WORK"
+tumorPfx=$1
+bloodPfx=$2
+tumorSamp=${tumorPfx:0:17}
+bloodSamp=${bloodPfx:0:17}
+tumorDir="$tumorPfx.WORK"
 # echo $tumor_dir
-echo $tumor_pfx $tumor_sample
-echo $blood_pfx $blood_sample
-if [ $tumor_sample != $blood_sample ]; then
+echo $tumorPfx $tumorSamp
+echo $bloodPfx $bloodSamp
+if [ $tumorSamp != $bloodSamp ]; then
 	echo "Non-matching sample names:"
-	echo "	$tumor_sample"
-	echo "	$blood_sample"
+	echo "	$tumorSamp"
+	echo "	$bloodSamp"
 	exit 1;
 fi
 
 ## locations of relevant and needed files
-gatk_jar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
-ref_dir="/work/00001/mattcowp/Hs_reference_datasets"
+gatkJar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
 hgReference="$ref_dir/Homo_sapiens.GRCh37.72.dna.fa"
-dbSNP="$ref_dir/dbsnp_137.b37.vcf"
-hapMap="$ref_dir/hapmap_3.3.b37.vcf"
-G1000_HiConf_SNPs="$ref_dir/1000G_phase1.snps.high_confidence.b37.vcf"
-G1000_Phase1_Indels="$ref_dir/1000G_phase1.indels.b37.vcf"
-G1000_Omni_SNPs="$ref_dir/1000G_omni2.5.b37.vcf"
[…]

### 2015-03-20 PAPER/paper_d3.tex
not sure if there is anything

@@
 \pagestyle{fancy} % options: empty , plain , fancy
-\renewcommand{\headrulewidth}{0pt} % customise the layout...
+\renewcommand{\headrulewidth}{0pt} % customize the layout...
 \lhead{}\chead{}\rhead{}
@@
 
+%%%Bibliography
+\usepackage{natbib}
+
 %%% END Article customizations
@@
 \title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
-\author{Dakota Z. Derryberry, Claus O. Wilke, Matthew C. Cowperthwaite} %**ask them about author order, I'm out of it...
+\author{Dakota Z. Derryberry, Matthew C. Cowperthwaite, Claus O. Wilke} %**ask them about author order, I'm out of it...
 
@@
 
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \cite{TCGA-GBM, TCGA-GBM-13}.
+The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
 TCGA's database includes hundreds of samples of each of several tumor types, including 580 and counting samples alone of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
-GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \cite{GBM-stats}.
-In an effort to improve patient outcomes, TCGA and other groups \cite{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \cite{pathways}, discover different GBM subtypes \cite{subtypes}, and develop a variety of computational models to find GBM driver mutations \cite{drivers}.
+GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \citep{GBM-stats}.
+In an effort to improve patient outcomes, TCGA and other groups \citep{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
 
-% How the hell do I cite anything in this paragraph? There is NOTHING TO CITE. (Except that power point intro???)
 Despite widespread use of TCGA data in these and other research efforts, and significant monetary investment in collecting the data, there has been very little investigation of the best way to sequence the 580+ paired tumor/normal samples in the TCGA GBM database.
@@
 However, we can find no research looking at the accuracy of variant calling at this or any sampling depth, for common or rare variants, in normal or cancerous tissue.
-Given that cancer is thought to be genetically heterogenous \cite{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \cite{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.
+Given that cancer is thought to be genetically heterogenous \citep{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \citep{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.
 
@@
 We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol.
-For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \cite{SomaticSniper}.
+For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}.
 We found significant overlap (around 50\%) between technical replicates, but also significant differences.
@@
 Contrary to expectations, the quality of results (measured by the percentage overlap between technical replicates) did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples.
-This may be in part due to know mutational hotspots in some tumors \cite{Karen}.
+This may be in part due to know mutational hotspots in some tumors \citep{Karen}.
 Using SomaticSniper and some custom protocols to analyze the likelihood that individual variants were true positives (actual somatic mutations in the tumor, and not sequencing or analysis errors or various sorts) was unsuccessful, because almost no putative SNPs were identified as probably SNPs.
@@
 
-\subsection{Just how similar are two iterations of sequencing the same tumor?}
+\subsection*{Just how similar are two iterations of sequencing the same tumor?}
 
@@
 DNA sequencing is not error-free.
-Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis.
+Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}.
 Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified.
@@
 
-For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub, re-generated fastq files with picard, re-aligned the fastq files to hg19 with bwa, performed indel re-alignment and base recalibration with GATK, and finally called somatic mutations with SomaticSniper.
+For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
 We then compared the lists of putative SNPs generated by the WGS and WGA bamfiles for each patient, and calculated the length of each list and the length of the overlap, or intersection, between the two lists (Figure 2).
@@
 Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 100 to over 10,000).
-Different cancers mutate at different rates: some pediatric cancers that arise from single and double hit mutations \cite{pediatric}, while some tumors might have a mutator phenotype, which usually results from errors in DNA repair pathways \cite{mutator}.
-GBM specifically is thought to have a relatively low mutation rate \cite{GBM_mut_rate}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater.
+Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}; other tumors show a mutator phenotype, leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
+GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater.
 One possible explanation is a degraded DNA sample, or bad data.
@@
 
-\subsection{Does SNP-calling software increase or decrease the degree similarity between replicates?}
+\subsection*{Does SNP-calling software increase or decrease the degree similarity between replicates?}
 
@@
 Because sequencing is not error free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed.
-If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and it's purity, this is often not possible.
-In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed.
+If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and its purity, this is often not possible.
+In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed \citep{SomaticSniper, mut_calling}.
 
 Software platforms to distinguish somatic mutations from germ line mutations and sequencing error are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones.
-Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads.
+SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones \citep{SomaticSniper}.
+Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads \cite{GATK}.
 In this paper, we used eight different methods, or filters, to identify somatic mutations, summarized in Table 1.
@@
 GBM, primary brain cancer, is an evolutionary process by which mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole.
-Treatment for GBM is difficult and has poor outcomes \cite{GBM-treatments}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM.
+Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM.
 Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
@@
 The second removes loss of heterozygosity (LOH) mutations.
-Multiple sources suggest that LOH mutations may be essential to cancer \cite{LOH}.
+Multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
 This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
@@
 First, in this analysis we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
-This metric is potentially problematic for two reasons: (i) cancer is highly heterogeneous \cite{hetero}, and so a legitimate somatic SNP might show up in one replicate and not another; and (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates.
-Although repeatability does not and could not perfectly measure confidence, it should correlate.
+This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
 Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).
@@
 
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in whole exome sequencing and about 30\% in whole genome sequencing.
+We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in WGS samples and about 30\% in WGA samples.
 This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
@@
 
-\subsection{Data and back-end processing}
+\subsection*{Data and back-end processing}
 
-All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \cite{TCGA-GBM}.
-We downloaded three bamfiles for each of 55 patients using CGHub \cite{CGHub}.
+All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
+We downloaded three bamfiles for each of 55 patients using CGHub \citep{CGHub}.
 For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate.
-In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library.
+In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
 
@@
 git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard.
-Next, we used BWA to align the fastq files to hg19, and samtools to sort, index, clean, and de-duplicate the new bamfile.
-We used GATK to do indel realignment and base recalibration, according to the standards best practices for genomics data \cite{best-practices},.
-Finally, we predicted somatic variants with SomaticSniper \cite{SomaticSniper}, with the output given in VCF format.
+Our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard \citep{picard}.
+Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools}to sort, index, clean, and de-duplicate the new bamfile.
+We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standards best practices for genomics data \citep{best-practices}.
+Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
 
-\subsection{Filtering and data analysis}
+\subsection*{Filtering and data analysis}
 
@@
 These scripts are also available in the public git repository.
-We plotted all data and did all statistics with standard plotting and statistics functions in R \cite{Rsoftware}.
+We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
 This code is also available in the public git repository.
 
-\section*{}
+%\section*{}
 
 \bibliography{TCGA_bib}
-\bibliographystyle{plain}
+\bibliographystyle{peerj}
 

### 2015-03-20 PAPER/paper_d1.tex

@@
 We used technical sequencing replicates from 55 GBM tumors to look at the reproducbility of the raw data using a standard computational pipeline.
-Assuming that those mutations that appear in each of the technical replicates are true positives, and all others are false positives (which in fact may or may not be true), we want to see what proportion of mutations recovered are ture positives, and whether or not the proportion is increased by SNP-calling methods designed to remove false positives.
+Assuming that those mutations that appear in each of the technical replicates are true positives, and all others are false positives (which in fact may or may not be true), we want to see what proportion of mutations recovered are true positives, and whether or not the proportion is increased by SNP-calling methods designed to remove false positives.
 

### 2015-04-17 PAPER/paper_d4.tex
another draft
editing draft 4

--- 2015-03-20 PAPER/paper_d3.tex
@@
 \usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
+\usepackage[T1]{fontenc}
+\usepackage{authblk}
 
@@
 \usepackage{natbib}
+\usepackage{url}
 
@@
 
-%%% The "real" document content comes below...
-
+%%% Title and author are here
 \title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
-\author{Dakota Z. Derryberry, Matthew C. Cowperthwaite, Claus O. Wilke} %**ask them about author order, I'm out of it...
+\author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
+\author[3]{Matthew C. Cowperthwaite}
+\author[1,2]{Claus O. Wilke} 
+\affil[1]{The University of Texas at Austin, Cell \& Molecular Biology}
+\affil[2]{The University of Texas at Austin, Integrative Biology}
+\affil[3]{St. David's NeuroTexas Institute Research Foundation}
 
@@
 
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability between sequencing runs of a single tumor, and to determine if the similarity between the two is increased or decreased by SNP-callers like SomaticSniper.
-To investigate this, we measured the extent of the overlap between two technical sequencing replicates from the same tumor (n=55), that is how many specific point mutations were found by both sequencing runs, using strictly computational means.
-We further examined whether the degree of overlap was higher or lower when using the SNP calling program SomaticSniper.
-We found that about half of the putative mutations found in one sequencing run of a given sample are also found in the second, and that this percentage remains steady throughout orders of magnitude of variation in the total number of mutations called (50 to 7000).
-We further found that using a SNP caller removes the overlap completely, probably because the SNP caller weeds out such a high percentage of the total putative mutations (between 95\% and 100\%).
-We conclude that there is variation in the frequency of mutations in GBMs, and that some parts of the SNP caller function more accurately than others.
+We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling when using the same data analysis pipeline on two instances of sequencing the same tumor, and to determine what additional filtering might increase or decrease the total similarity.
+We analyzed 55 pairs of technical replicates using the same computational pipeline and measured the extent of the overlap between two replicates, that is how many specific point mutations were found in both replicates.
+We further tested whether additional filtering increased or decreased the size of the overlap.
+We found that about half of the putative mutations found in one sequencing run of a given sample are also found in the second, and that this percentage remains steady throughout orders of magnitude of variation in the total number of mutations called (23 to 10,966).
+We further found that using a SNP and subsequent filtering removes the overlap completely.
+We concluded that there is variation in the frequency of mutations in GBMs, and that while some parts of the SNP-caller preferentially removed putative mutations found in only one replicate, others removed primarily putative mutations found in both.
 
@@
 The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
-TCGA's database includes hundreds of samples of each of several tumor types, including 580 and counting samples alone of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
-GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotheraputics \citep{GBM-stats}.
-In an effort to improve patient outcomes, TCGA and other groups \citep{Parsons} have used large scale genomic data to discover which genes and pathways are mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
+TCGA's database includes hundreds of samples of each of several tumor types, including 528 and counting samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
+GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
+To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
 
-Despite widespread use of TCGA data in these and other research efforts, and significant monetary investment in collecting the data, there has been very little investigation of the best way to sequence the 580+ paired tumor/normal samples in the TCGA GBM database.
-Most samples in the database include, at a minimum, whole exam sequencing of both the tumor and a matched blood sample at a coverage depth of between 8X and 80X, usually around 10--15X in the blood and 20--50X in the tumor.
-This is low, but similar to the 30X recommended for variant calling by the Broad Institute Introduction to NGS.
-However, we can find no research looking at the accuracy of variant calling at this or any sampling depth, for common or rare variants, in normal or cancerous tissue.
-Given that cancer is thought to be genetically heterogenous \citep{heterogenous1, heterogenous2}, and that rare variants in GBM may be functionally important \citep{rare}, we think that some inquiry into the accuracy and precision of standard cancer sequencing methods is warranted.
+But these and other efforts have produced very few results for patients.
+It may be that we simply need more new ways to analyze the data.
+Another factor, however, may be that the data are noisy.
+It is well known that sequencing and SNP-calling pipelines are not error-free.
+Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal to noise ratio in this data set may be even worse.
+Despite widespread use of the TCGA dataset, and significant monetary investment in collecting and analyzing the data, we found very little research concerning the best way to maximize the quality of the sequence data and SNP-calls.
+Because different SNP-calling pipelines have been known to return different results on the same data \citep{SNPcall}, we turned this question around and asked, will the same SNP-calling pipeline return the same results on two sequencing runs from the same tissue?
+Does SNP-calling software and filtering increase or decrease the degree similarity?
 
-In this paper, we begin this inquiry with two simple questions: Given two technical sequencing replicates of a single tumor, how similar are they?
-Does SNP-calling software increase or decrease the degree similarity?
 We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol.
-For each of these 55 samples, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}.
+For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}.
 We found significant overlap (around 50\%) between technical replicates, but also significant differences.
 As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-Still, the number of mutations in the WGS replicates only varied by orders of magnitude, from under 100 to almost 10,000.
-Contrary to expectations, the quality of results (measured by the percentage overlap between technical replicates) did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples.
+Still, the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples.
 This may be in part due to know mutational hotspots in some tumors \citep{Karen}.
-Using SomaticSniper and some custom protocols to analyze the likelihood that individual variants were true positives (actual somatic mutations in the tumor, and not sequencing or analysis errors or various sorts) was unsuccessful, because almost no putative SNPs were identified as probably SNPs.
+Our attempt to use SomaticSniper and some custom filters to increase the similarity between technical replicates was unsuccessful: filtering removed almost the entirety of the overlap between the replicates.
 
@@
 
-\subsection*{Just how similar are two iterations of sequencing the same tumor?}
+\subsection*{How similar are two iterations of sequencing the same tumor?}
 
@@
 Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a sample that is sequenced, so that a partially penetrant mutation may appear fixed or not at all.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or not at all.
 In general, we would like to know how often these error occur.
 One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming.
-A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between the iterations.
-Theoretically, any mutation fixed throughout the tumor should appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors would not.
-(Incompletely penetrant mutations would also not be fixed in all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
+A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between replicates.
+Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
+(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
 Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.
@@
 TCGA's GBM data set includes 2 technical replicates each of 55 tumors.
-In this case, the technical replicates are not identical; one includes an additional amplification of DNA before preparing a library; we refer to this replicate as the WGA replicate, and the other as the WGS replicate.
-Despite this difference, we can still assume that any mutation fixed throughout the tumor should appear in both replicates, and so those putative mutations appearing in both replicates are more likely to be real, fixed somatic variants than those found in only one replicate.
-We further hypothesized that the WGA samples should have a greater number of amplification errors, and thus there should be more putative SNPs per sample, than the corresponding WGS samples.
+In this case, the technical replicates are not identical.
+One protocol included additional amplification step \citep{TCGA}, and we refer to this replicate as the WGA replicate, and the other as the WGS replicate.
+Despite this difference, we still assumed that any fully penetrant mutation would appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one.
+Thus overall, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
+We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.
 
 For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the lists of putative SNPs generated by the WGS and WGA bamfiles for each patient, and calculated the length of each list and the length of the overlap, or intersection, between the two lists (Figure 2).
-We further calculated the percent of the WGA list that overlapped with the WGS list and the percent of the WGS list that overlapped with the WGA list.
-We found that the number of mutations found in one replicate correlated strongly with the number of mutations found in the other, though as expected the WGA replicate, with the additional amplification step, had slightly more mutations (Figure 3).
-We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, mostly between 20--40\% in WGA replicates and mostly between 40--60\% in WGS replicates.
+We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called for each replicate and the number of individual somatic mutations called by both replicates (hereafter, the length of the overlap, see Figure 2).
+We further calculated the percentage of of mutations in each replicate that occurred in the overlap between the two.
+
+There were between 110 and 8,192 mutations (average=844, median=328, standard deviation=1306) in the WGS replicates and between 23 and 10,966 mutations in the WGA replicates (average=1531, median=694, standard deviation=2230).
+Across all samples, the number of mutations in the a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 3).
+As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures 3 and 4).
+We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates (median=31\%, min=1\%, max=74\%, stdev=20\%) and mostly 44\% in WGS replicates (median=45\%, min=3\%, max=71\%, stdev=13\%).
 As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).
@@
 %(b) a description of what you found carrying out this experiment
-Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 100 to over 10,000).
-Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}; other tumors show a mutator phenotype, leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
-GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while the majority of our samples had low mutation frequencies in line with this theory, several samples also had mutation frequencies an order of magnitude greater.
+Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 23 to over 10,966).
+Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
+GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
 One possible explanation is a degraded DNA sample, or bad data.
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-We found no correlation between the number of putative mutations and the percentage of the sample that was overlapping (Figure 5).
-That the percentage overlap does not vary with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation rate than others, or indeed than is generally supposed in GBM.
+We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure 5), though it is possible that with more more samples this could become a weak correlation.
+Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
 
@@
 
-% Experiment: which filters do what (figures 4 and 5) % Experiment: those two filters are different (figures 6 and 7)
+% Experiment: which filters do what (figures 6 and 7) 
+% Experiment: those two filters are different (figures 8 and 9)
 %(a) a brief (one sentence to one paragraph) summary of what you did and why
@@
 The aim of sequencing tumors is generally to find somatic mutations, those mutations that arise in a tumor and are not present in the rest of the organism.
-To do this, researchers sequence a patient's tumor and blood and align the two to each other; Differences between the two sequences are putative somatic mutations.
-Because sequencing is not error free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed.
+To do this, researchers sequence a patient's tumor and blood and align the two to each other; differences between the two sequences are putative somatic mutations.
+Because sequencing is not error-free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed.
 If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and its purity, this is often not possible.
-In this case, computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing error may be employed \citep{SomaticSniper, mut_calling}.
+In this case, we may use computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
 
-Software platforms to distinguish somatic mutations from germ line mutations and sequencing error are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-SomaticSniper, the platform used in this research, uses several probabilistic calculations based on features of the dataset to sort out true putative mutations from false ones \citep{SomaticSniper}.
-Another example, the GATK QualityScore, calculates the probability that a mutation is a true mutations based on features of the reads \cite{GATK}.
-In this paper, we used eight different methods, or filters, to identify somatic mutations, summarized in Table 1.
-The first three are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds).
-The last five we coded ourselves in Python.
+Software platforms to distinguish somatic mutations from germ line mutations and sequencing errors are plentiful, and each one typically employs multiple methods to identify true positive mutations.
+The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \cite{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and mutations with higher quality scores are considered more likely than those with low quality scores to be true somatic mutations.
+Additionally, after removing reads with low quality scores, there are additional filtering steps used to distinguish somatic mutations from errors of all sorts.
+The entries in Table 1 describe each of the ways we filtered the raw SNPs called for each sample.
+The first three entries are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds).
+The last five are additional filters that we coded ourselves in Python.
 Each of these five filters represents a quality of the data or putative mutation that is generally thought to indicate that it is probably a false positive.
@@
 Put another way, does filtering out putative somatic mutations with these features increase or decrease the proportion of true somatic mutations in the remaining dataset?
-We found that after removing putative mutations tagged by any of the eight filters, the number of putative mutations per replicate decreased from 50--10,000 to 0--14.
+We found that after removing putative mutations tagged by any of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14.
 The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, so the overall percentage also decreases.
@@
 
-We next looked at the individual effects of each of the five filters that we coded, and one of the SomaticSniper filters (Variant Allele Quality, or VAQ).
+We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ).
 We first looked at the total number of putative SNPs removed by each filter (Figure 6).
@@
 \begin{table}
-\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor bamfiles relative to blood bamfiles are real somatic variants, as opposed to sequencing error or other variants.}
+\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
 \begin{tabular}{ p{2.25cm} p{4.25cm} p{14cm} }
@@
 \caption{\textbf{Back-end processing pipeline.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18.
-For each bamfile, we used picard to regerate fastq files, bwa to realign the fastq files to hg19, GATK to recalibrate bases and indels, and SomaticSniper to call somatic mutations.}
+For each bamfile, we used picard to regerate fastq files, bwa to realign the fastq files to hg19, GATK to recalibrate bases and indels.
+We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.}
 \end{figure}
@@
 \caption{\textbf{Definition of overlap between replicates.} For each sample, we generated a SomaticSniper VCF with putative somatic mutations for each of two replicates.
-We took the overlap between the two lists for the most likely true positives.}
+We calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
 \end{figure}
@@
 \caption{\textbf{Number of putative SNPs in WGS v.
-WGA, as called by SomaticSniper.} Each point is a patient.
-The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.}
+WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
+The number of mutations found in one replicate correlates with the number of mutations found in the other replicate, cor=0.51, t = 4.3, df = 53, p-value = 8.35e-05.}
 \end{figure}
@@
 \includegraphics[width=5in]{unfiltered_overlap_WGS_WGA_together_densities.pdf} }
-\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient.
+\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
@@
 \includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample.}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample.
+(cor=0.25, p=.06, not significant)}
 \end{figure}

### 2015-04-17 PAPER/paper_d4.tex
more draft4 edits

@@
 We first looked at the total number of putative SNPs removed by each filter (Figure 6).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, each of which removed 200 or more putative SNPs on average.
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of other mutations, also removed around 50--100 mutations each per sample (Figure 6).
-The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 0--1 putative SNPs on average, but did on occasion remove as many as 50.
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 (median=164, min=17, max=11,439, stdev=1,372) and 539 (median=169, min=3, max=8,538, stdev=1,051) putative SNPs, respectively.
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16 (median=8, min=0, max=348, stdev=33), 28 (median=16, min=0, max=535, stdev=45), and 46 (median=19, min=0, max=1,515, stdev=115) putative mutations per sample (Figure 6).
+The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average (median=1, min=1, max=36, sd=3).
 

### 2015-04-17 PAPER/paper_d4.tex
finished d4 edits

@@
 %%% Title and author are here
-\title{Reproducibility of SNP calling in multiple sequencing runs from single tumors}
+\title{Reproducibility of SNP-calling in multiple sequencing runs from single tumors}
 \author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
@@
 In this case, the technical replicates are not identical.
-One protocol included additional amplification step \citep{TCGA}, and we refer to this replicate as the WGA replicate, and the other as the WGS replicate.
+One protocol included additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA replicate, and the other as the WGS replicate.
 Despite this difference, we still assumed that any fully penetrant mutation would appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one.
@@
 We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category less likely to be true positives (only one replicate).
-To answer this, we graphed the percent of the overlap only (per sample) that was filtered by each of the six filters (Figure 7).
-We found that the three filters removing overlap with dbSNP and other putative mutations removed almost none or none of the overlap, despite removing 50--100 putative mutations on average.
-In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed about half of the overlap between replicates (Figure 7).
+Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category less likely to be false positives (only one replicate)?
+To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 7).
+We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\% (median=2\%, min=0\%, max=47\%, stdev=7\%), 4\% (median=4\%, min=0\%, max=22\%, stdev=4\%), or 3\% (median=2\%, min=0\%, max=20\%, stdev=3\%) of the overlap, or almost none, respectively.
+In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed on average 53\% (median=57\%, min=0\%, max=100\%, stdev=28\%) and 51\% (median=52\%, min=0\%, max=99\%, stdev=29\%), or about half each, of the overlap respectively (Figure 7).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
@@
 We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap.
+The two graphs are almost, but not quite, perfect inverses.
+But given that (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time, this is expected.
 We concluded that the LOH and VAQ filters were not removing the same putative mutations.
@@
 
-GBM, primary brain cancer, is an evolutionary process by which mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole.
-Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but could hopefully be improved by a more complete understanding of the somatic mutations present in GBM.
+GBM is an evolutionary disease that develops when mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
+Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM.
 Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-This data is enormously valuable to the research community, but the accuracy and reproducibility of these data are unknown, and knowing them could only increase the utility and value of the data.
-We make a first pass at evaluating the accuracy and repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.
+These data are enormously valuable to the research community, but their accuracy and reproducibility are unknown, and the knowing could only increase the utility of the data.
+We make a first pass at evaluating the repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.
 
-We looked at the similarity between technical GBM genomic sequencing replicates for 55 samples in TCGA.
+We looked at the similarity between technical sequencing replicates for 55 GBM samples in TCGA.
 We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
-The number of mutations present in both replicates was anywhere between 50 and 5,000 putative mutations.
+The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.
 We found that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
@@
 
-Filtering the raw computational data using both custom filters and those in the software SomaticSniper eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates.
-Combined, the two removed most or all of those mutations present in both replicates.
-Despite this, two still appeared to remove different sets of mutations.
-Of the remaining four, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative mutations within a 10 bp window of other putative mutations or within a 10 bp window of a putative indel, recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration.
-The third removed overlap with dbGaP.
-Our analysis suggests that these three filters are useful, and do clean up the purely computational data in a meaningful way.
-By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates.
-Of these, one, VAQ is built into SomaticSniper.
-The second removes loss of heterozygosity (LOH) mutations.
-Multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
+Filtering the raw computational data using both quality scores from GATK and SomaticSniper as well as five additional custom filters eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
+On some level, this is unsurprising: these filters were designed to be used on wild type genomes, where it is generally assumed that more differences are due to error than to real changes (e.g.
+LOH in this case is more likely to be an error than a change).
+They are designed to reduce or eliminate differences between the reference and the sample.
+When we do cancer genomics, our goal is the opposite: to highlight changes.
+Therefor, it is possible that we need entirely different filtering protocols.
+There are also more theoretical reasons to consider altering the filtering protocols for cancer genomes.
+For example, multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
 This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
 
+Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates, and combined, they removed most or all of those mutations present in both replicates; the two filters removed almost disjoint sets of putative mutations.
+Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
+Two of these three filters, those removing putative mutations within a 10 bp window of putative indels or other putative SNPs, recognized a feature (clustered mutations) that suggests a local problem with the reads or alignment, and removed this data from consideration.
+The third removed overlap with dbGaP.
+Our analysis suggests that these three filters do clean up the data in a meaningful way.
+By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates.
+
 Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
+First, in this analysis we used repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
 This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
 Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
-Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).
-In this sample, we have shown that while some filters remove primarily putative SNPs in only one of two replicates (more likely to be false positives), two filters, VAQ and LOH, remove most of the overlap (more likely to be true positives).
-While our results suggest that VAQ and LOH filters are probably removing some true positives, that does not preclude them removing false positives.
-Therefore, if we remove these filters, we may retain more true positives, but we also retain more false positives.
+Having an independent measure of confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).
 
@@
 There are many more equally widely used SNP callers available, which might do better or worse than the one we have chosen here.
-Additionally, we have shown that the VAQ filter does not avoid putative mutations in both replicates; however, we did not look at the SomaticScore, also a feature of SomaticSniper, or the GATK quality score.
+Also, we did not even look at every quality metric available in the SNP-caller we did choose (we did not look at the effects individually of the SomaticScore or the GATK quality score).
+In the future, we would like to evaluate other SNP-callers and other quality metrics.
 
@@
 This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
-While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six strictly computational filters on the putative mutations in each technical replicate and the overlap between them, for all 55 samples.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters for the same purpose.
 We found that some filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest using just the first set in the future, when mutation validation by PCR is not an option.
+We suggest using only the first set in the future, when mutation validation by PCR is not an option.
 
@@
 Our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools}to sort, index, clean, and de-duplicate the new bamfile.
-We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standards best practices for genomics data \citep{best-practices}.
+Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new bamfile.
+We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
 Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
@@
 This recommendation is discussed in the Results and Discussion.
-\item (LOH) It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale cancer genomics, because it to easily results from sequencing errors.
-We were able to exclude LOH variants using a custom python script available in the public git repository.
+\item (LOH) It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
+We excluded LOH variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (10bp-SNP) It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because this often indicates an error in reads or sequence alignment.
-We were able to exclude 10bp-SNP variants using a custom python script available in the public git repository.
+\item (10bp-SNP) It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+We excluded 10bp-SNP variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (10bp-INDEL) It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because this often indicates an error in reads or sequence alignment.
-We were able to exclude 10bp-INDEL variants using a custom python script available in the public git repository.
+\item (10bp-INDEL) It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (dbSNP) It is universal or near universal practice to exclude variants that overlap with dbSNP, because this often indicates an amplification error and not a true somatic variant.
-We were able to exclude dbSNP variants using a custom python script available in the public git repository.
+\item (dbSNP) It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
+We excluded dbSNP variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item ($<10\%$) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because this often indicates sequencing error.
-We were able to exclude $<10\%$ variants using a custom python script available in the public git repository.
+\item ($<10\%$) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
+We excluded $<10\%$ variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
@@
 \includegraphics[width=6in]{boxplot_number_filtered.pdf} }
-\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x-axis$ gives the name of those filters that removed putative SNPs, and the $y-axis$ gives, on a log scale, the number of mutations removed by a given filter in a given sample.
-Filters removing LOH and low VAQ putative SNPs removed around 100--200 putative SNPs on average.
-Three other filters, removing overlap with dbSNP and putative SNPs close to other putative mutations, removed on average about 50 mutations per sample.}
+\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x-axis$ gives the name of each filter (detail on Table 1), and the $y-axis$ gives, on a log scale, the number of mutations removed by a given filter in a given sample.
+In all cases, the LOH and VAQ filters removed the majority of the putative SNPs for a given sample.}
 \end{figure}
@@
 \includegraphics[width=6in]{boxplot_percent_overlap_filtered.pdf} }
-\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x-axis$ describes the filters, and the $y-axis$ gives the percentage of the WGS/WGA overlap removed by the filter, per sample.
+\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x-axis$ gives the names of the filters, and the $y-axis$ gives the percentage of the WGS/WGA overlap removed by the filter, per sample.
 Filters removing LOH and putative SNPs with low VAQ removed a significant percentage of the overlap; nothing else did.}
@@
 \includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
-\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the number of mutations filtered out by LOH increases.
-This is the opposite of the pattern observed in VAQ mutations (Figure 6).}
+\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by LOH increases.
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure 6).}
 \end{figure}
@@
 \caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by VAQ decreases.
-This is the opposite of the pattern observed in LOH mutations (Figure 5).}
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure 5).}
 \end{figure}

### 2015-04-29 PAPER/paper_d5.tex
another draft...

--- 2015-04-17 PAPER/paper_d4.tex
@@
 Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}.
-Loss of heterozygosity also occurs in sequencing, if only one allele of a polymorphic site is amplified.
+Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified.
 Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or not at all.
@@
 For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called for each replicate and the number of individual somatic mutations called by both replicates (hereafter, the length of the overlap, see Figure 2).
+We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the length of the overlap, see Figure 2).
 We further calculated the percentage of of mutations in each replicate that occurred in the overlap between the two.
 
-There were between 110 and 8,192 mutations (average=844, median=328, standard deviation=1306) in the WGS replicates and between 23 and 10,966 mutations in the WGA replicates (average=1531, median=694, standard deviation=2230).
-Across all samples, the number of mutations in the a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 3).
+There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 3).
 As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures 3 and 4).
-We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates (median=31\%, min=1\%, max=74\%, stdev=20\%) and mostly 44\% in WGS replicates (median=45\%, min=3\%, max=71\%, stdev=13\%).
+We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table 1).
 As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).
@@
 Software platforms to distinguish somatic mutations from germ line mutations and sequencing errors are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \cite{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and mutations with higher quality scores are considered more likely than those with low quality scores to be true somatic mutations.
+The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and mutations with higher quality scores are considered more likely than those with low quality scores to be true somatic mutations.
 Additionally, after removing reads with low quality scores, there are additional filtering steps used to distinguish somatic mutations from errors of all sorts.
-The entries in Table 1 describe each of the ways we filtered the raw SNPs called for each sample.
-The first three entries are the quality scores generated by the GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds).
+The entries in Table 2 describe each of the ways we filtered the raw SNPs called in each sample.
+The first three entries are the quality scores generated by GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds).
 The last five are additional filters that we coded ourselves in Python.
@@
 We first looked at the total number of putative SNPs removed by each filter (Figure 6).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 (median=164, min=17, max=11,439, stdev=1,372) and 539 (median=169, min=3, max=8,538, stdev=1,051) putative SNPs, respectively.
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16 (median=8, min=0, max=348, stdev=33), 28 (median=16, min=0, max=535, stdev=45), and 46 (median=19, min=0, max=1,515, stdev=115) putative mutations per sample (Figure 6).
-The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average (median=1, min=1, max=36, sd=3).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (see Table 1 for summary statistics).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure 6, summary statistics in Table 1).
+The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.
 
 We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category less likely to be false positives (only one replicate)?
+Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)?
 To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 7).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\% (median=2\%, min=0\%, max=47\%, stdev=7\%), 4\% (median=4\%, min=0\%, max=22\%, stdev=4\%), or 3\% (median=2\%, min=0\%, max=20\%, stdev=3\%) of the overlap, or almost none, respectively.
-In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed on average 53\% (median=57\%, min=0\%, max=100\%, stdev=28\%) and 51\% (median=52\%, min=0\%, max=99\%, stdev=29\%), or about half each, of the overlap respectively (Figure 7).
+We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (summary statistics in Table 1).
+In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure 7, summary statistics in Table 1).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
@@
 
-We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 2. The custom python code to connect the pipeline is available in a public 
+We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 3. The custom python code to connect the pipeline is available in a public 
 git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
@@
 \section*{Tables and Figures}
+
+%% Summary Statistics Table
+\begin{landscape}
+\begin{table}
+\caption{\textbf{TABLE CAPTION} Table explanation.}
+\begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
+	Quantity & average & median & min & max & stdev \\
+	\hline
+	No. mutations, WGS & 844 & 328 & 110 & 8,192 & 1,306 \\
+	No. mutations, WGA & 1,531 & 694 & 23 & 10,966 & 2,230 \\
+	\% mutations in overlap, WGS & 31\% & 31\% & 1\% & 74\% & 20\% \\
+	\% mutations in overlap, WGA & 44\% & 45\% & 3\% & 71\% & 13\% \\
+	No. putative mutations removed, VAQ & 309 & 164 & 17 & 11,439 & 1,372 \\
+	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
+	No. putative mutations removed, 10bp-SNP & 46 & 19 & 0 & 1,515 & 115 \\
+	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
+	No. putative mutations removed, dbSNP & 16 & 8 & 0 & 348 & 33 \\
+	No. putative mutations removed, $<$10\% & 1 & 1 & 1 & 36 & 3 \\
+	\% overlap removed, VAQ & 53\% & 57\% & 0\% & 100\% & 28\% \\
+	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
+	\% overlap removed, 10bp-SNP & 3\% & 2\% & 0\% & 20\% & 3\% \\
+	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
+	\% overlap removed, dbSNP & 3\% & 2\% & 0\% & 47\% & 7\% \\
+\end{tabular}
+\end{table}
+\end{landscape}
 

### 2015-04-29 MAKE_VCF/IN_SEQUENCE/SNPs_GATK.sh
changes to SNPs_GATK to makr duplicates instead of remove duplicates

@@ -6,10 +6,13 @@ module load gatk
 module load samtools/0.1.19 # used older version of samtools only on the 190 run
 module load picard 
 
-## this script does the GATK processing and variant calibration on all of the sorted .bam and .bam.bai files
+## this script does the GATK processing and variant calibration on all of the out.sorted.matefixed.bam (and corresponding .bam.bai) files
 
 dataDir="/scratch/00001/mattcowp/dakota"
 cd $dataDir
+
+echo "Begin $0: `date`">>$tumorPfx.gatk.log
+echo "-----------------------------------------------------------------------------">>$tumorPfx.gatk.log
 
 ## get sample names and file names, and make sure the two files have the same sample name
 tumorPfx=$1
@@ -18,8 +21,8 @@ tumorSamp=${tumorPfx:0:17}
 bloodSamp=${bloodPfx:0:17}
 tumorDir="$tumorPfx.WORK"
 # echo $tumor_dir
-echo $tumorPfx $tumorSamp &>>../$tumorPfx.gatk.log
-echo $bloodPfx $bloodSamp &>>../$tumorPfx.gatk.log
+echo $tumorPfx $tumorSamp &>>$tumorPfx.gatk.log
+echo $bloodPfx $bloodSamp &>>$tumorPfx.gatk.log
 if [ $tumorSamp != $bloodSamp ]; then
 	echo "Non-matching sample names:"
 	echo "	$tumorSamp"
@@ -38,68 +41,124 @@ G1000Phase1Indels="$refDir/1000G_phase1.indels.b37.vcf"
 G1000OmniSNPs="$refDir/1000G_omni2.5.b37.vcf"
 G1000Mills="$refDir/Mills_and_1000G_gold_standard.indels.b37.vcf"
 
-echo "Generating somatic calls for $tumorPfx" >> $tumorPfx.gatk.log
+echo "Generating somatic calls for $tumorPfx" &>>$tumorPfx.gatk.log
 
 # make a temporary directory in which this will actually run
 mkdir $tumorDir
-echo "Moving to working directory $tumorDir" >> $tumorPfx.gatk.log
+echo "Moving to working directory $tumorDir" &>>$tumorPfx.gatk.log
 cd $tumorDir
 
[…]

### 2015-05-14 PAPER/paper_d5.tex
poster text added
combined figures 1 and 2 into new figure 1

@@
 There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 3).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 2).
 As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures 3 and 4).
 We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table 1).
-As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 4).
+As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 3).
 
-% Experiment: number of mutations (figure 5)
+% Experiment: number of mutations (figure 4)
 %(a) a brief (one sentence to one paragraph) summary of what you did and why
@@
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure 5), though it is possible that with more more samples this could become a weak correlation.
+We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure 4), though it is possible that with more more samples this could become a weak correlation.
 Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
@@
 We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ).
-We first looked at the total number of putative SNPs removed by each filter (Figure 6).
+We first looked at the total number of putative SNPs removed by each filter (Figure 5).
 We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (see Table 1 for summary statistics).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure 6, summary statistics in Table 1).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure 5, summary statistics in Table 1).
 The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.
@@
 Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)?
-To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 7).
+To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 6).
 We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (summary statistics in Table 1).
-In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure 7, summary statistics in Table 1).
+In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure 6, summary statistics in Table 1).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
@@
 Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample.
-We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure 8), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure 9).
+We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure 7), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure 8).
 We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap.
@@
 These data are enormously valuable to the research community, but their accuracy and reproducibility are unknown, and the knowing could only increase the utility of the data.
-We make a first pass at evaluating the repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.
+We made a first pass at evaluating the repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.
 
@@
 \centerline{
-\includegraphics[width=6in]{figure1.pdf} }
-\caption{\textbf{Back-end processing pipeline.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18.
-For each bamfile, we used picard to regerate fastq files, bwa to realign the fastq files to hg19, GATK to recalibrate bases and indels.
-We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.}
+\includegraphics[width=5in]{figs1and2combo.pdf} }
+\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18.
+For each bamfile, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
+We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
+When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
 \end{figure}
@@
 %% Figure 2
-\begin{figure}
-\centerline{
-\includegraphics[width=6in]{figure2.pdf} }
-\caption{\textbf{Definition of overlap between replicates.} For each sample, we generated a SomaticSniper VCF with putative somatic mutations for each of two replicates.
-We calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
-\end{figure}
-
-%% Figure 3
 \begin{figure}
@@
 
-%% Figure 4
+%% Figure 3
 \begin{figure}
@@
 
-%% Figure 5
+%% Figure 4
 \begin{figure}
@@
 
-%% Figure 6
+%% Figure 5
 \begin{figure}
@@
 
-%%Figure 7
+%%Figure 6
 \begin{figure}
@@
 
-%%Figure 8
+%%Figure 7
 \begin{figure}
@@
 \caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by LOH increases.
-This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure 6).}
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure 5).}
 \end{figure}
 
-%%Figure 9
+%%Figure 8
 \begin{figure}
@@
 \caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by VAQ decreases.
-This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure 5).}
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure 4).}
 \end{figure}

### 2015-05-18 PAPER/paper.tex
Intro and first part of Results

--- 2015-05-14 PAPER/paper_d5.tex
@@
 \usepackage{lscape}
+\usepackage{amsmath}
 % These packages are all incorporated in the memoir class to one degree or another...
@@
 
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling when using the same data analysis pipeline on two instances of sequencing the same tumor, and to determine what additional filtering might increase or decrease the total similarity.
-We analyzed 55 pairs of technical replicates using the same computational pipeline and measured the extent of the overlap between two replicates, that is how many specific point mutations were found in both replicates.
+We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling.
+We used the same computational pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
 We further tested whether additional filtering increased or decreased the size of the overlap.
-We found that about half of the putative mutations found in one sequencing run of a given sample are also found in the second, and that this percentage remains steady throughout orders of magnitude of variation in the total number of mutations called (23 to 10,966).
-We further found that using a SNP and subsequent filtering removes the overlap completely.
+We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966).
+We further found that using filtering after SNP-calling removed the overlap completely.
 We concluded that there is variation in the frequency of mutations in GBMs, and that while some parts of the SNP-caller preferentially removed putative mutations found in only one replicate, others removed primarily putative mutations found in both.
@@
 
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
-TCGA's database includes hundreds of samples of each of several tumor types, including 528 and counting samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
-GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
+The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
+TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
+GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%.
+Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
 To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
 
-But these and other efforts have produced very few results for patients.
-It may be that we simply need more new ways to analyze the data.
-Another factor, however, may be that the data are noisy.
+However, these and other efforts have produced very few results for patients in terms of improved treatments.
+It may be that we simply haven't yet analyzed the data in the most informative manner.
+Another factor may be that the data are noisy.
 It is well known that sequencing and SNP-calling pipelines are not error-free.
-Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal to noise ratio in this data set may be even worse.
-Despite widespread use of the TCGA dataset, and significant monetary investment in collecting and analyzing the data, we found very little research concerning the best way to maximize the quality of the sequence data and SNP-calls.
-Because different SNP-calling pipelines have been known to return different results on the same data \citep{SNPcall}, we turned this question around and asked, will the same SNP-calling pipeline return the same results on two sequencing runs from the same tissue?
-Does SNP-calling software and filtering increase or decrease the degree similarity?
+For example, different SNP-calling pipelines can return different results on the same data \citep{SNPcall}.
+Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.
+Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls.
+Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return the comparable results on two sequencing runs from the same tissue.
+And further, we asked whether added filtering after SNP calling increases or decreases the degree of similarity.
 
-We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol.
+We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
 For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}.
@@
 As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-Still, the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples.
-This may be in part due to know mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SomaticSniper and some custom filters to increase the similarity between technical replicates was unsuccessful: filtering removed almost the entirety of the overlap between the replicates.
+We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency between samples.
+This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
+Our attempt to use SomaticSniper filters, as well as some custom filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
 
@@
 
-\subsection*{How similar are two iterations of sequencing the same tumor?}
+\subsection*{How similar are two replicate sequencing runs of the same tumor?}
 
@@
 Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or not at all.
-In general, we would like to know how often these error occur.
-One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming.
-A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between replicates.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or completely absent.
+In general, we would like to know how often these errors occur.
+One way of investigating this question would be to use PCR to verify each individual mutation.
+However, this method is expensive and time consuming.
+A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates.
 Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
-Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.
+(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiment, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
+Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
 
@@
 In this case, the technical replicates are not identical.
-One protocol included additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA replicate, and the other as the WGS replicate.
-Despite this difference, we still assumed that any fully penetrant mutation would appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one.
-Thus overall, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
+One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
+Despite the difference in sequencing protocol, we would expect any fully penetrant mutation to appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one.
+Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
 We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.
 
-For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the length of the overlap, see Figure 2).
-We further calculated the percentage of of mutations in each replicate that occurred in the overlap between the two.
+For each patient (n=55), we called mutations in both technical replicates and in the patient's blood sample, using the same computational pipeline (see Figure~\ref{fig:pipeline} and Methods): We downloaded TCGA BAM files with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
+We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the overlap, see Figure~\ref{fig:pipeline}).
+We further calculated the percentage of mutations in each replicate that occurred in the overlap between the two.
 
+\textbf{copy edit from here}
 There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 2).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures 3 and 4).
-We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table 1).
-As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 3).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure~\ref{fig:unfiltered_overlap}).
+As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:unfiltered_overlap} and~\ref{fig:unfiltered_total_muts}).
+We further found that the percent overlap between the two samples, calculated by $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table~\ref{tab:summary_stats}).
+As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
 
@@
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure 4), though it is possible that with more more samples this could become a weak correlation.
+We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}), though it is possible that with more more samples this could become a weak correlation.
 Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
@@
 We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ).
-We first looked at the total number of putative SNPs removed by each filter (Figure 5).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (see Table 1 for summary statistics).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure 5, summary statistics in Table 1).
+We first looked at the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
 The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.
@@
 Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)?
-To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 6).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (summary statistics in Table 1).
-In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure 6, summary statistics in Table 1).
+To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
+We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (Table~\ref{tab:summary_stats}).
+In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
@@
 Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample.
-We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure 7), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure 8).
+We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure~\ref{fig:LOH_all}), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure~\ref{fig:VAQ_all}).
 We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap.
@@
 All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
-We downloaded three bamfiles for each of 55 patients using CGHub \citep{CGHub}.
-For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate.
+We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
+For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
 In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
 
-We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 3. The custom python code to connect the pipeline is available in a public 
+We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table~\ref{tab:commands}.
+The custom python code to connect the pipeline is available in a public
 git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-Our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new bamfile.
+Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
+Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
 We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
@@
 After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNPs according to the eight filters described in Table 1, using a combination of command line options and custom python code (available in a public git repository).
+We then filtered the lists of putative SNPs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
 The eight filters were enacted as follows:
 
-\begin{enumerate}
-\item (GATK) The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
+\begin{itemize}
+\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
 We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item (SS) The SomaticScore is a similar metric calculated by SomaticSniper.
+\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
 As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
 We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item (VAQ) The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
+\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
 SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (LOH) It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
+\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
 We excluded LOH variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (10bp-SNP) It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+\item[10bp-SNP:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
 We excluded 10bp-SNP variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (10bp-INDEL) It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
 We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item (dbSNP) It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
+\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
 We excluded dbSNP variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\item ($<10\%$) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
+\item[$<10\%$:]) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
 We excluded $<10\%$ variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
-\end{enumerate}
+\end{itemize}
 
@@
 \caption{\textbf{TABLE CAPTION} Table explanation.}
+\label{tab:summary_stats}
 \begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
@@
 \begin{table}
+\label{tab:filters}
 \caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
@@
 \begin{table}
+\label{tab:commands}
 \caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
@@
 	\hline
-	picard & regenerate fastq files from bamfile aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
+	picard & regenerate fastq files from BAM file aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
 	bwa & align fastq files to hg19 & \texttt{bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai} \\
-	bwa, samtools & convert aligned fastq files into new bamfile & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
-	samtools & sort and index new bamfile & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
-	samtools & remove duplicate reads from bamfiles & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
+	bwa, samtools & convert aligned fastq files into new BAM file & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
+	samtools & sort and index new BAM file & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
+	samtools & remove duplicate reads from BAM files & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
 	GATK & indel realignment & \texttt{java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam} \\
 	GATK & base recalibration & \texttt{java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp} \\
-	samtools & index recalibrated bamfile & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
+	samtools & index recalibrated BAM file & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
 	SomaticSniper & call somatic mutations, generate VCF & \texttt{bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf} \\
@@
 \includegraphics[width=5in]{figs1and2combo.pdf} }
-\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor bamfile (WGS), a C282 tumor bamfile (WGA), and a normal bamfile, all aligned to hg18.
-For each bamfile, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
+\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
+For each BAM file, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
 We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
 When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
+\label{fig:pipeline}
 \end{figure}
@@
 The number of mutations found in one replicate correlates with the number of mutations found in the other replicate, cor=0.51, t = 4.3, df = 53, p-value = 8.35e-05.}
+\label{fig:C282_v_C484}
 \end{figure}
@@
 }
+\label{fig:unfiltered_overlap}
 \end{figure}
@@
 (cor=0.25, p=.06, not significant)}
+\label{fig:unfiltered_total_muts}
 \end{figure}
@@
 In all cases, the LOH and VAQ filters removed the majority of the putative SNPs for a given sample.}
+\label{fig:boxplot_number_filtered}
 \end{figure}
@@
 Filters removing LOH and putative SNPs with low VAQ removed a significant percentage of the overlap; nothing else did.}
+\label{fig:boxplot_percent_overlap_filtered}
 \end{figure}
@@
 \caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by LOH increases.
-This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure 5).}
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}).}
+\label{fig:LOH_all}
 \end{figure}
@@
 \caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by VAQ decreases.
-This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure 4).}
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}).}
+\label{fig:VAQ_all}
 \end{figure}

### 2015-05-21 PAPER/paper.tex
revised manuscript.

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{color}
 % These packages are all incorporated in the memoir class to one degree or another...
@@
 
-\textbf{copy edit from here}
-There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure~\ref{fig:unfiltered_overlap}).
+There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Pearson $r=0.51$, $df=53$, $P=8.35\times10^{-5}$, Figure~\ref{fig:unfiltered_overlap}).
 As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:unfiltered_overlap} and~\ref{fig:unfiltered_total_muts}).
-We further found that the percent overlap between the two samples, calculated by $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table~\ref{tab:summary_stats}).
+We further found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
 As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
@@
 %(b) a description of what you found carrying out this experiment
-Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 23 to over 10,966).
-Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
+Although the percent overlap in WGS and WGA samples was fairly constant across samples, and the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations varied by orders of magnitude (Table~\ref{tab:summary_stats}).
+It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
 GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample, or bad data.
+One possible explanation is a degraded DNA sample or otherwise bad data.
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}), though it is possible that with more more samples this could become a weak correlation.
-Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
+However, we found no significant correlation (Pearson $r=0.25$, {\color{red}$df=XXX$}, $P=0.06$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
+While the correlation might turn significant with a larger data set, the observed correlation coefficient was also low, corresponding to only 6.25\% of the variation in the data explained by the number of putative mutations.
+Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
 
-\subsection*{Does SNP-calling software increase or decrease the degree similarity between replicates?}
+\subsection*{Does more sophisticated SNP filtering software increase or decrease the degree of similarity between replicates?}
 
@@
 %(b) a description of what you found carrying out this experiment
-The aim of sequencing tumors is generally to find somatic mutations, those mutations that arise in a tumor and are not present in the rest of the organism.
-To do this, researchers sequence a patient's tumor and blood and align the two to each other; differences between the two sequences are putative somatic mutations.
-Because sequencing is not error-free, and because a patient's blood tissue might also have somatic mutations different from the tumor, not all of the putative somatic mutations are mutations, and some method of distinguishing true somatic mutations from sequencing and other errors is needed.
-If possible, this is done by individually validating each putative mutation with PCR; however, due to restraints on time and money, as well as the amount of tumor tissue available and its purity, this is often not possible.
-In this case, we may use computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
+As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
+Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations.
+The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores.
+In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts.
+Table~\ref{tab:filters} lists eight distinct SNP filters we evaluated.
+The first three are based on the quality scores generated by GATK and SomaticSniper.
+We simply remove from the dataset anything that fails to meet conventional quality-score thresholds.
+The other five are additional filters that we developed for this project.
+Each of these five filters represents an aspect of the data or putative mutation that is generally thought to indicate that a given SNP call is a false positive.
 
-Software platforms to distinguish somatic mutations from germ line mutations and sequencing errors are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and mutations with higher quality scores are considered more likely than those with low quality scores to be true somatic mutations.
-Additionally, after removing reads with low quality scores, there are additional filtering steps used to distinguish somatic mutations from errors of all sorts.
-The entries in Table 2 describe each of the ways we filtered the raw SNPs called in each sample.
-The first three entries are the quality scores generated by GATK and SomaticSniper (we simply remove from the dataset anything that fails to meet these thresholds).
-The last five are additional filters that we coded ourselves in Python.
-Each of these five filters represents a quality of the data or putative mutation that is generally thought to indicate that it is probably a false positive.
+We first asked whether filtering the data increases or decreases the percentage of the sample that is overlapping between the two technical replicates.
+We expect that this analysis informs us about whether filtering out putative somatic mutations with these features affects the proportion of true somatic mutations in the remaining dataset.
+We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14.
+The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, i.e., the overall overlap percentage also decreased.
+We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as all the noise).
 
-Our initial question was, does filtering the data increase or decrease the percentage of the sample that is overlapping between the two technical replicates?
-Put another way, does filtering out putative somatic mutations with these features increase or decrease the proportion of true somatic mutations in the remaining dataset?
-We found that after removing putative mutations tagged by any of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14.
-The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, so the overall percentage also decreases.
-We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as the noise).
-
-We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ).
-We first looked at the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
+We next looked at the individual effects of six of the eight filters.
+({{\color{red}Is this parenthetical remark correct:} We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
 The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.
 
-We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)?
-To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (Table~\ref{tab:summary_stats}).
-In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
-Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.
+We next asked how many of the SNPs removed by each filter were SNPs present in the overlap between technical replicates, and how many were in just one sample?
+Put differently, what percentage of putative SNPs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
+To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
+We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
+By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
+Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
 
 %% this is figures 8 and 9, which need to be added to the document and explained now...
-Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample.
-We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure~\ref{fig:LOH_all}), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure~\ref{fig:VAQ_all}).
-We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap.
-The two graphs are almost, but not quite, perfect inverses.
-But given that (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time, this is expected.
-We concluded that the LOH and VAQ filters were not removing the same putative mutations.
+Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
+We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ {\color{red}(please double-check the value 100)} there was no strong trend.
+Either filter was removing between 0 and 100\% of the overlap for some samples.
+For samples with more overlap, however, LOH was the primary filter removing overlap.
+
+We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH_all}) and the VAQ graph (Figure~\ref{fig:VAQ_all}) looked like exact inverses of each other.
+Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
+In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.
 
@@
 
-GBM is an evolutionary disease that develops when mutations arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
+GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
 Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM.
@@
 These data are enormously valuable to the research community, but their accuracy and reproducibility are unknown, and the knowing could only increase the utility of the data.
-We made a first pass at evaluating the repeatability of the data by comparing 55 technical replicates in the TCGA GBM dataset.
+Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
 
-We looked at the similarity between technical sequencing replicates for 55 GBM samples in TCGA.
 We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
@@
 
-Filtering the raw computational data using both quality scores from GATK and SomaticSniper as well as five additional custom filters eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-On some level, this is unsurprising: these filters were designed to be used on wild type genomes, where it is generally assumed that more differences are due to error than to real changes (e.g.
-LOH in this case is more likely to be an error than a change).
-They are designed to reduce or eliminate differences between the reference and the sample.
-When we do cancer genomics, our goal is the opposite: to highlight changes.
-Therefor, it is possible that we need entirely different filtering protocols.
+Filtering the raw computational data using both quality scores from GATK and SomaticSniper, as well as five additional custom filters, eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
+To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNPs.
+For example, LOH in this case is more likely to be an error than an actual mutation.
+When we do cancer genomics, however, our goal is the opposite, to highlight changes.
+Therefore, it is possible that we need entirely different filtering protocols.
 There are also more theoretical reasons to consider altering the filtering protocols for cancer genomes.
 For example, multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
-This, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
+This hypothesis, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
 
-Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that found in both technical replicates, and combined, they removed most or all of those mutations present in both replicates; the two filters removed almost disjoint sets of putative mutations.
+Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that we found in both technical replicates.
+In combination, they removed most or all of those mutations present in both replicates, since the two filters consistently removed nearly completely disjoint sets of putative mutations.
 Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative mutations within a 10 bp window of putative indels or other putative SNPs, recognized a feature (clustered mutations) that suggests a local problem with the reads or alignment, and removed this data from consideration.
-The third removed overlap with dbGaP.
+Two of these three filters, those removing putative SNPs within a 10 bp window of putative indels or other putative SNPs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
+The third removes overlap with dbGaP.
 Our analysis suggests that these three filters do clean up the data in a meaningful way.
-By contrast, it may be more useful to discard the two filters that removed principally data from the overlap of the two replicates.
+By contrast, it may be more useful not to apply the two filters that removed principally data from the overlap of the two replicates.
 
@@
 First, in this analysis we used repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
-This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
 Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
-Having an independent measure of confidence in SNP calls (repeatability), even an imperfect one, can help us gauge the accuracy of other measures (filtering).
+Having an independent measure of confidence in SNP calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
 
-Second, we looked at results from one SNP caller.
-There are many more equally widely used SNP callers available, which might do better or worse than the one we have chosen here.
-Also, we did not even look at every quality metric available in the SNP-caller we did choose (we did not look at the effects individually of the SomaticScore or the GATK quality score).
-In the future, we would like to evaluate other SNP-callers and other quality metrics.
+Second, we studied only one particular SNP caller, SomaticSniper \citep{SomaticSniper}.
+There are other, equally widely used SNP callers available {\color{red}please cite some}, which might do better or worse than the one we have chosen here.
+Also, we did not even consider every possible quality metric available in the SNP-caller we did choose.
+For example, we did not look at the effects of the SomaticScore or the GATK quality score individually.
+In the future, it would be worthwhile to evaluate other SNP-callers and other quality metrics.
 
 We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in WGS samples and about 30\% in WGA samples.
-This effect remains even at a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
-While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters for the same purpose.
-We found that some filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest using only the first set in the future, when mutation validation by PCR is not an option.
+The overlap exists even for samples with a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters that are commonly applied to validate SNPs.
+We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
+We suggest that when mutation validation by PCR is not an option, only the filters that removed little overlap between samples should be used for computational SNP validation.
 
@@
 This recommendation is discussed in the Results and Discussion.
-\item[$<10\%$:]) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
+\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
 We excluded $<10\%$ variants using a custom python script available in the public git repository.
@@
 \begin{table}
-\caption{\textbf{TABLE CAPTION} Table explanation.}
+\caption{\color{red}\textbf{TABLE CAPTION} Table explanation.}
 \label{tab:summary_stats}
@@
 \begin{table}
+\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
 \label{tab:filters}
-\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
 \begin{tabular}{ p{2.25cm} p{4.25cm} p{14cm} }
@@
 \begin{table}
-\label{tab:commands}
 \caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
 The rows are in order of use.}
+\label{tab:commands}
 \begin{tabular}{ p{2.5cm} p{6cm} p{12cm} }
@@
 We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
+When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.
+{\color{red}Comment on the figure: The two colored boxes called ``SomaticSniper VCF'' need labels ``WGS'' and ``WGA'' next to them.}}
 \label{fig:pipeline}
@@
 \includegraphics[width=5in]{C282_v_C484.pdf} }
-\caption{\textbf{Number of putative SNPs in WGS v.
-WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+\caption{\textbf{Number of putative SNPs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate, cor=0.51, t = 4.3, df = 53, p-value = 8.35e-05.}
+The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Pearson $r=0.51$, $df = 53$, $P = 8.35\times 10^{-5}$).}
 \label{fig:C282_v_C484}
@@
 \includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample.
-(cor=0.25, p=.06, not significant)}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample (Pearson $r=0.25$, {\color{red}$df=XXX$}, $P=0.06$).}
 \label{fig:unfiltered_total_muts}
@@
 \includegraphics[width=6in]{boxplot_number_filtered.pdf} }
-\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x-axis$ gives the name of each filter (detail on Table 1), and the $y-axis$ gives, on a log scale, the number of mutations removed by a given filter in a given sample.
+\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
 In all cases, the LOH and VAQ filters removed the majority of the putative SNPs for a given sample.}
@@
 \includegraphics[width=6in]{boxplot_percent_overlap_filtered.pdf} }
-\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x-axis$ gives the names of the filters, and the $y-axis$ gives the percentage of the WGS/WGA overlap removed by the filter, per sample.
+\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
 Filters removing LOH and putative SNPs with low VAQ removed a significant percentage of the overlap; nothing else did.}
@@
 \includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
-\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by LOH increases.
+\caption{\textbf{{\color{red}Figure needs a better title.
+The current one is unclear.} Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
 This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}).}
@@
 \includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} }
-\caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x-axis$), the percentage of the overlap filtered out by VAQ decreases.
+\caption{ \textbf{{\color{red}Figure needs a better title.
+The current one is unclear.} Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
 This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}).}

### 2015-05-30 PAPER/paper.tex
more figure editing...

@@
 \centerline{
-\includegraphics[width=5in]{figs1and2combo.pdf} }
+\includegraphics[width=5in]{pipeline.pdf} }
 \caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
@@
 We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.
-{\color{red}Comment on the figure: The two colored boxes called ``SomaticSniper VCF'' need labels ``WGS'' and ``WGA'' next to them.}}
+When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
 \label{fig:pipeline}
@@
 \includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample (Pearson $r=0.25$, {\color{red}$df=XXX$}, $P=0.06$).}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
 \label{fig:unfiltered_total_muts}
@@
 \includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
-\caption{\textbf{{\color{red}Figure needs a better title.
-The current one is unclear.} Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
-This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}).}
+\caption{\textbf{Percentage of replicate SNPs filtered out by LOH, as a function of the total number of overlapping SNPs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}), indicating that these two filters remove almost disjoint sets of putative SNPs.}
 \label{fig:LOH_all}
@@
 \includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} }
-\caption{ \textbf{{\color{red}Figure needs a better title.
-The current one is unclear.} Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
-This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}).}
+\caption{ \textbf{Percentage of replicate SNPs filtered out by VAQ, as a function of the total number of overlapping SNPs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}), indicating that these two filters remove almost disjoint sets of putative SNPs.}
 \label{fig:VAQ_all}

### 2015-05-30 PAPER/paper.tex
final paper edits

@@
 Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls.
-Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return the comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNP calling increases or decreases the degree of similarity.
+Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return comparable results on two sequencing runs from the same tissue.
+And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity.
 
@@
 Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiment, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
+(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
 Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
@@
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Pearson $r=0.25$, {\color{red}$df=XXX$}, $P=0.06$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
-While the correlation might turn significant with a larger data set, the observed correlation coefficient was also low, corresponding to only 6.25\% of the variation in the data explained by the number of putative mutations.
+However, we found no significant correlation (Pearson $r=-0.04$, $df=53$, $P=0.76$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
 Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
@@
 We next looked at the individual effects of six of the eight filters.
-({{\color{red}Is this parenthetical remark correct:} We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
+(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
 We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}).
@@
 Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ {\color{red}(please double-check the value 100)} there was no strong trend.
+We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
 Either filter was removing between 0 and 100\% of the overlap for some samples.
@@
 Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we used repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
+First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
 This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
@@
 Second, we studied only one particular SNP caller, SomaticSniper \citep{SomaticSniper}.
-There are other, equally widely used SNP callers available {\color{red}please cite some}, which might do better or worse than the one we have chosen here.
+There are other, equally widely used SNP callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
 Also, we did not even consider every possible quality metric available in the SNP-caller we did choose.
@@
 
-After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
+After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
 We then filtered the lists of putative SNPs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
@@
 \begin{table}
-\caption{\color{red}\textbf{TABLE CAPTION} Table explanation.}
+\caption{\textbf{Summary Statistics.} This table describes the results of our data processing pipeline across samples and pairs of replicates, in terms of the number of putative mutations found, length of the overlap between replicates, number of mutations removed (per sample) by each filter, and percent of the overlap removed (per pair of replicates) by each filter.}
 \label{tab:summary_stats}

### 2015-06-01 PAPER/paper.tex
finishing touches

@@
 We further found that using filtering after SNP-calling removed the overlap completely.
-We concluded that there is variation in the frequency of mutations in GBMs, and that while some parts of the SNP-caller preferentially removed putative mutations found in only one replicate, others removed primarily putative mutations found in both.
+We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
 
@@
 Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity.
+And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity among replicates.
 
 We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
-For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}.
+For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNP-filtering approaches.
 We found significant overlap (around 50\%) between technical replicates, but also significant differences.
@@
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SomaticSniper filters, as well as some custom filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
+Our attempt to use SNP filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
 
@@
 DNA sequencing is not error-free.
-Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}.
+Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}.
 Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified.
@@
 There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Pearson $r=0.51$, $df=53$, $P=8.35\times10^{-5}$, Figure~\ref{fig:unfiltered_overlap}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:unfiltered_overlap} and~\ref{fig:unfiltered_total_muts}).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Pearson $r=0.51$, $df=53$, $P=8.35\times10^{-5}$, Figure~\ref{fig:C282_v_C484}).
+As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_overlap}).
 We further found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
@@
 We expect that this analysis informs us about whether filtering out putative somatic mutations with these features affects the proportion of true somatic mutations in the remaining dataset.
-We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14.
+We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14 (Table~\ref{tab:summary_stats}).
 The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, i.e., the overall overlap percentage also decreased.
@@
 Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-These data are enormously valuable to the research community, but their accuracy and reproducibility are unknown, and the knowing could only increase the utility of the data.
+These data are enormous potential value to the research community, but their accuracy and reproducibility are unknown.
 Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
@@
 The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.
-We found that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
-Further, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations.
-This suggests the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
+We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
+Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations.
+These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
 
@@
 
+\cleardoublepage
 \section*{Tables and Figures}
@@
 \includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of a given WGS samples that is in the overlap between replicates against the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
 \label{fig:unfiltered_total_muts}

### 2015-06-02 PAPER/paper.tex
Matt's edits are included in the source file.
I think the paper is MUCH improved over the previous version.  Well done!

My main comment is that I would not call somatic mutations as SNPs

@@
 We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling.
-We used the same computational pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
+We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
 We further tested whether additional filtering increased or decreased the size of the overlap.
@@
 Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
+To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
 
+% Matt: I would use SNV, rather than SNP, since SNP is more commonly used to refer to germline variants.  Or just use "somatic mutation calling pipeline"
 However, these and other efforts have produced very few results for patients in terms of improved treatments.
@@
 %(b) a description of what you found carrying out this experiment
+%
+% Matt:
+%  a.  LOH can be real or an amplification artifact
+%  b.  isn't penetrance more about phenotype?  i would use polymorphic (for partially penetrant) and fixed (for fully penetrant)
+%  c.  i think the main reason for lack of multi-sample sequencing is that TCGA and other have used pathology specimens (i.e. there is only one per patient) out of convenience
 
@@
 In general, we would like to know how often these errors occur.
-One way of investigating this question would be to use PCR to verify each individual mutation.
+One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
 However, this method is expensive and time consuming.
@@
 
-TCGA's GBM data set includes 2 technical replicates each of 55 tumors.
+% Matt:
+%  a.  the first citation looks weird in the PDF.  it says (Network, 2008), which I think should be (TCGA Network, 2008)
+
+TCGA's GBM data set includes 2 technical replicates for each of 55 tumors.
 In this case, the technical replicates are not identical.
@@
 The other five are additional filters that we developed for this project.
-Each of these five filters represents an aspect of the data or putative mutation that is generally thought to indicate that a given SNP call is a false positive.
+Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNP call is a false positive.
 
@@
 Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-These data are enormous potential value to the research community, but their accuracy and reproducibility are unknown.
+These data are of enormous potential value to the research community, but their accuracy and reproducibility are unknown.
 Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
 
-We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
+We found that, on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
 The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.

### 2015-06-04 PAPER/paper.tex
matt's final edits to paper

@@
 %%% Title and author are here
-\title{Reproducibility of SNP-calling in multiple sequencing runs from single tumors}
+\title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
 \author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
@@
 
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling.
+We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNV-calling.
 We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
@@
 We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966).
-We further found that using filtering after SNP-calling removed the overlap completely.
+We further found that using filtering after SNV-calling removed the overlap completely.
 We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
@@
 
-% Matt: I would use SNV, rather than SNP, since SNP is more commonly used to refer to germline variants.  Or just use "somatic mutation calling pipeline"
 However, these and other efforts have produced very few results for patients in terms of improved treatments.
@@
 Another factor may be that the data are noisy.
-It is well known that sequencing and SNP-calling pipelines are not error-free.
-For example, different SNP-calling pipelines can return different results on the same data \citep{SNPcall}.
+It is well known that sequencing and SNV-calling pipelines are not error-free.
+For example, different SNV-calling pipelines can return different results on the same data \citep{SNPcall}.
 Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.
-Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls.
-Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity among replicates.
+Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls.
+Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue.
+And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
 
 We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
-For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNP-filtering approaches.
+For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches.
 We found significant overlap (around 50\%) between technical replicates, but also significant differences.
@@
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SNP filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
+Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
 
@@
 %(b) a description of what you found carrying out this experiment
-%
-% Matt:
-%  a.  LOH can be real or an amplification artifact
-%  b.  isn't penetrance more about phenotype?  i would use polymorphic (for partially penetrant) and fixed (for fully penetrant)
-%  c.  i think the main reason for lack of multi-sample sequencing is that TCGA and other have used pathology specimens (i.e. there is only one per patient) out of convenience
 
@@
 Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}.
-Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or completely absent.
+Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation may appear fixed or completely absent.
 In general, we would like to know how often these errors occur.
@@
 A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates.
-Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
+Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
+(Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
 Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
-
-% Matt:
-%  a.  the first citation looks weird in the PDF.  it says (Network, 2008), which I think should be (TCGA Network, 2008)
 
@@
 One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would expect any fully penetrant mutation to appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one.
+Despite the difference in sequencing protocol, we would expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
 Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
-We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.
+We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNVs per sample, than the corresponding WGS samples.
 
@@
 
-\subsection*{Does more sophisticated SNP filtering software increase or decrease the degree of similarity between replicates?}
+\subsection*{Does more sophisticated SNV filtering software increase or decrease the degree of similarity between replicates?}
 
@@
 In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts.
-Table~\ref{tab:filters} lists eight distinct SNP filters we evaluated.
+Table~\ref{tab:filters} lists eight distinct SNV filters we evaluated.
 The first three are based on the quality scores generated by GATK and SomaticSniper.
@@
 The other five are additional filters that we developed for this project.
-Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNP call is a false positive.
+Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNV call is a false positive.
 
@@
 We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
-The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.
+(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
+We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
+The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
 
-We next asked how many of the SNPs removed by each filter were SNPs present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNPs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
+We next asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
+Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
 To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
+We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
 By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
@@
 Filtering the raw computational data using both quality scores from GATK and SomaticSniper, as well as five additional custom filters, eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNPs.
+To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNVs.
 For example, LOH in this case is more likely to be an error than an actual mutation.
@@
 Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative SNPs within a 10 bp window of putative indels or other putative SNPs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
+Two of these three filters, those removing putative SNVs within a 10 bp window of putative indels or other putative SNVs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
 The third removes overlap with dbGaP.
@@
 Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNP.
-This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNP might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
+This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
 Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
-Having an independent measure of confidence in SNP calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
+Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
 
-Second, we studied only one particular SNP caller, SomaticSniper \citep{SomaticSniper}.
-There are other, equally widely used SNP callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
-Also, we did not even consider every possible quality metric available in the SNP-caller we did choose.
+Second, we studied only one particular SNV caller, SomaticSniper \citep{SomaticSniper}.
+There are other, equally widely used SNV callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
+Also, we did not even consider every possible quality metric available in the SNV-caller we did choose.
 For example, we did not look at the effects of the SomaticScore or the GATK quality score individually.
-In the future, it would be worthwhile to evaluate other SNP-callers and other quality metrics.
+In the future, it would be worthwhile to evaluate other SNV-callers and other quality metrics.
 
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNPs in WGS samples and about 30\% in WGA samples.
-The overlap exists even for samples with a high number of putative SNPs, suggesting that some GBMs may have significantly more somatic mutation than others.
-While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters that are commonly applied to validate SNPs.
+We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNVs in WGS samples and about 30\% in WGA samples.
+The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters that are commonly applied to validate SNVs.
 We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest that when mutation validation by PCR is not an option, only the filters that removed little overlap between samples should be used for computational SNP validation.
+We suggest that when mutation validation by PCR is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
 
@@
 After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNPs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
+We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
 The eight filters were enacted as follows:
@@
 This recommendation is discussed in the Results and Discussion.
-\item[10bp-SNP:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-SNP variants using a custom python script available in the public git repository.
+\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+We excluded 10bp-SNV variants using a custom python script available in the public git repository.
 This recommendation is discussed in the Results and Discussion.
@@
 	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
-	No. putative mutations removed, 10bp-SNP & 46 & 19 & 0 & 1,515 & 115 \\
+	No. putative mutations removed, 10bp-SNV & 46 & 19 & 0 & 1,515 & 115 \\
 	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
@@
 	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
-	\% overlap removed, 10bp-SNP & 3\% & 2\% & 0\% & 20\% & 3\% \\
+	\% overlap removed, 10bp-SNV & 3\% & 2\% & 0\% & 20\% & 3\% \\
 	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
@@
 \begin{table}
-\caption{\textbf{SNP prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
+\caption{\textbf{SNV prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
 \label{tab:filters}
@@
 	\hline
-	GATK & GATK & Removes putative SNPs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration) \\
-	SS & SomaticSniper & Removes putative SNPs with a SomaticScore less than 40 \\
-	VAQ & SomaticSniper & Removes putative SNPs with SomaticSniper Varaint Allele Quality scores less than 20 \\
-	LOH & SomaticSniper, python & Removes putative SNPs that are identified as loss of heterozygosity \\
-	10bp-SNP & python & Removes putative SNPs located within a 10 bp window of any other putative SNP \\
-	10bp-INDEL & python & Removes putative SNPs located within a 10 bp window of indels \\
-	dbSNP & python & Removes putative SNPs that overlap with dbSNP coverage \\
-	$<$10\% & python & Removes putative SNPs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10\% \\
+	GATK & GATK & Removes putative SNVs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration) \\
+	SS & SomaticSniper & Removes putative SNVs with a SomaticScore less than 40 \\
+	VAQ & SomaticSniper & Removes putative SNVs with SomaticSniper Varaint Allele Quality scores less than 20 \\
+	LOH & SomaticSniper, python & Removes putative SNVs that are identified as loss of heterozygosity \\
+	10bp-SNV & python & Removes putative SNVs located within a 10 bp window of any other putative SNV \\
+	10bp-INDEL & python & Removes putative SNVs located within a 10 bp window of indels \\
+	dbSNP & python & Removes putative SNVs that overlap with dbSNP coverage \\
+	$<$10\% & python & Removes putative SNVs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10\% \\
 \end{tabular}
@@
 \includegraphics[width=5in]{C282_v_C484.pdf} }
-\caption{\textbf{Number of putative SNPs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+\caption{\textbf{Number of putative SNVs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
@@
 \includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNPs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNVs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
 \label{fig:unfiltered_total_muts}
@@
 \includegraphics[width=6in]{boxplot_number_filtered.pdf} }
-\caption{\textbf{Number of putative SNPs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
-In all cases, the LOH and VAQ filters removed the majority of the putative SNPs for a given sample.}
+\caption{\textbf{Number of putative SNVs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
+In all cases, the LOH and VAQ filters removed the majority of the putative SNVs for a given sample.}
 \label{fig:boxplot_number_filtered}
@@
 \caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
-Filters removing LOH and putative SNPs with low VAQ removed a significant percentage of the overlap; nothing else did.}
+Filters removing LOH and putative SNVs with low VAQ removed a significant percentage of the overlap; nothing else did.}
 \label{fig:boxplot_percent_overlap_filtered}
@@
 \includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
-\caption{\textbf{Percentage of replicate SNPs filtered out by LOH, as a function of the total number of overlapping SNPs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
-This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}), indicating that these two filters remove almost disjoint sets of putative SNPs.}
+\caption{\textbf{Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
 \label{fig:LOH_all}
@@
 \includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} }
-\caption{ \textbf{Percentage of replicate SNPs filtered out by VAQ, as a function of the total number of overlapping SNPs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
-This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}), indicating that these two filters remove almost disjoint sets of putative SNPs.}
+\caption{ \textbf{Percentage of replicate SNVs filtered out by VAQ, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
 \label{fig:VAQ_all}

### 2015-06-04 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r

@@ -24,8 +24,8 @@ dev.off()
 
 pdf( "../FIGURE_PDFS/boxplot_number_filtered.pdf", width=6, height=5 )
 par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
-axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNP", "10bp-INDEL", "dbSNP", "<10%" ) )
+boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNVs filtered, per sample" )
+axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 dev.off()
 
 ### still working on two problems:
@@ -52,5 +52,5 @@ dev.off()
 pdf( "../FIGURE_PDFS/boxplot_percent_overlap_filtered.pdf", width=6, height=5 )
 par(bty='n', mar=c(5.1,4.1,1.1,0))
 boxplot( overlap_data[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="% of unfiltered overlap filtered, per sample", pch=20 )
-axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNP", "10bp-INDEL", "dbSNP", "<10%" ) )
+axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 dev.off()

### 2015-06-04 FIGURES/FIGURE_SCRIPTS/histograms_of_mut_frequency.r

@@ -25,7 +25,7 @@ dev.off()
 #png( "../FIGURE_PDFS/C282_v_C484.png" )
 pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
 par(bty='n')
-plot(data$C282_count, data$C484_count, log='xy', pch=20, xlab="No. of putative SNPs, WGA", ylab="No. of putative SNPs, WGS" )
+plot(data$C282_count, data$C484_count, log='xy', pch=20, xlab="No. of putative SNVs, WGA", ylab="No. of putative SNVs, WGS" )
 abline(0,1)
 dev.off()
 

### 2015-06-04 FIGURES/FIGURE_SCRIPTS/histograms_of_unfiltered_overlap.R

@@ -25,7 +25,7 @@ dev.off()
 pdf( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.pdf" )
 # when necessary, use the following title: No. of putative SNPs does not correlate with % WGS/WGA overlap
 par(bty = 'n')
-plot( data$TOTAL, data$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="log of no. of putative SNPs (unfiltered)", ylab="percent overlap between WGS and WGA (unfiltered)" )
+plot( data$TOTAL, data$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="log of no. of putative SNVs (unfiltered)", ylab="percent overlap between WGS and WGA (unfiltered)" )
 abline( 0,1 )
 dev.off()
 

### 2015-06-04 PAPER/paper.tex
Define SNV on first use.

@@
 
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNV-calling.
+We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in calling single-nucleotide variants (SNVs).
 We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
@@
 Another factor may be that the data are noisy.
-It is well known that sequencing and SNV-calling pipelines are not error-free.
-For example, different SNV-calling pipelines can return different results on the same data \citep{SNPcall}.
+It is well known that sequencing and variant-calling pipelines are not error-free.
+For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}.
 Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.

### 2015-06-05 PAPER/paper.tex
added to bibliography

@@
 \author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
-\author[3]{Matthew C. Cowperthwaite}
-\author[1,2]{Claus O. Wilke} 
+\author[2,3]{Matthew C. Cowperthwaite}
+\author[1,4]{Claus O. Wilke} 
 \affil[1]{The University of Texas at Austin, Cell \& Molecular Biology}
-\affil[2]{The University of Texas at Austin, Integrative Biology}
-\affil[3]{St. David's NeuroTexas Institute Research Foundation}
+\affil[2]{St. David's NeuroTexas Institute Research Foundation}
+\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin}
+\affil[4]{The University of Texas at Austin, Integrative Biology}
 
@@
 Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls.
+One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}.
 Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue.

### 2015-06-23 MAKE_VCF/PARALLEL/12_batchFastExon.sh
removed some unused scripts for processing bamfiles in parallel

@@ -1,180 +1,0 @@
-#!/bin/bash
-
-module load bwa/0.7.7
-module load samtools
-module load picard
-module load python
-module load pylauncher
-module load gatk/2.5.2
-
-# these are command line arguments; pfx is the sample name (the TCGA barcode; everything before .bam or .fastq)
-pfx=$1
-splitSize=$2
-batchSize=$3
-coresPerRun=$4
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-queue="normal"
-
-# 1. .bam to .fastq conversion
-cd /scratch/00001/mattcowp/dakota
-# echo $pfx
-cd $pfx
-echo "$pfx:" `date` > ../$pfx.log
-
-# we have to convert a bam into a fastq
-# might have to change path to picard; check this in idev
-# echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
-# (java -d64 -Xmx4g -jar /opt/apps/picard/1.98/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
-# echo "SamToFastQ done: " `date` >> ../$pfx.log
-
-# exit: this is the test point for bam to fastq conversion
-# I tested it successfully on sample TCGA-02-2468-01A-01D-1494-08.5
-
-# 2. split and clean up the fastq files
-# which is done by the script split_fastq_threaded.py
-rm split.*
-echo "" >> ../$pfx.log
-echo "python $HOME/GBM_genomics/MAKE_VCF/split_fastq_threaded.py $pfx $splitSize" > split.script
-echo "Waiting while splitting and cleaning the fastq file."  `date` >> ../$pfx.log
-python $HOME/GBM_genomics/MAKE_VCF/launcher.py split.script 2
-echo "Fastq split finished :" `date` >> ../$pfx.log
-rm -rf pylauncher_tmp*
-
-# I guess we're making a note to show that we finished running pylauncher here? 
-echo "-----"
-echo "$pfx:" `date` > $pfx.log
-
-# we need to make sure we are in the right working directory, just cause it's super important
-if [[ "$PWD" != *"$pfx"* ]];
-then
[…]

### 2015-06-23 MAKE_VCF/PARALLEL/batchFastExon.sh

@@ -1,200 +1,0 @@
-#!/bin/bash
-
-# this is heavily modified and meant for making the 12 2 files
-
-module load bwa/0.7.4
-module load samtools
-module load picard
-module load python
-module load pylauncher
-module load gatk/2.5.2
-
-# these are command line arguments; pfx is the sample name (the TCGA barcode; everything before .bam or .fastq)
-pfx=$1
-splitSize=$2
-batchSize=$3
-coresPerRun=$4
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-queue="normal"
-
-# 0. Decrypt the bam file and extract, split and filter the reads (remove unpaired reads)
-# Note:   I added a optional initial flag, "UNPACK_ONLY", that will run the decryption and SamToFastQ steps.
-#         There are very low memory steps, but are also very slow.  Therefore, it's much more efficient
-#         to unpack everything with 8 samples per node and then process everything else at 4 cores per
-#         node.  Unpack is roughly 3/4 of the total processing time.
-echo $1;
-
-if [[ "$1" == *"UNPACK"* ]];
-then
-
-    # shift to right command-lines args                                                                                                                                
-    pfx=$2
-    splitSize=$3
-    batchSize=$4
-    coresPerRun=$5
-    echo "$pfx:" `date` > $pfx.log
-    echo "UNPACK:  Sample: " $pfx  `date` >> $pfx.log
-
-    # make the dir and                                                                                                                                                  
-    mkdir $pfx
-    cd $pfx
-    # delete old bams
-    rm $pfx.bam
-    echo "Waiting for $pfx to decrypt from ../tcga_bams"  `date` >> ../$pfx.log
-    decrypt $pfx.bam ../tcga_bams $pkFile
-    echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
-    (java -d64 -Xmx4g -jar /opt/apps/picard/1.92/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
[…]

### 2015-06-23 MAKE_VCF/PARALLEL/launcher.py

@@ -1,14 +1,0 @@
-#!/usr/bin/env python
-
-import os
-import sys
-import pylauncher
-
-cmdLineFile = sys.argv[1]
-numCoresPerCmd = 4
-if len(sys.argv) == 3:
-    numCoresPerCmd = int(sys.argv[2])
-
-pylauncher.ClassicLauncher( cmdLineFile, cores=numCoresPerCmd )
-
-sys.exit()

### 2015-06-23 MAKE_VCF/PARALLEL/pt_2_only_batchFastExon.sh

@@ -1,167 +1,0 @@
-# this script begins only after the .fastq file (step 1 of the original batchFastExon) has been made.
-
-#!/bin/bash
-
-module load bwa/0.7.7
-module load samtools
-module load picard
-module load python
-module load pylauncher
-module load gatk/2.5.2
-
-# these are command line arguments; pfx is the sample name (the TCGA barcode; everything before .bam or .fastq)
-pfx=$1
-splitSize=$2
-batchSize=$3
-coresPerRun=$4
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-queue="normal"
-
-# 2. split and clean up the fastq files
-# which is done by the script split_fastq_threaded.py
-rm split.*
-echo "" >> ../$pfx.log
-echo "python $HOME/GBM_genomics/MAKE_VCF/split_fastq_threaded.py $pfx $splitSize" > split.script
-echo "Waiting while splitting and cleaning the fastq file."  `date` >> ../$pfx.log
-python $HOME/GBM_genomics/MAKE_VCF/launcher.py split.script 2
-echo "Fastq split finished :" `date` >> ../$pfx.log
-rm -rf pylauncher_tmp*
-
-# I guess we're making a note to show that we finished running pylauncher here? 
-echo "-----"
-echo "$pfx:" `date` > $pfx.log
-
-# we need to make sure we are in the right working directory, just cause it's super important
-if [[ "$PWD" != *"$pfx"* ]];
-then
-    echo "Changing directory to /scratch/00001/mattcowp/dakota/$pfx"  `date` >> $pfx.log
-    cd /scratch/00001/mattcowp/dakota/$pfx
-fi;
-
-# exit: working on debugging. Ran the job on this file: TCGA-02-2468-01A-01D-1494-08.6
-# log says good, checked in idev, and for this one, it all checks out
-
-#  check to ensure that we go an equal number of splits, if there
-# are no r1 or r2 files, then try to re-run split script
-numR1Files=`ls r1.* | wc -l`
-echo "Number of r1 files: " $numR1Files  `date` >> ../$pfx.log
-numR2Files=`ls r2.* | wc -l`
[…]

### 2015-06-23 MAKE_VCF/PARALLEL/revisions_1.sh

@@ -1,61 +1,0 @@
-# start this script in $SCRATCH/DOWNSAMPLING_DATA
-# this takes the downsampling data and makes a .bam, NOT A VCF
-
-module load bwa/0.7.4
-module load samtools
-module load python
-module load pylauncher
-
-# pfx is the sample name (everything before .bam or .fastq), in this case GSAF barcodes
-# split_size is 2,000,000
-pfx=$1
-splitSize=$2
-batchSize=$3
-coresPerRun=$4
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-queue="normal"
-
-# make a directory into which the .bam file will go, and move to it
-mkdir -p $pfx
-cd $pfx
-# split, and clean up the fastq files
-rm split.*
-
-echo "" >> ../$pfx.log
-echo "python $SCRATCH/DOWNSAMPLING/split_fastq_threaded.py $pfx $splitSize"  `date` > split.script
-echo "Waiting while splitting and cleaning the fastq file."  `date` >> ../$pfx.log
-python $HOME/DOWNSAMPLING/launcher.py split.script 2
-echo "Fastq split finished :" `date` >> ../$pfx.log
-rm -rf pylauncher_tmp*
-
-echo "---fastq split and cleaned---"
-echo "$pfx:" `date` > $pfx.log
-
-#  check to ensure that we go an equal number of splits, if there                                              
-# are no r1 or r2 files, then try to re-run split script                                                       
-numR1Files=`ls r1.* | wc -l`
-echo "Number of r1 files: " $numR1Files  `date` >> ../$pfx.log
-numR2Files=`ls r2.* | wc -l`
-echo "Number of r2 files: " $numR2Files  `date` >> ../$pfx.log
-if [ $numR1Files -eq 0 -o $numR2Files -eq 0 ]
-then
-    echo "python $HOME/DOWNSAMPLING/split_fastq_threaded.py $pfx $splitSize"  `date` > split.script
-    echo "Initial split seems suspect:  r1=$numR1Files, r2=$numR2Files"  `date` >> $pfx.log
-    echo "Waiting while splitting and cleaning the fastq file."  `date` >> ../$pfx.log
-    python $HOME/DOWNSAMPLING/launcher.py split.script 2
-    echo "Fastq split finished :" `date` >> ../$pfx.log
-    rm -rf pylauncher_tmp*
-    numR1Files=`ls r1.* | wc -l`
[…]

### 2015-06-23 MAKE_VCF/PARALLEL/run_writer.sh

@@ -1,24 +1,0 @@
-#!/bin/bash
-
-for r in $@
-do
-    echo '#!/bin/bash' > $r.sbat
-    echo '' >> $r.sbat
-    echo '#SBATCH -J' $r >> $r.sbat
-    echo '#SBATCH -o' $r.out >> $r.sbat
-    echo '#SBATCH -e' $r.err >> $r.sbat
-    echo '#SBATCH -n 16' >> $r.sbat
-    echo '#SBATCH -p normal' >> $r.sbat
-    echo '#SBATCH -A mattcowp' >> $r.sbat
-    echo '#SBATCH -t 24:00:00' >> $r.sbat
-
-    echo '' >> $r.sbat
-    echo 'cd $SCRATCH/TCGA_NextGen_Analysis/hg19_bams' >> $r.sbat
-    echo 'echo $PWD' >> $r.sbat
-    echo '' >> $r.sbat
-
-    echo 'echo 'Run:' `date`' >> $r.sbat
-    echo "bash $r" >> $r.sbat
-    echo 'echo 'Done: ' `date`' >> $r.sbat
-
-done;

### 2015-06-23 MAKE_VCF/PARALLEL/split_fastq_threaded.py

@@ -1,96 +1,0 @@
-#!/usr/bin/python
-
-# run command: python $HOME/GBM_genomics/MAKE_VCF/split_fastq_threaded.py $pfx $splitSize
-# $pfx is a barcode for the sample, and $splitSize is a command line argument
-
-import os
-import sys
-import Queue
-import threading
-import time
-from itertools import groupby, count
-from Bio import SeqIO
-
-def fastqSplit( fileName, pt, chunk=1000000 ):
-    outFilePfx = "r" + str(pt)
-    print fileName
-    print outFilePfx
-    with open( fileName, 'r' ) as datafile:
-        groups = groupby( datafile, key=lambda k, line=count(): next(line) // chunk )
-        n = 0
-        for k, group in groups:
-            with open( outFilePfx+"."+str(n), 'w' ) as outfile:
-                outfile.write(''.join(group))
-                outfile.close()
-            n += 1
-    datafile.close()
-
-def removeUnpairedMates( sample, ptNum ):
-    inFastQ = sample + "." + str(ptNum) + ".fastq"
-    outFastQ = sample + "." + str(ptNum) + ".clean.fastq"
-    print "Cleaning %s --> %s" % ( inFastQ, outFastQ )
-    cleanFastq = open( outFastQ, 'w' )
-    for rec in SeqIO.parse( open(inFastQ), 'fastq' ):
-        mateInfo = rec.name[-2:]
-        # print rec.name
-        # print mateInfo
-        if mateInfo == '/1' or mateInfo == '/2':
-            SeqIO.write( rec, cleanFastq, 'fastq' )
-    cleanFastq.close()
-    # rename the "clean" file to the starting file
-    os.rename( outFastQ, inFastQ )
-
-# set up the global threading queue
-threadQueue = Queue.Queue()
-class FastQThread( threading.Thread ):
-
-    def __init__( self, queue ):
-        threading.Thread.__init__( self )
-        self.queue = queue
-
-    def run( self ):
-        while True:
-            # get the sample information from the queue
-            info = self.queue.get()
-            function = info[0]
-            args = info[1:]
-            try:
-                function( *args )
-            except:
-                print "something wrong in " + self.name
-                self.queue.task_done()
-            # send the done signal
-            self.queue.task_done()
-
-startTime = time.time()
-def main():
-
-    # spawn two threads
-    for i in range( 2 ):
-        t = FastQThread( threadQueue )
[…]

### 2015-06-23 MAKE_VCF/SNPcalling.sh
moved some files around

@@ -1,0 +1,123 @@
+#!/bin/bash
+
+## command line variables (this is taken from the previous script, GATK_SNPs.sh)
+pfx=$1
+tumor=$2
+blood=$3
+SS=$4
+MT=$5
+ST=$6
+VS=$7
+
+## make paths to working directory, data directory, the data, and other important places and files
+dataDir="/scratch/00001/mattcowp/dakota"
+bloodDir="/scratch/00001/mattcowp/dakota/normal"
+workDir="$dataDir/$pfx$tumor.WORK"
+tumorbam="$workDir/$pfx$tumor.realn.recal.bam"
+normalbam="$workDir/$pfx$blood.realn.recal.bam"
+
+## make paths to files you will need, including hg19, 
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+
+## additional references for Strelka
+
+## additional references for VarScan
+
+#####################################
+## First SNP-Caller: SomaticSniper ##
+#####################################
+
+## run SomaticSniper
+function runSS {
+	cd $workDir
+	echo "run SomaticSniper: " `date` >> ./$pfx.SNPcalls.log
+	$somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f $hgReference $tumorbam $normalbam $pfx$tumor.SS.vcf >> ./$pfx.SNPcalls.log
+	echo "SomaticSniper finished: " `date` >> ./$pfx.SNPcalls.log
+}
+
+###############################
+## Second SNP-Caller: MuTect ##
+###############################
+
+## run MuTect
+function runMT {
+	
+	# go to the working directory
+	cd $workDir
+	
+	# set variables
+	muTectDir="/work/01839/dakotaz/local/bin/MuTect"
+	muTectJar="$muTectDir/muTect-1.1.1.jar"
+	myRefDir="/work/01839/dakotaz/referenceDS"
+	COSMIC="$myRefDir/Cosmic.hg19.vcf"
+	dbSNP="$refDir/dbsnp_137.b37.vcf"
+	
+	# unzip dbSNP in scratch (because it is huge)
+	# gunzip -c $dbSNPgz > $workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf
+	# dbSNP="$workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf"
+	
+	# and finally the run command...
[…]

### 2015-06-23 MAKE_VCF/SNPs_GATK.sh

@@ -1,0 +1,164 @@
+#!/bin/bash
+
+# load modules (gatk current default is 2.7.2)
+# module load jdk64 # this line added just for 190 run
+module load gatk
+module load samtools/0.1.19 # used older version of samtools only on the 190 run
+module load picard 
+
+## this script does the GATK processing and variant calibration on all of the out.sorted.matefixed.bam (and corresponding .bam.bai) files
+
+dataDir="/scratch/00001/mattcowp/dakota"
+cd $dataDir
+
+echo "Begin $0: `date`">>$tumorPfx.gatk.log
+echo "-----------------------------------------------------------------------------">>$tumorPfx.gatk.log
+
+## get sample names and file names, and make sure the two files have the same sample name
+tumorPfx=$1
+bloodPfx=$2
+tumorSamp=${tumorPfx:0:17}
+bloodSamp=${bloodPfx:0:17}
+tumorDir="$tumorPfx.WORK"
+# echo $tumor_dir
+echo $tumorPfx $tumorSamp &>>$tumorPfx.gatk.log
+echo $bloodPfx $bloodSamp &>>$tumorPfx.gatk.log
+if [ $tumorSamp != $bloodSamp ]; then
+	echo "Non-matching sample names:"
+	echo "	$tumorSamp"
+	echo "	$bloodSamp"
+	exit 1;
+fi
+
+## locations of relevant and needed files
+gatkJar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+dbSNP="$refDir/dbsnp_137.b37.vcf"
+hapMap="$refDir/hapmap_3.3.b37.vcf"
+G1000HiConfSNPs="$refDir/1000G_phase1.snps.high_confidence.b37.vcf"
+G1000Phase1Indels="$refDir/1000G_phase1.indels.b37.vcf"
+G1000OmniSNPs="$refDir/1000G_omni2.5.b37.vcf"
+G1000Mills="$refDir/Mills_and_1000G_gold_standard.indels.b37.vcf"
+
+echo "Generating somatic calls for $tumorPfx" &>>$tumorPfx.gatk.log
+
+# make a temporary directory in which this will actually run
+mkdir $tumorDir
[…]

### 2015-06-23 MAKE_VCF/SNPs_GATK_NoNormal.sh

@@ -1,0 +1,99 @@
+#!/bin/bash
+
+# load modules (gatk current default is 2.7.2)
+module load jdk64
+module load gatk
+module load samtools/0.1.19
+module load picard 
+
+## this script does the GATK processing and variant calibration on all of the sorted .bam and .bam.bai files
+
+dataDir="/scratch/00001/mattcowp/dakota"
+cd $dataDir
+
+## get sample names and file names, and make sure the two files have the same sample name
+tumorPfx1=$1
+tumorPfx2=$2
+sampPfx=$3
+tumorSamp1=${tumorPfx1:0:17}
+tumorSamp2=${tumorPfx2:0:17}
+tumorDir="$sampPfx.NoNormal.WORK"
+# echo $tumor_dir
+echo $tumorPfx1 $tumorSamp1
+echo $tumorPfx2 $tumorSamp2
+if [ $tumorSamp1 != $tumorSamp2 ]; then
+	echo "Non-matching sample names:"
+	echo "	$tumorSamp1"
+	echo "	$tumorSamp2"
+	exit 1;
+fi
+
+## locations of relevant and needed files
+gatkJar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+dbSNP="$refDir/dbsnp_137.b37.vcf"
+hapMap="$refDir/hapmap_3.3.b37.vcf"
+G1000HiConfSNPs="$refDir/1000G_phase1.snps.high_confidence.b37.vcf"
+G1000Phase1Indels="$refDir/1000G_phase1.indels.b37.vcf"
+G1000OmniSNPs="$refDir/1000G_omni2.5.b37.vcf"
+G1000Mills="$refDir/Mills_and_1000G_gold_standard.indels.b37.vcf"
+
+echo "Generating somatic calls for $sampPfx" > $sampPfx.gatk.log
+
+# make a temporary directory in which this will actually run
+mkdir $tumorDir
+cd $tumorDir
+
+# remove duplicate reads from bamfiles, leaving only the read with the highest map quality
+samtools rmdup ../$tumorPfx1/$tumorPfx1.out.sorted.matefixed.bam $tumorPfx1.dedup.bam &
[…]

### 2015-06-23 MAKE_VCF/align_fastq.sh

@@ -1,0 +1,54 @@
+#!/bin/bash
+
+# the purpose of this script is to align the two fastq files of a TCGA sample (paired ends) to hg19
+
+module load bwa
+module load samtools
+
+# pfx is the sample name
+pfx=$1
+tissue=$2
+outprefix="$pfx.out"
+queue="normal"
+# reference hg19
+refDir="/work/00001/mattcowp/Hs_reference_datasets"
+hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
+# what data we're running on...
+data_dir="/scratch/00001/mattcowp/dakota"
+if [ $tissue = "blood" ]
+then
+	data_dir="/scratch/00001/mattcowp/dakota/normal"
+fi
+fastq1="$data_dir/$pfx/$pfx.1.fastq"
+fastq2="$data_dir/$pfx/$pfx.2.fastq"
+
+# move to working directory
+cd $data_dir/$pfx
+
+# the bwa manual says this is important
+# but Matt says take it out...
+# bwa index -p RefSeqbwaidx -a bwtsw $hgReference
+
+# bwa aln first fastq
+bwa aln -q 30 -t 8 $hgReference $fastq1 > $fastq1.aln.sai 2>$fastq1.aln.bwa.log &
+# bwa aln second fastq
+bwa aln -q 30 -t 8 $hgReference $fastq2 > $fastq2.aln.sai 2>$fastq2.aln.bwa.log &
+wait
+echo "bwa aln for $pfx done"
+
+# need to specify read groups (RG), which is the argument of  -r in bwa sempe
+# first get the library information from the TCGA bam:
+LB=`samtools view -H ./$pfx.bam | grep -m 1 '@RG' | awk '{print substr($5,4) }'`
+if [ ${#LB} -eq 0 ] # this is the case if the bam has no library information at all
+then
+	LB="Catch-1"
+fi
+# then set the @RG header
+RG="@RG\tID:1\tPL:ILLUMINA\tSM:$pfx\tLB:$LB\tDS:ref=hg19"
+
+# bwa sampe | samtools view
+(bwa sampe -a 600 -P -r "$RG" $hgReference $fastq1.aln.sai $fastq2.aln.sai $fastq1 $fastq2 | samtools view -bSh -o $outprefix.bam -) 2>$outprefix.log
+# samtools sort
+samtools sort -@ 16 $outprefix.bam $outprefix.sorted 2>>$outprefix.log
[…]

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/SNPcalling.sh
removed a redundant directory

@@ -1,123 +1,0 @@
-#!/bin/bash
-
-## command line variables (this is taken from the previous script, GATK_SNPs.sh)
-pfx=$1
-tumor=$2
-blood=$3
-SS=$4
-MT=$5
-ST=$6
-VS=$7
-
-## make paths to working directory, data directory, the data, and other important places and files
-dataDir="/scratch/00001/mattcowp/dakota"
-bloodDir="/scratch/00001/mattcowp/dakota/normal"
-workDir="$dataDir/$pfx$tumor.WORK"
-tumorbam="$workDir/$pfx$tumor.realn.recal.bam"
-normalbam="$workDir/$pfx$blood.realn.recal.bam"
-
-## make paths to files you will need, including hg19, 
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-
-## additional references for Strelka
-
-## additional references for VarScan
-
-#####################################
-## First SNP-Caller: SomaticSniper ##
-#####################################
-
-## run SomaticSniper
-function runSS {
-	cd $workDir
-	echo "run SomaticSniper: " `date` >> ./$pfx.SNPcalls.log
-	$somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f $hgReference $tumorbam $normalbam $pfx$tumor.SS.vcf >> ./$pfx.SNPcalls.log
-	echo "SomaticSniper finished: " `date` >> ./$pfx.SNPcalls.log
-}
-
-###############################
-## Second SNP-Caller: MuTect ##
-###############################
-
-## run MuTect
-function runMT {
-	
-	# go to the working directory
-	cd $workDir
-	
-	# set variables
-	muTectDir="/work/01839/dakotaz/local/bin/MuTect"
-	muTectJar="$muTectDir/muTect-1.1.1.jar"
-	myRefDir="/work/01839/dakotaz/referenceDS"
-	COSMIC="$myRefDir/Cosmic.hg19.vcf"
-	dbSNP="$refDir/dbsnp_137.b37.vcf"
-	
-	# unzip dbSNP in scratch (because it is huge)
-	# gunzip -c $dbSNPgz > $workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf
-	# dbSNP="$workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf"
-	
-	# and finally the run command...
[…]

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/SNPs_GATK.sh

@@ -1,164 +1,0 @@
-#!/bin/bash
-
-# load modules (gatk current default is 2.7.2)
-# module load jdk64 # this line added just for 190 run
-module load gatk
-module load samtools/0.1.19 # used older version of samtools only on the 190 run
-module load picard 
-
-## this script does the GATK processing and variant calibration on all of the out.sorted.matefixed.bam (and corresponding .bam.bai) files
-
-dataDir="/scratch/00001/mattcowp/dakota"
-cd $dataDir
-
-echo "Begin $0: `date`">>$tumorPfx.gatk.log
-echo "-----------------------------------------------------------------------------">>$tumorPfx.gatk.log
-
-## get sample names and file names, and make sure the two files have the same sample name
-tumorPfx=$1
-bloodPfx=$2
-tumorSamp=${tumorPfx:0:17}
-bloodSamp=${bloodPfx:0:17}
-tumorDir="$tumorPfx.WORK"
-# echo $tumor_dir
-echo $tumorPfx $tumorSamp &>>$tumorPfx.gatk.log
-echo $bloodPfx $bloodSamp &>>$tumorPfx.gatk.log
-if [ $tumorSamp != $bloodSamp ]; then
-	echo "Non-matching sample names:"
-	echo "	$tumorSamp"
-	echo "	$bloodSamp"
-	exit 1;
-fi
-
-## locations of relevant and needed files
-gatkJar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-dbSNP="$refDir/dbsnp_137.b37.vcf"
-hapMap="$refDir/hapmap_3.3.b37.vcf"
-G1000HiConfSNPs="$refDir/1000G_phase1.snps.high_confidence.b37.vcf"
-G1000Phase1Indels="$refDir/1000G_phase1.indels.b37.vcf"
-G1000OmniSNPs="$refDir/1000G_omni2.5.b37.vcf"
-G1000Mills="$refDir/Mills_and_1000G_gold_standard.indels.b37.vcf"
-
-echo "Generating somatic calls for $tumorPfx" &>>$tumorPfx.gatk.log
-
-# make a temporary directory in which this will actually run
-mkdir $tumorDir
[…]

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/SNPs_GATK_NoNormal.sh

@@ -1,99 +1,0 @@
-#!/bin/bash
-
-# load modules (gatk current default is 2.7.2)
-module load jdk64
-module load gatk
-module load samtools/0.1.19
-module load picard 
-
-## this script does the GATK processing and variant calibration on all of the sorted .bam and .bam.bai files
-
-dataDir="/scratch/00001/mattcowp/dakota"
-cd $dataDir
-
-## get sample names and file names, and make sure the two files have the same sample name
-tumorPfx1=$1
-tumorPfx2=$2
-sampPfx=$3
-tumorSamp1=${tumorPfx1:0:17}
-tumorSamp2=${tumorPfx2:0:17}
-tumorDir="$sampPfx.NoNormal.WORK"
-# echo $tumor_dir
-echo $tumorPfx1 $tumorSamp1
-echo $tumorPfx2 $tumorSamp2
-if [ $tumorSamp1 != $tumorSamp2 ]; then
-	echo "Non-matching sample names:"
-	echo "	$tumorSamp1"
-	echo "	$tumorSamp2"
-	exit 1;
-fi
-
-## locations of relevant and needed files
-gatkJar="$TACC_GATK_DIR/GenomeAnalysisTK.jar"
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-dbSNP="$refDir/dbsnp_137.b37.vcf"
-hapMap="$refDir/hapmap_3.3.b37.vcf"
-G1000HiConfSNPs="$refDir/1000G_phase1.snps.high_confidence.b37.vcf"
-G1000Phase1Indels="$refDir/1000G_phase1.indels.b37.vcf"
-G1000OmniSNPs="$refDir/1000G_omni2.5.b37.vcf"
-G1000Mills="$refDir/Mills_and_1000G_gold_standard.indels.b37.vcf"
-
-echo "Generating somatic calls for $sampPfx" > $sampPfx.gatk.log
-
-# make a temporary directory in which this will actually run
-mkdir $tumorDir
-cd $tumorDir
-
-# remove duplicate reads from bamfiles, leaving only the read with the highest map quality
-samtools rmdup ../$tumorPfx1/$tumorPfx1.out.sorted.matefixed.bam $tumorPfx1.dedup.bam &
[…]

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/align_fastq.sh

@@ -1,54 +1,0 @@
-#!/bin/bash
-
-# the purpose of this script is to align the two fastq files of a TCGA sample (paired ends) to hg19
-
-module load bwa
-module load samtools
-
-# pfx is the sample name
-pfx=$1
-tissue=$2
-outprefix="$pfx.out"
-queue="normal"
-# reference hg19
-refDir="/work/00001/mattcowp/Hs_reference_datasets"
-hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
-# what data we're running on...
-data_dir="/scratch/00001/mattcowp/dakota"
-if [ $tissue = "blood" ]
-then
-	data_dir="/scratch/00001/mattcowp/dakota/normal"
-fi
-fastq1="$data_dir/$pfx/$pfx.1.fastq"
-fastq2="$data_dir/$pfx/$pfx.2.fastq"
-
-# move to working directory
-cd $data_dir/$pfx
-
-# the bwa manual says this is important
-# but Matt says take it out...
-# bwa index -p RefSeqbwaidx -a bwtsw $hgReference
-
-# bwa aln first fastq
-bwa aln -q 30 -t 8 $hgReference $fastq1 > $fastq1.aln.sai 2>$fastq1.aln.bwa.log &
-# bwa aln second fastq
-bwa aln -q 30 -t 8 $hgReference $fastq2 > $fastq2.aln.sai 2>$fastq2.aln.bwa.log &
-wait
-echo "bwa aln for $pfx done"
-
-# need to specify read groups (RG), which is the argument of  -r in bwa sempe
-# first get the library information from the TCGA bam:
-LB=`samtools view -H ./$pfx.bam | grep -m 1 '@RG' | awk '{print substr($5,4) }'`
-if [ ${#LB} -eq 0 ] # this is the case if the bam has no library information at all
-then
-	LB="Catch-1"
-fi
-# then set the @RG header
-RG="@RG\tID:1\tPL:ILLUMINA\tSM:$pfx\tLB:$LB\tDS:ref=hg19"
-
-# bwa sampe | samtools view
-(bwa sampe -a 600 -P -r "$RG" $hgReference $fastq1.aln.sai $fastq2.aln.sai $fastq1 $fastq2 | samtools view -bSh -o $outprefix.bam -) 2>$outprefix.log
-# samtools sort
-samtools sort -@ 16 $outprefix.bam $outprefix.sorted 2>>$outprefix.log
[…]

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/bam_to_fastq.sh

@@ -1,21 +1,0 @@
-#!/bin/bash
-
-# this script takes a TCGA .bam (aligned to hg18) and converts it to a pair of unaligned .fastq files
-
-module load samtools
-module load picard
-
-# these are command line arguments; pfx is the sample name (the TCGA barcode)
-pfx=$1
-wrk_dir=$2
-queue="normal"
-
-# 1. pfx.bam to pfx.fastq conversion
-# cd /scratch/00001/mattcowp/dakota
-cd $wrk_dir
-cd $pfx
-echo "$pfx:" `date` > ../$pfx.log
-
-echo "Running SamToFastQ on $pfx...."  `date` >> ../$pfx.log
-(java -d64 -Xmx4g -jar /opt/apps/picard/1.98/SamToFastq.jar I=$pfx.bam F=$pfx.1.fastq F2=$pfx.2.fastq 2>&1) >> $pfx\_SamToFastQ.log
-echo "SamToFastQ done: " `date` >> ../$pfx.log

### 2015-06-23 MAKE_VCF/IN_SEQUENCE/gtdownload.sh

@@ -1,11 +1,0 @@
-## this file tests the gtdownload command...
-## /work/01839/dakotaz/cghub/bin/gtdownload is the executable
-## -v is the verbose option so that I know what the program is doing
-## -c is the credential to download protected data, good until May 1, 2016
-## -p is the path to where we want to store the file being downloaded
-## -d is the file to download, tested here with a one file manifest
-
-manifest=$1
-path=$2
-
-time /work/01839/dakotaz/cghub/bin/gtdownload -v -c /work/01839/dakotaz/cghub.key -p $path -d $manifest

### 2015-06-23 MAKE_VCF/SNPcalling.sh
took out some un-used code in this project

@@ -4,10 +4,6 @@
 pfx=$1
 tumor=$2
 blood=$3
-SS=$4
-MT=$5
-ST=$6
-VS=$7
 
 ## make paths to working directory, data directory, the data, and other important places and files
 dataDir="/scratch/00001/mattcowp/dakota"
@@ -20,104 +16,8 @@ normalbam="$workDir/$pfx$blood.realn.recal.bam"
 refDir="/work/00001/mattcowp/Hs_reference_datasets"
 hgReference="$refDir/Homo_sapiens.GRCh37.72.dna.fa"
 
-## additional references for Strelka
-
-## additional references for VarScan
-
-#####################################
-## First SNP-Caller: SomaticSniper ##
-#####################################
-
 ## run SomaticSniper
-function runSS {
-	cd $workDir
-	echo "run SomaticSniper: " `date` >> ./$pfx.SNPcalls.log
-	$somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f $hgReference $tumorbam $normalbam $pfx$tumor.SS.vcf >> ./$pfx.SNPcalls.log
-	echo "SomaticSniper finished: " `date` >> ./$pfx.SNPcalls.log
-}
-
-###############################
-## Second SNP-Caller: MuTect ##
-###############################
-
-## run MuTect
-function runMT {
-	
-	# go to the working directory
-	cd $workDir
-	
-	# set variables
-	muTectDir="/work/01839/dakotaz/local/bin/MuTect"
-	muTectJar="$muTectDir/muTect-1.1.1.jar"
-	myRefDir="/work/01839/dakotaz/referenceDS"
-	COSMIC="$myRefDir/Cosmic.hg19.vcf"
-	dbSNP="$refDir/dbsnp_137.b37.vcf"
-	
-	# unzip dbSNP in scratch (because it is huge)
-	# gunzip -c $dbSNPgz > $workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf
-	# dbSNP="$workDir/dbSNP.human_9606_b142_GRCh37p13.All.vcf"
-	
-	# and finally the run command...
-	java -Xmx12g -jar $muTectJar --analysis_type MuTect --reference_sequence $hgReference --cosmic $COSMIC --dbsnp $dbSNP --input_file:normal $normalbam --input_file:tumor $tumorbam --out $pfx$tumor.MT.out --coverage_file $pfx$tumor.MTcoverage.wig.txt
-	
-	# clean up 
-	# rm $dbSNP
-	
-}
[…]

### 2015-06-23 MAKE_VCF/gtdownload.sh

@@ -1,5 +1,6 @@
-## this file tests the gtdownload command...
-## /work/01839/dakotaz/cghub/bin/gtdownload is the executable
+#!/bin/bash
+
+## /work/01839/dakotaz/local/bin/cghub/bin/gtdownload is the executable
 ## -v is the verbose option so that I know what the program is doing
 ## -c is the credential to download protected data, good until May 1, 2016
 ## -p is the path to where we want to store the file being downloaded
@@ -8,4 +9,4 @@
 manifest=$1
 path=$2
 
-time /work/01839/dakotaz/cghub/bin/gtdownload -v -c /work/01839/dakotaz/cghub.key -p $path -d $manifest
+time /work/01839/dakotaz/local/bin/cghub/bin/gtdownload -v -c /work/01839/dakotaz/cghub.key -p $path -d $manifest

### 2015-06-24 PAPER/paper.tex
moved the keynote doucument with figure 1 to FIGURE_SCRIPTS

@@
-\documentclass[11pt]{article} % use larger type; default would be 10pt
+\documentclass[11 pt]{article} % use larger type; default would be 10pt
 \usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
@@
 \title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
-\author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
+\author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
 \author[2,3]{Matthew C. Cowperthwaite}
 \author[1,4]{Claus O. Wilke} 
-\affil[1]{The University of Texas at Austin, Cell \& Molecular Biology}
-\affil[2]{St. David's NeuroTexas Institute Research Foundation}
-\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin}
-\affil[4]{The University of Texas at Austin, Integrative Biology}
+\affil[1]{Cell \& Molecular Biology, The University of Texas at Austin, Austin, TX USA}
+\affil[2]{St. David's NeuroTexas Institute Research Foundation, Austin, TX, USA}
+\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin, Austin TX, USA}
+\affil[4]{Integrative Biology, The University of Texas at Austin, Austin, TX, USA}
 
@@
 We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
+
+\clearpage
 

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
consolodating figure scripts

@@ -1,0 +1,17 @@
+## This is the R script (run on a Macbook) to create the figures from the data in FIGURE_DATA
+
+###############
+## LOAD DATA ##
+###############
+
+data23 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
+
+#####################################################
+## FIGURE 2: Number of putative SNVs in WGS v. WGA ##
+#####################################################
+
+# pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
+par(bty='n')
+plot(data23$C282_count, data23$C484_count, log='xy', pch=20, xlab="No. of putative SNVs, WGA", ylab="No. of putative SNVs, WGS" )
+abline(0,1)
+# dev.off()

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/histograms_of_mut_frequency.r

@@ -1,5 +1,5 @@
 ## This is for figure 3 in the paper
-## and hte plot with the trendline (3)
+## and hte plot with the trendline (2)
 
 # setwd( "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/GBM_genomics/FIGURES/" )
 

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed and re-named Figure2

@@ -10,8 +10,26 @@ data23 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
 ## FIGURE 2: Number of putative SNVs in WGS v. WGA ##
 #####################################################
 
-# pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
+pdf( "../FIGURE_PDFS/Figure2.pdf" )
 par(bty='n')
-plot(data23$C282_count, data23$C484_count, log='xy', pch=20, xlab="No. of putative SNVs, WGA", ylab="No. of putative SNVs, WGS" )
+plot(data23$C282_count, data23$C484_count, log='xy', pch=20, xlab="Number of putative SNVs, WGA", ylab="Number of putative SNVs, WGS" )
 abline(0,1)
-# dev.off()
+dev.off()
+
+############################
+## FIGURE 2 CORERELATIONS ##
+############################
+
+# First the Pearson correlation on the raw data
+cor.test(data23$C282_count, data23$C484_count)
+# OUTPUT: t = 4.2622, df = 53, p-value = 8.35e-05, 95 percent confidence interval: 0.2770757 0.6794588, cor=0.5052344
+
+# Next the Pearson correlation on the log-transformed data
+cor.test( log(data23$C282_count), log(data23$C484_count) )
+# OUTPUT: t = 3.6456, df = 53, p-value = 0.0006086, 95 percent confidence interval: 0.2070558 0.6373449, cor=0.4477571
+
+# Finally the Spearman correlation
+cor.test(data23$C282_count, data23$C484_count, method="spearman")
+# OUTPUT: S = 16142.79, p-value = 0.001511, rho=0.4176482
+
+## FIGURE 3: ##

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed Figure 3

@@ -4,32 +4,68 @@
 ## LOAD DATA ##
 ###############
 
-data23 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
+data2 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
+data3 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
 
-#####################################################
-## FIGURE 2: Number of putative SNVs in WGS v. WGA ##
-#####################################################
+#########################################################
+## FIGURE 2: Number of putative SNVs in WGS versus WGA ##
+#########################################################
 
-pdf( "../FIGURE_PDFS/Figure2.pdf" )
-par(bty='n')
-plot(data23$C282_count, data23$C484_count, log='xy', pch=20, xlab="Number of putative SNVs, WGA", ylab="Number of putative SNVs, WGS" )
-abline(0,1)
-dev.off()
+# pdf( "../FIGURE_PDFS/Figure2.pdf" )
+# par(bty='n')
+# plot(data2$C282_count, data2$C484_count, log='xy', pch=20, xlab="Number of putative SNVs, WGA", ylab="Number of putative SNVs, WGS" )
+# abline(0,1)
+# dev.off()
 
 ############################
 ## FIGURE 2 CORERELATIONS ##
 ############################
 
 # First the Pearson correlation on the raw data
-cor.test(data23$C282_count, data23$C484_count)
+# cor.test(data2$C282_count, data2$C484_count)
 # OUTPUT: t = 4.2622, df = 53, p-value = 8.35e-05, 95 percent confidence interval: 0.2770757 0.6794588, cor=0.5052344
 
 # Next the Pearson correlation on the log-transformed data
-cor.test( log(data23$C282_count), log(data23$C484_count) )
+# cor.test( log(data2$C282_count), log(data2$C484_count) )
 # OUTPUT: t = 3.6456, df = 53, p-value = 0.0006086, 95 percent confidence interval: 0.2070558 0.6373449, cor=0.4477571
 
 # Finally the Spearman correlation
-cor.test(data23$C282_count, data23$C484_count, method="spearman")
+# cor.test(data2$C282_count, data2$C484_count, method="spearman")
 # OUTPUT: S = 16142.79, p-value = 0.001511, rho=0.4176482
 
-## FIGURE 3: ##
+#########################################################################
+## FIGURE 3: Density of percentage overlap between WGS and WGA samples ##
[…]

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed Figure4

@@ -5,16 +5,16 @@
 ###############
 
 data2 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
-data3 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
+data34 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
 
 #########################################################
 ## FIGURE 2: Number of putative SNVs in WGS versus WGA ##
 #########################################################
 
 # pdf( "../FIGURE_PDFS/Figure2.pdf" )
-# par(bty='n')
-# plot(data2$C282_count, data2$C484_count, log='xy', pch=20, xlab="Number of putative SNVs, WGA", ylab="Number of putative SNVs, WGS" )
-# abline(0,1)
+par(bty='n')
+plot(data2$C282_count, data2$C484_count, log='xy', pch=20, xlab="Number of putative SNVs, WGA", ylab="Number of putative SNVs, WGS" )
+abline(0,1)
 # dev.off()
 
 ############################
@@ -37,35 +37,55 @@ data3 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
 ## FIGURE 3: Density of percentage overlap between WGS and WGA samples ##
 #########################################################################
 
-pdf( "../FIGURE_PDFS/Figure3.pdf" )
+# pdf( "../FIGURE_PDFS/Figure3.pdf" )
 # figure title (should it ever be needed): Greater % of WGS sample found in overlap than WGA
-amp_frac = density( data3$OVERLAP/data3$TOTAL_WGA.C282., from=0 )
-seq_frac = density( data3$OVERLAP/data3$TOTAL_WGS.C484., from=0 )
+amp_frac = density( data34$OVERLAP/data34$TOTAL_WGA.C282., from=0 )
+seq_frac = density( data34$OVERLAP/data34$TOTAL_WGS.C484., from=0 )
 par(bty='n')
 plot( seq_frac$x, seq_frac$y, type='l', col='blue', main="", xlab="Percentage of putative SNVs in a sample also recovered in the samples's technical replicate", ylab="Density")
 legend( "topright", c( "WGS", "WGA" ), col=c( "blue", "red" ), lty=1 )
 lines( amp_frac$x, amp_frac$y, col='red' )
-dev.off()
+# dev.off()
 
[…]

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed Figure 5

@@ -6,6 +6,7 @@
 
 data2 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
 data34 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
+data56 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
 
 #########################################################
 ## FIGURE 2: Number of putative SNVs in WGS versus WGA ##
@@ -51,12 +52,12 @@ lines( amp_frac$x, amp_frac$y, col='red' )
 ## FIGURE 4: Percentage of overlap versus (log) number of putative SNVs per sample ##
 #####################################################################################
 
-pdf( "../FIGURE_PDFS/Figure4.pdf" )
+# pdf( "../FIGURE_PDFS/Figure4.pdf" )
 # when necessary, use the following title: Number of putative SNPs does not correlate with percentage WGS/WGA overlap
 par(bty = 'n')
 plot( data34$TOTAL, data34$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="Number of putative SNVs in unfiltered data (log)", ylab="Percentage overlap between WGS and WGA in unfiltered data" )
 abline( 0,1 )
-dev.off()
+# dev.off()
 
 ############################
 ## FIGURE 4 CORERELATIONS ##
@@ -74,7 +75,17 @@ dev.off()
 # cor.test(data34$TOTAL, data34$PERCENT_OVERLAP_by_WGA.282, method="spearman")
 # OUTPUT: S = 29268, p-value = 0.6847, rho=-0.05584416
 
-## FIGURE 5: ##
+######################################################################
+## FIGURE 5: Number of putative SNPs removed by each of six filters ##
+######################################################################
+
+# make density of each filter removal
+
+# pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
+par(bty='n', mar=c(5.1,4.1,1.1,0))
+boxplot( data56[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number of putative SNVs filtered, per sample" )
+axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
+# dev.off()
 
 ## FIGURE 6: ##
 

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed Figure 6

@@ -6,7 +6,8 @@
 
 data2 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
 data34 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
-data56 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+data5 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+data6 = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
 
 #########################################################
 ## FIGURE 2: Number of putative SNVs in WGS versus WGA ##
@@ -83,11 +84,17 @@ abline( 0,1 )
 
 # pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
 par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( data56[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number of putative SNVs filtered, per sample" )
+boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number of putative SNVs filtered, per sample" )
 axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 # dev.off()
 
 ## FIGURE 6: ##
+
+pdf( "../FIGURE_PDFS/Figure6.pdf", width=6, height=5 )
+par(bty='n', mar=c(5.1,4.1,1.1,0))
+boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="Percentage of overlap filtered, per sample", pch=20 )
+axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
+dev.off()
 
 ## FIGURE 7: ##
 

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed Figure 7

@@ -8,6 +8,12 @@ data2 = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
 data34 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
 data5 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
 data6 = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+data78a = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
+data78a[ 3:14 ] <- list( NULL )
+data78b = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
+data78b[ 2:7 ] <- list( NULL )
+data78b[ 4:5 ] <- list( NULL )
+data78b[ 6:7 ] <- list( NULL )
 
 #########################################################
 ## FIGURE 2: Number of putative SNVs in WGS versus WGA ##
@@ -88,15 +94,33 @@ boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, yl
 axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 # dev.off()
 
-## FIGURE 6: ##
+#################################################################
+## FIGURE 6: Percentage of WGS/WGSA overlap removed, by filter ##
+#################################################################
 
-pdf( "../FIGURE_PDFS/Figure6.pdf", width=6, height=5 )
+# pdf( "../FIGURE_PDFS/Figure6.pdf", width=6, height=5 )
 par(bty='n', mar=c(5.1,4.1,1.1,0))
 boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="Percentage of overlap filtered, per sample", pch=20 )
 axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-dev.off()
+# dev.off()
 
-## FIGURE 7: ##
[…]

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure 8

@@ -117,15 +117,17 @@ names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT
 # note that this is pretty close to 1.00 every time, hence the inversion...
 LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
 
-pdf( "../FIGURE_PDFS/Figure7.pdf" )
+# pdf( "../FIGURE_PDFS/Figure7.pdf" )
 par(bty='n')
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by LOH", pch=20 )
-dev.off()
+# dev.off()
 
 ## FIGURE 8: ##
 
-
-
+pdf( "../FIGURE_PDFS/Figure8.pdf" )
+par(bty='n')
+plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by VAQ", pch=20 )
+dev.off()
 
 
 

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/LOH_VAQ_analysis.r
cleaned up the figure scripts folder

@@ -1,36 +1,0 @@
-# we are redoing the LOH/VAQ analysis, because the figures are screwed...
-
-#######################################
-# STEP 1: make the correct data frame #
-#######################################
- 
-overlap_data = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
-overlap_data[ 3:14 ] <- list( NULL )
-
-filter_data = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
-filter_data[ 2:7 ] <- list( NULL )
-filter_data[ 4:5 ] <- list( NULL )
-filter_data[ 6:7 ] <- list( NULL )
-
-LOH_VAQ_data <- merge( overlap_data, filter_data, by="SAMPLE" )
-names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_5_COUNT" ] <- "VAQ_COUNT"
-names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_COUNT" ] <- "LOH_COUNT"
-names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_5_PERCENT" ] <- "VAQ_PERCENT"
-names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT"
-
-# note that this is pretty close to 1.00 every time, hence the inversion...
-LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
-
-#############################################################
-# STEP 2: make figure 8, x-axis=OVERLAP, y-axis=LOH_PERCENT #
-#############################################################
-
-par(bty='n')
-plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Length of overlap", ylab="% of overlap filtered out by LOH", pch=20 )
-
-#############################################################
-# STEP 3: make figure 9, x-axis=OVERLAP, y-axis=VAQ_PERCENT #
-#############################################################
-
-par(bty='n')
-plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Length of overlap", ylab="% of overlap filtered out by VAQ", pch=20 )

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/histograms_of_filter_removals.r

@@ -1,56 +1,0 @@
-## FIGURES IN PAPER: 6 and 7
-## it's the two box plots
-
-# this script looks in different ways at which mutations are removed when I run the filters
-
-# first look at the STD_40 data
-# first, just everything removed, each file individually, how many by each filter 3-8
-data = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
-
-# make density of each filter removal
-d3 = density( data$FILTER_3 )
-d4 = density( data$FILTER_4 )
-d5 = density( data$FILTER_5 )
-d6 = density( data$FILTER_6 )
-d7 = density( data$FILTER_7 )
-d8 = density( data$FILTER_8 )
-
-# make the box plot that will show everything
-png( "../FIGURE_PDFS/boxplot_number_filtered.png", width = 480, height = 360, res=55 )
-par(bty='n')
-boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNPs filtered, per sample" )
-axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "Removes Variant\nAllele Quality\n(VAQ)<20", "Removes\nLOH", "Removes putative\nSNPs within\n10 bp window\nof other SNPs", "Removes putative\nSNPs within\n10 bp window\nof indels", "Removes\noverlap\nwith dbSNP\ncoverage", "Removes when\nalt. allele\ncoverage<=10%" ) )
-dev.off()
-
-pdf( "../FIGURE_PDFS/boxplot_number_filtered.pdf", width=6, height=5 )
-par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( data[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="# of putative SNVs filtered, per sample" )
-axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-dev.off()
-
-### still working on two problems:
-### how to make the figure horizontally longer so that all labels fit, and 
-### how to move the x-axis lables down relative to the x-axis, again so they fit
-
-
-#axis( 1, at=1:6 )
[…]

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/histograms_of_mut_frequency.r

@@ -1,48 +1,0 @@
-## This is for figure 3 in the paper
-## and hte plot with the trendline (2)
-
-# setwd( "/share/WilkeLab/work/dzd58/TCGA_Reanalysis/GBM_genomics/FIGURES/" )
-
-# this script makes histograms of mutation frequency for the different data sets
-
-# first load the data
-data = read.table( "../FIGURE_DATA/unfiltered_WGS-WGA_counts.txt", header=T )
-
-# density plot of C282
-pdf( "../FIGURE_PDFS/unfiltered_C282_density.pdf" )
-d1 = density( data$C282_count, from=0, to=10000 )
-plot(d1)
-dev.off()
-
-# density plot of C484
-pdf( "../FIGURE_PDFS/unfiltered_C484_density.pdf" )
-d2 = density( data$C484_count, from=0, to=10000 )
-plot(d2)
-dev.off()
-
-# plot of (paired) C282 against C484, with a line at 1 to see how things fall
-# anything below the line is in line with the hypothesis, above is bad
-#png( "../FIGURE_PDFS/C282_v_C484.png" )
-pdf( "../FIGURE_PDFS/C282_v_C484.pdf" )
-par(bty='n')
-plot(data$C282_count, data$C484_count, log='xy', pch=20, xlab="No. of putative SNVs, WGA", ylab="No. of putative SNVs, WGS" )
-abline(0,1)
-dev.off()
-
-# the two density plots on top of each other, to see what is going on
-pdf( "../FIGURE_PDFS/unfiltered_C282_C484_density.pdf" )
-par(bty='n')
-plot( d2$x, d2$y, type='l', col='red', xlab="Density", ylab="% of sample contained in overlap")
-lines( d1$x, d1$y, col='blue' )
-dev.off()
-
-# and finally, the differene: how much bigger is the one than the other?
-pdf( "../FIGURE_PDFS/unfiltered_C282-C484.pdf" ) 
-d3 = density( data$C282_count-data$C484_count )
-plot( d3$x, d3$y, type='l', col='blue')
-dev.off()
-
-
-
-
-

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/histograms_of_unfiltered_overlap.R

@@ -1,44 +1,0 @@
-## This is for figure 4 and 5 in the paper
-## that is the density plot (4)
-## and the log figure that is random
-
-# load the data
-data = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
-
-# density plot of the percent overlap for the 55 WGA-S samples
-png( "../FIGURE_PDFS/unfiltered_overlap_density.png" )
-# pdf( "../FIGURE_PDFS/unfiltered_overlap_density.pdf" )
-par(bty = 'n')
-overlap = density( data$PERCENT_OVERLAP_by_WGA.282, from=0, to=1 )
-plot( overlap, main="Density of [WGS/WGA overlap]/[Total WGA mutations]", xlab="[WGS/WGA overlap]/[Total WGA mutations]" )
-dev.off()
-
-# density plot of the percent difference for the 55 WGA-S samples
-pdf( "../FIGURE_PDFS/unfiltered_difference_density.pdf" )
-par(bty='n')
-difference = density( data$PERCENT_DIFFERENCE_by_WGA.282, from=0, to=1 )
-plot( difference, main="Density of [WGS/WGA difference]/[Total WGA mutations]", xlab="[WGS/WGA difference]/[Total WGA mutations]" )
-dev.off()
-
-# plot of the size of the overlap v. total # unfiltered mutations
-#png( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.png" )
-pdf( "../FIGURE_PDFS/unfiltered_total_muts_v_percent_overlap.pdf" )
-# when necessary, use the following title: No. of putative SNPs does not correlate with % WGS/WGA overlap
-par(bty = 'n')
-plot( data$TOTAL, data$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="log of no. of putative SNVs (unfiltered)", ylab="percent overlap between WGS and WGA (unfiltered)" )
-abline( 0,1 )
-dev.off()
-
-# for each of WGA(282) and WGS(484), is there a clear difference in percentage of mutations that are in the overlap?
-# in other words, what is teh density of (a_only/a_all) - (b_only/b_all)?
-#png( "../FIGURE_PDFS/unfiltered_overlap_WGS_WGA_together_densities.png" )
-pdf( "../FIGURE_PDFS/unfiltered_overlap_WGS_WGA_together_densities.pdf" )
[…]

### 2015-06-24 FIGURES/FIGURE_SCRIPTS/more_filter_figures.r

@@ -1,57 +1,0 @@
-# the purpose of this script is to make some figures...
-
-## 1. For LOH, VAQ, plot size of overlap (x-axis) against percent of overlap removed by relevant filter (y-axis)
-
-overlap_data = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T )
-filter_data = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T )
-
-all_data <- merge( overlap_data, filter_data, by="SAMPLE" )
-
-# LOH (filter 7)
-LOH_vars = c( "OVERLAP", "FILT_7_COUNT" )
-LOH_data = all_data[ LOH_vars ]
-FILT_7_PERCENT_new = LOH_data$FILT_7_COUNT / LOH_data$OVERLAP
-LOH_data$FILT_7_PERCENT_new = FILT_7_PERCENT_new
-
-pdf( "/Users/dakota/github/GBM_genomics/FIGURES/FIGURE_PDFS/LOH_VAQ/LOH_all.pdf" )
-par(bty='n')
-plot( LOH_data$OVERLAP, LOH_data$FILT_7_PERCENT_new, xlab="Length of overlap", ylab="% of overlap filtered out by LOH", pch=20 )
-dev.off()
-
-#png( "/Users/dakota/Desktop/FIG_RD/LOH_partial.png" )
-#par(bty='n')
-#plot( LOH_data$OVERLAP, LOH_data$FILT_7_PERCENT_new, xlim=c( 0, 500 ), xlab="Length of overlap", ylab="% of overlap filtered out by LOH", pch=20 )
-#dev.off()
-
-# VAQ (filter 5)
-VAQ_vars = c( "OVERLAP", "FILT_5_PERCENT" )
-VAQ_data = all_data[ VAQ_vars ]
-
-pdf( "/Users/dakota/github/GBM_genomics/FIGURES/FIGURE_PDFS/LOH_VAQ/VAQ_all.pdf" )
-par(bty='n')
-plot( VAQ_data$OVERLAP, VAQ_data$FILT_5_PERCENT, xlab="Length of overlap", ylab="% of overlap filtered out by VAQ", pch=20 )
-dev.off()
-
-#png( "/Users/dakota/Desktop/FIG_RD/VAQ_partial.png" )
-#par(bty='n')
[…]

### 2015-09-18 PAPER/paper.tex
adding diffset analysis for reviewers

@@
 The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
+To analyze this data, researchers typically sequence the blood and tumor of a single patient independently and then align them to each other to find somatic variants, any of which could be cancer-causing or -sustaining.
 TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
@@
 We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency between samples.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the number of SNVs found in a tumor) between samples.
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
@@
 Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation may appear fixed or completely absent.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation (a site with more than one alternate allele, in addition to the reference) may appear fixed or completely absent.
 In general, we would like to know how often these errors occur.
@@
 First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are indistinguishable from SNVs in computational analyses and will only be removed from the data if they are recorded in dbSNP; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
 Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
@@
 The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters that are commonly applied to validate SNVs.
+PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-concordance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
+Nevertheless, when less fine-tuned methods are appropriate, or are the only thing possible, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
 We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
@@
 \centerline{
-\includegraphics[width=5in]{pipeline.pdf} }
+\includegraphics[width=5in]{Figure1.pdf} }
 \caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
@@
 \centerline{
-\includegraphics[width=5in]{C282_v_C484.pdf} }
+\includegraphics[width=5in]{Figure2.pdf} }
 \caption{\textbf{Number of putative SNVs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
@@
 \centerline{
-\includegraphics[width=5in]{unfiltered_overlap_WGS_WGA_together_densities.pdf} }
+\includegraphics[width=5in]{Figure3.pdf} }
 \caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
@@
 \centerline{
-\includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
+\includegraphics[width=5in]{Figure4.pdf} }
 \caption{\textbf{Percent of overlap versus (log) number of putative SNVs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
@@
 \centerline{
-\includegraphics[width=6in]{boxplot_number_filtered.pdf} }
+\includegraphics[width=6in]{Figure5.pdf} }
 \caption{\textbf{Number of putative SNVs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
@@
 \centerline{
-\includegraphics[width=6in]{boxplot_percent_overlap_filtered.pdf} }
+\includegraphics[width=6in]{Figure6.pdf} }
 \caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
@@
 \centerline{
-\includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} }
+\includegraphics[width=5in]{Figure7.pdf} }
 \caption{\textbf{Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
@@
 \centerline{
-\includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} }
+\includegraphics[width=5in]{Figure8.pdf} }
 \caption{ \textbf{Percentage of replicate SNVs filtered out by VAQ, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.

### 2015-09-18 ANALYSIS/diffset.py

@@ -1,0 +1,1 @@
+## this is a new script

### 2015-09-18 ANALYSIS/diffset.py
further analysis of diffset

@@ -1,1 +1,81 @@
 ## this is a new script
+import os
+
+## in GBM_Genomics live ANALYSIS (home of this script) and FIGURES
+filtData="../FIGURES/FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt"
+overData="../FIGURES/FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt"
+
+## set up a dictionary, smapDict, with sample: { filter: { total, overlap, diffset } }
+## also set up a dictionary, filtDict, with filter: { total:[], overlap:[], diffset:[] }
+sampDict={}
+
+filtDict={}
+filtDict[ "F2" ]={}
+filtDict[ "F2" ][ "total" ]=[]
+filtDict[ "F2" ][ "overlap" ]=[]
+filtDict[ "F2" ][ "diffset" ]=[]
+filtDict[ "F3" ]={}
+filtDict[ "F3" ][ "total" ]=[]
+filtDict[ "F3" ][ "overlap" ]=[]
+filtDict[ "F3" ][ "diffset" ]=[]
+filtDict[ "F4" ]={}
+filtDict[ "F4" ][ "total" ]=[]
+filtDict[ "F4" ][ "overlap" ]=[]
+filtDict[ "F4" ][ "diffset" ]=[]
+filtDict[ "F5" ]={}
+filtDict[ "F5" ][ "total" ]=[]
+filtDict[ "F5" ][ "overlap" ]=[]
+filtDict[ "F5" ][ "diffset" ]=[]
+filtDict[ "F6" ]={}
+filtDict[ "F6" ][ "total" ]=[]
+filtDict[ "F6" ][ "overlap" ]=[]
+filtDict[ "F6" ][ "diffset" ]=[]
+filtDict[ "F7" ]={}
+filtDict[ "F7" ][ "total" ]=[]
+filtDict[ "F7" ][ "overlap" ]=[]
+filtDict[ "F7" ][ "diffset" ]=[]
+filtDict[ "F8" ]={}
+filtDict[ "F8" ][ "total" ]=[]
+filtDict[ "F8" ][ "overlap" ]=[]
+filtDict[ "F8" ][ "diffset" ]=[]
+
+## FILTER TOTALS
+filtFh=open( filtData, 'r' )
[…]

### 2015-09-30 PAPER/paper.tex
don't know...

--- 2015-06-24 PAPER/paper.tex
@@
 \centerline{
-\includegraphics[width=5in]{C282_v_C484.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+\includegraphics[width=5in]{Figure2.pdf} }
+\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
@@
 \centerline{
-\includegraphics[width=5in]{unfiltered_overlap_WGS_WGA_together_densities.pdf} }
-\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
+\includegraphics[width=5in]{Figure3.pdf} }
+\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
@@
 \centerline{
-\includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNVs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\includegraphics[width=5in]{Figure4.pdf} }
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
 \label{fig:unfiltered_total_muts}

### 2015-09-30 PAPER/paper.tex

--- 2015-09-18 PAPER/paper.tex
@@
 \includegraphics[width=5in]{Figure2.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
@@
 \includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
+\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
@@
 \includegraphics[width=5in]{Figure4.pdf} }
-\caption{\textbf{Percent of overlap versus (log) number of putative SNVs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
 \label{fig:unfiltered_total_muts}

### 2015-09-30 PAPER/paper.tex
1-2/13 revision

@@
 We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the number of SNVs found in a tumor) between samples.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
@@
 Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation (a site with more than one alternate allele, in addition to the reference) may appear fixed or completely absent.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation, defined as a mutation that is not fixed and therefor is not present in all tumor cells, may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
+Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
 In general, we would like to know how often these errors occur.

### 2015-09-30 PAPER/paper.tex
3-6/13 rebuttal/revisions

@@
 The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
-To analyze this data, researchers typically sequence the blood and tumor of a single patient independently and then align them to each other to find somatic variants, any of which could be cancer-causing or -sustaining.
+To identify somatic mutations in cancer genomics data, researchers typically independently sequence the blood and tumor of a single patient and align the blood and tumor sequences to each other to identify somatic variants that are specific to the tumor DNA, any of which could be cancer-causing or -sustaining.
 TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
@@
 First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are indistinguishable from SNVs in computational analyses and will only be removed from the data if they are recorded in dbSNP; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
-Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
+This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood.
 Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.

### 2015-09-30 FIGURES/FIGURE_SCRIPTS/Figures.R
changed all the figure files again...

@@ -41,33 +41,36 @@ abline(0,1)
 # cor.test(data2$C282_count, data2$C484_count, method="spearman")
 # OUTPUT: S = 16142.79, p-value = 0.001511, rho=0.4176482
 
-#########################################################################
-## FIGURE 3: Density of percentage overlap between WGS and WGA samples ##
-#########################################################################
+##################################################################################
+## FIGURE 3 ORIGINAL: Density of percentage overlap between WGS and WGA samples ##
+##################################################################################
 
 # pdf( "../FIGURE_PDFS/Figure3.pdf" )
 # figure title (should it ever be needed): Greater % of WGS sample found in overlap than WGA
-amp_frac = density( data34$OVERLAP/data34$TOTAL_WGA.C282., from=0 )
-seq_frac = density( data34$OVERLAP/data34$TOTAL_WGS.C484., from=0 )
-par(bty='n')
-plot( seq_frac$x, seq_frac$y, type='l', col='blue', main="", xlab="Percentage of putative SNVs in a sample also recovered in the samples's technical replicate", ylab="Density")
-legend( "topright", c( "WGS", "WGA" ), col=c( "blue", "red" ), lty=1 )
-lines( amp_frac$x, amp_frac$y, col='red' )
+# amp_frac = density( data34$OVERLAP/data34$TOTAL_WGA.C282., from=0 )
+# seq_frac = density( data34$OVERLAP/data34$TOTAL_WGS.C484., from=0 )
+# par(bty='n')
+# plot( seq_frac$x, seq_frac$y, type='l', col='blue', main="", xlab="Percentage of putative SNVs in a sample also recovered in the samples's technical replicate", ylab="Density")
+# legend( "topright", c( "WGS", "WGA" ), col=c( "blue", "red" ), lty=1 )
+# lines( amp_frac$x, amp_frac$y, col='red' )
 # dev.off()
 
-#####################################################################################
-## FIGURE 4: Percentage of overlap versus (log) number of putative SNVs per sample ##
-#####################################################################################
+#############################################################################################################
+## FIGURE 3 MOD (ORIGINAL FIGURE 4): Percentage of overlap versus (log) number of putative SNVs per sample ##
[…]

### 2015-09-30 PAPER/paper.tex

@@
 There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Pearson $r=0.51$, $df=53$, $P=8.35\times10^{-5}$, Figure~\ref{fig:C282_v_C484}).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:Figure2}).
 As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_overlap}).
-We further found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
+We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
+As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
 
@@
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Pearson $r=-0.04$, $df=53$, $P=0.76$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
+However, we found no significant correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
 Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
@@
 The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-concordance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
+PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-condance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
 Nevertheless, when less fine-tuned methods are appropriate, or are the only thing possible, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
@@
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Pearson $r=0.51$, $df = 53$, $P = 8.35\times 10^{-5}$).}
+The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
 \label{fig:C282_v_C484}
@@
 \includegraphics[width=5in]{Figure4.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).}
 \label{fig:unfiltered_total_muts}

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Jacquard.R
added Jacquard analysis

@@ -1,0 +1,133 @@
+## first get data
+
+## this table has the size of the overlap and the size of the total, everything for the "before" Jacquard, for each sample
+data1 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T, stringsAsFactors=FALSE )
+## this table has the number of mutations excluded from the total for each sample for each filter
+data2 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
+## this table has the number of mutations for each sample/filter that were excluded FOR THE OVERLAP
+data3 = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
+
+## from data1, we want the following columns: SAMPLE, OVERLAP, TOTAL
+## (Jacquard^before( WGS, WGA )=OVERLAP/TOTAL
+data1$DIFFERENCE<-NULL
+data1$ONLY_WGA.C282.<-NULL
+data1$TOTAL_WGA.C282.<-NULL
+data1$ONLY_WGS.C484.<-NULL
+data1$TOTAL_WGS.C484.<-NULL
+data1$PERCENT_OVERLAP_by_ALL<-NULL
+data1$PERCENT_OVERLAP_by_WGA.282<-NULL
+data1$PERCENT_OVERLAP_by_WGS.484<-NULL
+data1$PERCENT_DIFFERENCE_by_ALL<-NULL
+data1$PERCENT_DIFFERENCE_by_WGA.282<-NULL
+data1$PERCENT_DIFFERENCE_by_WGS.484<-NULL
+
+data1$Jacquard.before<-data1$OVERLAP/data1$TOTAL
+
+## from data2 (excluded by filter from SAMPLE), we want SAMPLE, FILT_3, FILT_4, FILT_5, FILT_6, FILT_7 (renamed tot.)
+# change C282 and C484 to a new column, type
+data2$newtype<-"NA"
+data2$newname<-"NA"
+for (i in 1:length(data2$SAMPLE)){
+	oldname<-data2$SAMPLE[ i ]
+	newname<-substr( oldname, 6, 18 )
+	data2$newname[ i ]<-newname
[…]

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Jacquard.R
finished Jacquard analysis

@@ -123,8 +123,19 @@ mergedData$delta.Jacquard.F6<-((mergedData$Jacquard.F6-mergedData$Jacquard.befor
 mergedData$delta.Jacquard.F7<-((mergedData$Jacquard.F7-mergedData$Jacquard.before)/mergedData$Jacquard.before)*100
 mergedData$delta.Jacquard.F8<-((mergedData$Jacquard.F8-mergedData$Jacquard.before)/mergedData$Jacquard.before)*100
 
+## then make figure 6 for the paper
 
-## then average everything for the paper
+#pdf( "../FIGURE_PDFS/Jacquard1.pdf", width=6, height=5 )
+par(bty='n', mar=c(3.1,2.1,1.1,0))
+boxplot( mergedData[,c(50,52)], ylim=c(0,450), xaxt='n', ylab="Difference in Jacquard(WGS, WGA) before and after filtering", pch=20 )
+axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH" ) )
+#dev.off()
+
+#pdf( "../FIGURE_PDFS/Jacquard2.pdf", width=6, height=5 )
+par(bty='n', mar=c(5.1,4.1,1.1,0))
+boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab="Difference in Jacquard(WGS, WGA) before and after filtering", pch=20 )
+axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
+#dev.off()
 
 
 

### 2015-10-01 PAPER/paper.tex
more editing

@@
 Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
-In general, we would like to know how often these errors occur.
+In general, we would like to know how often these sorts of errors occur.
 One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
@@
 One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
+Despite the difference in sequencing protocol, we would theoretically expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
 Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
@@
 
-There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:Figure2}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_overlap}).
+There were an average of 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:C282_v_C484}).
+As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_total_muts}).
 We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
+As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
 
-% Experiment: number of mutations (figure 4)
+% Experiment: number of mutations (figure 3 **ORIGINALLY FIGURE 4** )
 %(a) a brief (one sentence to one paragraph) summary of what you did and why
 %(b) a description of what you found carrying out this experiment
-Although the percent overlap in WGS and WGA samples was fairly constant across samples, and the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations varied by orders of magnitude (Table~\ref{tab:summary_stats}).
+Although the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations in each sample varied by orders of magnitude (Table~\ref{tab:summary_stats}).
 It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
 GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample or otherwise bad data.
+One possible explanation is a degraded DNA sample, significant alignment error, or otherwise bad data.
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
@@
 
-%% Data processing commands table
-\begin{landscape}
-\begin{table}
-\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
-The rows are in order of use.}
-\label{tab:commands}
-\begin{tabular}{ p{2.5cm} p{6cm} p{12cm} }
-	software & purpose & command\\
-	\hline
-	picard & regenerate fastq files from BAM file aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
-	bwa & align fastq files to hg19 & \texttt{bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai} \\
-	bwa, samtools & convert aligned fastq files into new BAM file & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
-	samtools & sort and index new BAM file & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
-	samtools & remove duplicate reads from BAM files & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
-	GATK & indel realignment & \texttt{java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam} \\
-	GATK & base recalibration & \texttt{java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp} \\
-	samtools & index recalibrated BAM file & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
-	SomaticSniper & call somatic mutations, generate VCF & \texttt{bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf} \\
-\end{tabular}
-\end{table}
-\end{landscape}
-
 %% Figure 1
@@
 \includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
-The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
-}
-\label{fig:unfiltered_overlap}
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
+We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
+The correlation was calculated with respect to WGA.}
+\label{fig:unfiltered_total_muts}
 \end{figure}
@@
 \centerline{
-\includegraphics[width=5in]{Figure4.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).}
-\label{fig:unfiltered_total_muts}
-\end{figure}
-
-%% Figure 5
-\begin{figure}
-\centerline{
-\includegraphics[width=6in]{Figure5.pdf} }
+\includegraphics[width=6in]{Figure4.pdf} }
 \caption{\textbf{Number of putative SNVs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
@@
 
-%%Figure 6
+%%Figure 5
 \begin{figure}
@@
 \centerline{
-\includegraphics[width=6in]{Figure6.pdf} }
+\includegraphics[width=6in]{Figure5.pdf} }
 \caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
@@
 \centerline{
-\includegraphics[width=5in]{Figure7.pdf} }
+\includegraphics[width=5in]{Figure6.pdf} }
 \caption{\textbf{Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
@@
 \centerline{
-\includegraphics[width=5in]{Figure8.pdf} }
+\includegraphics[width=5in]{Figure7.pdf} }
 \caption{ \textbf{Percentage of replicate SNVs filtered out by VAQ, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Figures.R
made figure 7

@@ -111,9 +111,9 @@ axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "
 ## FIGURE 6 IS MADE IN THE SCRIPT Jacquard.R IN THIS FOLDER ##
 ##############################################################
 
-#######################################################################################################################################
+##############################################################################################################################################
 ## FIGURE 7 (ORIG FIGURES 7 AND 8): Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs ##
-#######################################################################################################################################
+##############################################################################################################################################
 
 LOH_VAQ_data <- merge( data78a, data78b, by="SAMPLE" )
 names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_5_COUNT" ] <- "VAQ_COUNT"
@@ -124,20 +124,11 @@ names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT
 # note that this is pretty close to 1.00 every time, hence the inversion...
 LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
 
-#pdf( "../FIGURE_PDFS/Figure6.pdf" )
-par(bty='n')
-plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by LOH", pch=20 )
-#dev.off()
-
-###############################
-## FIGURE 7 (ORIG FIGURE 8): ##
-###############################
-
 #pdf( "../FIGURE_PDFS/Figure7.pdf" )
-par(bty='n')
+par(bty='n', mfrow=c(1,2) )
+plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by LOH", pch=20 )
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by VAQ", pch=20 )
 #dev.off()
-
 
 
 

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Figures.R
finished this round of figure edits (probably only one to go...)

@@ -85,34 +85,24 @@ legend( 'topright', c( "overlap/WGA total", "overlap/WGS total" ), col=c( 'black
 # cor.test(data34$TOTAL, data34$PERCENT_OVERLAP_by_WGA.282, method="spearman")
 # OUTPUT: S = 29268, p-value = 0.6847, rho=-0.05584416
 
-######################################################################################
-## FIGURE 4 (ORIG FIGURE 5): Number of putative SNPs removed by each of six filters ##
-######################################################################################
-
-# make density of each filter removal
+############################################################################################
+## FIGURE 4 (ORIG FIGURE 5 and 6): Number of putative SNPs removed by each of six filters ##
+############################################################################################
 
 #pdf( "../FIGURE_PDFS/Figure4.pdf", width=6, height=5 )
-par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number of putative SNVs filtered, per sample" )
+par(bty='n', mfrow=c(2,1), mar=c(5.1,4.1,1.1,0))
+boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number putative SNVs filtered" )
 axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-#dev.off()
-
-#################################################################################
-## FIGURE 5 (ORIG FIGURE 6): Percentage of WGS/WGSA overlap removed, by filter ##
-#################################################################################
-
-#pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
-par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="Percentage of overlap filtered, per sample", pch=20 )
+boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="Percentage overlap filtered", pch=20 )
[…]

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Jacquard.R

@@ -1,6 +1,8 @@
-## first get data
+###################################
+## first do the Jaccard analysis ##
+###################################
 
-## this table has the size of the overlap and the size of the total, everything for the "before" Jacquard, for each sample
+## this table has the size of the overlap and the size of the total, everything for the "before" Jaccard, for each sample
 data1 = read.table( "../FIGURE_DATA/STD40_unfiltered_overlap.txt", header=T, stringsAsFactors=FALSE )
 ## this table has the number of mutations excluded from the total for each sample for each filter
 data2 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
@@ -8,7 +10,7 @@ data2 = read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_ST
 data3 = read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
 
 ## from data1, we want the following columns: SAMPLE, OVERLAP, TOTAL
-## (Jacquard^before( WGS, WGA )=OVERLAP/TOTAL
+## (Jaccard^before( WGS, WGA )=OVERLAP/TOTAL
 data1$DIFFERENCE<-NULL
 data1$ONLY_WGA.C282.<-NULL
 data1$TOTAL_WGA.C282.<-NULL
@@ -21,7 +23,7 @@ data1$PERCENT_DIFFERENCE_by_ALL<-NULL
 data1$PERCENT_DIFFERENCE_by_WGA.282<-NULL
 data1$PERCENT_DIFFERENCE_by_WGS.484<-NULL
 
-data1$Jacquard.before<-data1$OVERLAP/data1$TOTAL
+data1$Jaccard.before<-data1$OVERLAP/data1$TOTAL
 
 ## from data2 (excluded by filter from SAMPLE), we want SAMPLE, FILT_3, FILT_4, FILT_5, FILT_6, FILT_7 (renamed tot.)
 # change C282 and C484 to a new column, type
[…]

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/diffset.R

@@ -1,34 +1,34 @@
 ## LOAD DATA
 
-data=read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
-data2=read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
+data4=read.table( "../FIGURE_DATA/overlap_numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
+data5=read.table( "../FIGURE_DATA/numbers_excluded_by_filter_MUTATION_CALLS_STD40.txt", header=T, stringsAsFactors=FALSE )
 
 ## now re-name and delete columns
-names( data )[ 2 ] <- "F2overlap"
-names( data )[ 4 ] <- "F3overlap"
-names( data )[ 6 ] <- "F4overlap"
-names( data )[ 8 ] <- "F5overlap"
-names( data )[ 10 ] <- "F6overlap"
-names( data )[ 12 ] <- "F7overlap"
-names( data )[ 14 ] <- "F8overlap"
+names( data4 )[ 2 ] <- "F2overlap"
+names( data4 )[ 4 ] <- "F3overlap"
+names( data4 )[ 6 ] <- "F4overlap"
+names( data4 )[ 8 ] <- "F5overlap"
+names( data4 )[ 10 ] <- "F6overlap"
+names( data4 )[ 12 ] <- "F7overlap"
+names( data4 )[ 14 ] <- "F8overlap"
 
-data$FILT_2_PERCENT<-NULL
-data$FILT_3_PERCENT<-NULL
-data$FILT_4_PERCENT<-NULL
-data$FILT_5_PERCENT<-NULL
-data$FILT_6_PERCENT<-NULL
-data$FILT_7_PERCENT<-NULL
-data$FILT_8_PERCENT<-NULL
+data4$FILT_2_PERCENT<-NULL
+data4$FILT_3_PERCENT<-NULL
+data4$FILT_4_PERCENT<-NULL
+data4$FILT_5_PERCENT<-NULL
+data4$FILT_6_PERCENT<-NULL
+data4$FILT_7_PERCENT<-NULL
+data4$FILT_8_PERCENT<-NULL
 
 ## now merge the two data frames
-for (i in 1:length(data$SAMPLE)){
[…]

### 2015-10-01 PAPER/paper.tex
edited reviewer comment 13/13

@@
 We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
+(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}).
 We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_filtered} and Table~\ref{tab:summary_stats}).
 The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
@@
 Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
-To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
+To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_filtered}).
 We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
-By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
+By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_filtered}, Table~\ref{tab:summary_stats}).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
@@
 Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
+We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
 Either filter was removing between 0 and 100\% of the overlap for some samples.
@@
 
-We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH_all}) and the VAQ graph (Figure~\ref{fig:VAQ_all}) looked like exact inverses of each other.
+We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH-VAQ_all}) and the VAQ graph (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
 Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
@@
 
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table~\ref{tab:commands}.
+We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
 The custom python code to connect the pipeline is available in a public
@@
 \includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Number of putative SNVs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
-In all cases, the LOH and VAQ filters removed the majority of the putative SNVs for a given sample.}
-\label{fig:boxplot_number_filtered}
+\caption{\textbf{Different filters catch greater or smaller numbers of putative SNVs, and lower or higher percentages of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
+The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
+The <10\% filter removed very few putative SNVs and almost none of the overlap.
+The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
+}
+\label{fig:boxplot_filtered}
 \end{figure}
@@
 \begin{figure}
-% 379 samples, 311 patients
 \centerline{
-\includegraphics[width=6in]{Figure5.pdf} }
-\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
-Filters removing LOH and putative SNVs with low VAQ removed a significant percentage of the overlap; nothing else did.}
-\label{fig:boxplot_percent_overlap_filtered}
+\includegraphics[width=5in]{Figure5.pdf} }
+\caption{\textbf{A title.} a caption.}
+\label{fig:diffset}
 \end{figure}
 
-%%Figure 7
+%%Figure 6
 \begin{figure}
@@
 \includegraphics[width=5in]{Figure6.pdf} }
-\caption{\textbf{Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
-This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
-\label{fig:LOH_all}
-\end{figure}
-
-%%Figure 8
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure7.pdf} }
-\caption{ \textbf{Percentage of replicate SNVs filtered out by VAQ, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
-This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
-\label{fig:VAQ_all}
+\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
+The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
+\label{fig:LOH-VAQ_all}
 \end{figure}

### 2015-10-01 PAPER/paper.tex
more editing

@@
 
-% Experiment: which filters do what (figures 6 and 7) 
-% Experiment: those two filters are different (figures 8 and 9)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
+% Experiment: which filters do what (figures 4 and 5) 
 As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
@@
 The first three are based on the quality scores generated by GATK and SomaticSniper.
-We simply remove from the dataset anything that fails to meet conventional quality-score thresholds.
+We simply remove from the dataset anything that fails to meet these conventional quality-score thresholds.
 The other five are additional filters that we developed for this project.
@@
 We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}).
+(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We ran each filter independently on the whole dataset to see which putative SNVs it caught.
+We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}) on each sample.
 We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
@@
 
-We next asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
+%% start edits for Reviewer One comment 13 here%%
+We next looked at the individual effects the six of eight filters on the overlap between replicates.
+We asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
 Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
@@
 
-%% this is figures 8 and 9, which need to be added to the document and explained now...
+Finally, we looked at the individual effects of the six of eight filters on the difference between replicates of each sample (only those mutations found in WGS \textit{or} WGA, \textit{but not} both), and on the similarity of WGS to WGA as a whole.
+We first looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA, where a higher ratio indicates that more mutations are removed from the difference than the overlap.
+We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}).
+To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}).
+We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
+(We did no look at 1the <10\% filter for this analysis, because it did not catch enough data to be meaningful.)
+
+We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient (union/intersection, where the intersection is all those putative SNVs found in WGS \textit{or} WGA \textit{or} both) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
+By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}).
+This is most likely because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
+Removing so much of the data makes the intersection much, much smaller, therefor making the similarity between samples much larger, mostly independently of the size of the overlap.
+
+%% this is figures 6
 Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Jacquard.R
edited figure 6

@@ -242,16 +242,10 @@ median(all$F7ratio, na.rm=TRUE)
 ## THEN MAKE FIGURE 6 (COMPOUND FIGURE) FOR THE PAPER ##
 ########################################################
 
-#pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
-par(bty='n', mfrow=c(2,2), mar=c(5.1,4.1,1.1,0) )
+pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
+par(bty='n', mfrow=c(2,2), mar=c(3,4.1,2,1) )
 
-boxplot( mergedData[,c(52,50)], ylim=c(0,450), xaxt='n', ylab=expression(paste( Delta, "Jaccard( WGS, WGA )")), pch=20 )
-axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "LOH", "VAQ" ) )
-
-boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab="", pch=20 )
-axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-
-plot( all$F7ratio, ylim=c(0,80), pch=20, col='black', ylab="Ratio of intersection to union", xaxt='n', xlab="Samples" )
+plot( all$F7ratio, ylim=c(0,80), pch=20, col='black', ylab="Ratio of difference to overlap", xaxt='n', xlab="Samples" )
 points( all$F5ratio, pch=20, col='navy' )
 points( all$F3ratio, pch=20, col='red' )
 points( all$F4ratio, pch=20, col='yellow' )
@@ -260,7 +254,16 @@ legend( 'topright', c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ), col=c(
 
 boxplot( all[,c(26,28,25,24,27)], ylim=c(1,50), xaxt='n', pch=20, ylab="" )
 axis( 1, at=1:5, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ) )
-#dev.off()
+
[…]

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Jacquard.R
very small change

@@ -242,7 +242,8 @@ median(all$F7ratio, na.rm=TRUE)
 ## THEN MAKE FIGURE 6 (COMPOUND FIGURE) FOR THE PAPER ##
 ########################################################
 
-pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
+#pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
+
 par(bty='n', mfrow=c(2,2), mar=c(3,4.1,2,1) )
 
 plot( all$F7ratio, ylim=c(0,80), pch=20, col='black', ylab="Ratio of difference to overlap", xaxt='n', xlab="Samples" )
@@ -261,7 +262,7 @@ axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "LOH", "VAQ" ) )
 boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab="", pch=20 )
 axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 
-dev.off()
+#dev.off()
 
 
 

### 2015-10-01 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure numbering mistake

@@ -114,11 +114,11 @@ names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT
 # note that this is pretty close to 1.00 every time, hence the inversion...
 LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
 
-#pdf( "../FIGURE_PDFS/Figure6.pdf" )
+pdf( "../FIGURE_PDFS/Figure6.pdf" )
 par(bty='n', mfrow=c(1,2) )
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by LOH", pch=20 )
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by VAQ", pch=20 )
-#dev.off()
+dev.off()
 
 
 ##############

### 2015-10-01 PAPER/paper.tex
finished edits 10-11/13

@@
 
-Finally, we looked at the individual effects of the six of eight filters on the difference between replicates of each sample (only those mutations found in WGS \textit{or} WGA, \textit{but not} both), and on the similarity of WGS to WGA as a whole.
-We first looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA, where a higher ratio indicates that more mutations are removed from the difference than the overlap.
+We next looked at the individual effects of the six of eight filters on the difference between replicates of each sample, those mutations found only in WGS \textit{or} WGA, \textit{but not} both.
+We looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$), where a higher ratio indicates that more mutations are removed from the difference than the overlap.
 We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}).
@@
 We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
-(We did no look at 1the <10\% filter for this analysis, because it did not catch enough data to be meaningful.)
+(We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
 
-We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient (union/intersection, where the intersection is all those putative SNVs found in WGS \textit{or} WGA \textit{or} both) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
+We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient ($WGS \cap WGA / WGS \cup WGA$) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
 By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}).
-This is most likely because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
-Removing so much of the data makes the intersection much, much smaller, therefor making the similarity between samples much larger, mostly independently of the size of the overlap.
+We see this effect because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
+Removing so much of the data makes the intersection much, much smaller, and thus makes the Jaccard coefficient between samples much larger, mostly independently of the size of the overlap.
 
-%% this is figures 6
-Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
+%% this is figure 6
+Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most to all of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
 We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
@@
 
-We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH-VAQ_all}) and the VAQ graph (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
+We found an additional trend that was initially unexpected: The LOH and VAQ graphs (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
 Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
@@
 We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
+Each of the eight filters was run independently on the whole data set.
 The eight filters were enacted as follows:
@@
 \includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Different filters catch greater or smaller numbers of putative SNVs, and lower or higher percentages of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
+\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
 The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
@@
 \includegraphics[width=5in]{Figure5.pdf} }
-\caption{\textbf{A title.} a caption.}
+\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
+\textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis.
+\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
 \label{fig:diffset}

### 2015-10-01 PAPER/Supplementary.tex
supplementary materials

--- 2015-10-01 PAPER/paper.tex
@@
 %%% Title and author are here
-\title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
+\title{Supplementary materials: Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
 \author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
@@
 
-\section*{Abstract}
+\section{Figures}
 
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in calling single-nucleotide variants (SNVs).
-We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
-We further tested whether additional filtering increased or decreased the size of the overlap.
-We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966).
-We further found that using filtering after SNV-calling removed the overlap completely.
-We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
+%% Figure 2
+\begin{figure}
+\centerline{
+\includegraphics[width=5in]{orig_Figure3.pdf} }
+\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
+The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.}
+\label{fig:density}
+\end{figure}
 
-\clearpage
+\section{Tables}
 
-\section*{Introduction}
-
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
-To identify somatic mutations in cancer genomics data, researchers typically independently sequence the blood and tumor of a single patient and align the blood and tumor sequences to each other to identify somatic variants that are specific to the tumor DNA, any of which could be cancer-causing or -sustaining.
-TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
-GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%.
-Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
-
-However, these and other efforts have produced very few results for patients in terms of improved treatments.
-It may be that we simply haven't yet analyzed the data in the most informative manner.
-Another factor may be that the data are noisy.
-It is well known that sequencing and variant-calling pipelines are not error-free.
-For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}.
-Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.
-Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls.
-One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}.
-Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
-
-We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
-For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches.
-We found significant overlap (around 50\%) between technical replicates, but also significant differences.
-As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
-This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
-
-\section*{Results}
-
-\subsection*{How similar are two replicate sequencing runs of the same tumor?}
-
-% Experiment: size of overlap (figures 3 and 4)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-
-DNA sequencing is not error-free.
-Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}.
-Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation, defined as a mutation that is not fixed and therefor is not present in all tumor cells, may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
-Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
-In general, we would like to know how often these sorts of errors occur.
-One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
-However, this method is expensive and time consuming.
-A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates.
-Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
-Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
-
-TCGA's GBM data set includes 2 technical replicates for each of 55 tumors.
-In this case, the technical replicates are not identical.
-One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would theoretically expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
-Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
-We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNVs per sample, than the corresponding WGS samples.
-
-For each patient (n=55), we called mutations in both technical replicates and in the patient's blood sample, using the same computational pipeline (see Figure~\ref{fig:pipeline} and Methods): We downloaded TCGA BAM files with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the overlap, see Figure~\ref{fig:pipeline}).
-We further calculated the percentage of mutations in each replicate that occurred in the overlap between the two.
-
-There were an average of 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:C282_v_C484}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_total_muts}).
-We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
-
-% Experiment: number of mutations (figure 3 **ORIGINALLY FIGURE 4** )
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-Although the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations in each sample varied by orders of magnitude (Table~\ref{tab:summary_stats}).
-It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
-GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample, significant alignment error, or otherwise bad data.
-If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
-Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
-
-\subsection*{Does more sophisticated SNV filtering software increase or decrease the degree of similarity between replicates?}
-
-% Experiment: which filters do what (figures 6 and 7) 
-% Experiment: those two filters are different (figures 8 and 9)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
-Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores.
-In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts.
-Table~\ref{tab:filters} lists eight distinct SNV filters we evaluated.
-The first three are based on the quality scores generated by GATK and SomaticSniper.
-We simply remove from the dataset anything that fails to meet conventional quality-score thresholds.
-The other five are additional filters that we developed for this project.
-Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNV call is a false positive.
-
-We first asked whether filtering the data increases or decreases the percentage of the sample that is overlapping between the two technical replicates.
-We expect that this analysis informs us about whether filtering out putative somatic mutations with these features affects the proportion of true somatic mutations in the remaining dataset.
-We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14 (Table~\ref{tab:summary_stats}).
-The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, i.e., the overall overlap percentage also decreased.
-We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as all the noise).
-
-We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}).
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_filtered} and Table~\ref{tab:summary_stats}).
-The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
-
-We next asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
-To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_filtered}).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
-By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_filtered}, Table~\ref{tab:summary_stats}).
-Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
-
-%% this is figures 8 and 9, which need to be added to the document and explained now...
-Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
-Either filter was removing between 0 and 100\% of the overlap for some samples.
-For samples with more overlap, however, LOH was the primary filter removing overlap.
-
-We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH-VAQ_all}) and the VAQ graph (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
-Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
-In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.
-
-\section*{Discussion}
-
-GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
-Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM.
-Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-These data are of enormous potential value to the research community, but their accuracy and reproducibility are unknown.
-Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
-
-We found that, on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
-The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.
-We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
-Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations.
-These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
-
-Filtering the raw computational data using both quality scores from GATK and SomaticSniper, as well as five additional custom filters, eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNVs.
-For example, LOH in this case is more likely to be an error than an actual mutation.
-When we do cancer genomics, however, our goal is the opposite, to highlight changes.
-Therefore, it is possible that we need entirely different filtering protocols.
-There are also more theoretical reasons to consider altering the filtering protocols for cancer genomes.
-For example, multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
-This hypothesis, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
-
-Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that we found in both technical replicates.
-In combination, they removed most or all of those mutations present in both replicates, since the two filters consistently removed nearly completely disjoint sets of putative mutations.
-Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative SNVs within a 10 bp window of putative indels or other putative SNVs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
-The third removes overlap with dbGaP.
-Our analysis suggests that these three filters do clean up the data in a meaningful way.
-By contrast, it may be more useful not to apply the two filters that removed principally data from the overlap of the two replicates.
-
-Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
-Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood.
-Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
-
-Second, we studied only one particular SNV caller, SomaticSniper \citep{SomaticSniper}.
-There are other, equally widely used SNV callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
-Also, we did not even consider every possible quality metric available in the SNV-caller we did choose.
-For example, we did not look at the effects of the SomaticScore or the GATK quality score individually.
-In the future, it would be worthwhile to evaluate other SNV-callers and other quality metrics.
-
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNVs in WGS samples and about 30\% in WGA samples.
-The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-condance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
-Nevertheless, when less fine-tuned methods are appropriate, or are the only thing possible, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
-We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest that when mutation validation by PCR is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
-
-\section*{Methods}
-
-\subsection*{Data and back-end processing}
-
-All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
-We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
-For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
-In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
-
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
-The custom python code to connect the pipeline is available in a public
-git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
-We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
-Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
-
-\subsection*{Filtering and data analysis}
-
-After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
-The eight filters were enacted as follows:
-
-\begin{itemize}
-\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
-As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
-SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
-We excluded LOH variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-SNV variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
-We excluded dbSNP variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
-We excluded $<10\%$ variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\end{itemize}
-
-We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering.
-These scripts are also available in the public git repository.
-We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
-This code is also available in the public git repository.
-
-%\section*{}
-
-\bibliography{TCGA_bib}
-\bibliographystyle{peerj}
-
-\cleardoublepage
-\section*{Tables and Figures}
-
-%% Summary Statistics Table
+%% Data processing commands table
 \begin{landscape}
 \begin{table}
-\caption{\textbf{Summary Statistics.} This table describes the results of our data processing pipeline across samples and pairs of replicates, in terms of the number of putative mutations found, length of the overlap between replicates, number of mutations removed (per sample) by each filter, and percent of the overlap removed (per pair of replicates) by each filter.}
-\label{tab:summary_stats}
-\begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
-	Quantity & average & median & min & max & stdev \\
+\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
+The rows are in order of use.}
+\label{tab:commands}
+\begin{tabular}{ p{2.5cm} p{6cm} p{12cm} }
+	software & purpose & command\\
 	\hline
-	No. mutations, WGS & 844 & 328 & 110 & 8,192 & 1,306 \\
-	No. mutations, WGA & 1,531 & 694 & 23 & 10,966 & 2,230 \\
-	\% mutations in overlap, WGS & 31\% & 31\% & 1\% & 74\% & 20\% \\
-	\% mutations in overlap, WGA & 44\% & 45\% & 3\% & 71\% & 13\% \\
-	No. putative mutations removed, VAQ & 309 & 164 & 17 & 11,439 & 1,372 \\
-	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
-	No. putative mutations removed, 10bp-SNV & 46 & 19 & 0 & 1,515 & 115 \\
-	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
-	No. putative mutations removed, dbSNP & 16 & 8 & 0 & 348 & 33 \\
-	No. putative mutations removed, $<$10\% & 1 & 1 & 1 & 36 & 3 \\
-	\% overlap removed, VAQ & 53\% & 57\% & 0\% & 100\% & 28\% \\
-	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
-	\% overlap removed, 10bp-SNV & 3\% & 2\% & 0\% & 20\% & 3\% \\
-	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
-	\% overlap removed, dbSNP & 3\% & 2\% & 0\% & 47\% & 7\% \\
+	picard & regenerate fastq files from BAM file aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
+	bwa & align fastq files to hg19 & \texttt{bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai} \\
+	bwa, samtools & convert aligned fastq files into new BAM file & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
+	samtools & sort and index new BAM file & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
+	samtools & remove duplicate reads from BAM files & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
+	GATK & indel realignment & \texttt{java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam} \\
+	GATK & base recalibration & \texttt{java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp} \\
+	samtools & index recalibrated BAM file & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
+	SomaticSniper & call somatic mutations, generate VCF & \texttt{bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf} \\
 \end{tabular}
@@
 
-%% Filters table
-\begin{landscape}
-\begin{table}
-\caption{\textbf{SNV prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
-\label{tab:filters}
-\begin{tabular}{ p{2.25cm} p{4.25cm} p{14cm} }
-	filter & software & purpose \\
-	\hline
-	GATK & GATK & Removes putative SNVs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration) \\
-	SS & SomaticSniper & Removes putative SNVs with a SomaticScore less than 40 \\
-	VAQ & SomaticSniper & Removes putative SNVs with SomaticSniper Varaint Allele Quality scores less than 20 \\
-	LOH & SomaticSniper, python & Removes putative SNVs that are identified as loss of heterozygosity \\
-	10bp-SNV & python & Removes putative SNVs located within a 10 bp window of any other putative SNV \\
-	10bp-INDEL & python & Removes putative SNVs located within a 10 bp window of indels \\
-	dbSNP & python & Removes putative SNVs that overlap with dbSNP coverage \\
-	$<$10\% & python & Removes putative SNVs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10\% \\
-\end{tabular}
-\end{table}
-\end{landscape}
-
-%% Figure 1
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure1.pdf} }
-\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
-For each BAM file, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
-We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
-\label{fig:pipeline}
-\end{figure}
-
-%% Figure 2
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure2.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
-The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
-\label{fig:C282_v_C484}
-\end{figure}
-
-%% Figure 3
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
-We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
-The correlation was calculated with respect to WGA.}
-\label{fig:unfiltered_total_muts}
-\end{figure}
-
-%% Figure 4
-\begin{figure}
-\centerline{
-\includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Different filters catch greater or smaller numbers of putative SNVs, and lower or higher percentages of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
-The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
-The <10\% filter removed very few putative SNVs and almost none of the overlap.
-The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
-}
-\label{fig:boxplot_filtered}
-\end{figure}
-
-%%Figure 5
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure5.pdf} }
-\caption{\textbf{A title.} a caption.}
-\label{fig:diffset}
-\end{figure}
-
-%%Figure 6
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure6.pdf} }
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
-\label{fig:LOH-VAQ_all}
-\end{figure}
-
 \end{document}

### 2015-10-22 ANALYSIS/seqqual.py
final analysis begun

@@ -1,0 +1,3 @@
+# runs on stampede
+# extracts information about how mapping quality
+# outputs table in form <sample ID> <patient ID> <> <> <>

### 2015-10-22 ANALYSIS/idxstats.sh
added the script used to calculate values for last analysis to git

@@ -1,0 +1,53 @@
+#!bin/bash
+
+## the purpose of this script is to gather the summary stats needed to do the last figures and the summary table in the GBM_Genomics paper revision
+
+## environment
+# module load samtools
+
+## first you need to loop over each filename
+for typeDir in /scratch/01839/dakotaz/alignments/2015-05.TCGA-GBM-53techreps/*/; do
+	
+	# get the type variable
+	type1=${typeDir#/scratch/01839/dakotaz/alignments/2015-05.TCGA-GBM-53techreps/}
+	type=${type1%/}
+	# echo $type
+	
+	# get each sample
+	for sampDir in $typeDir/*/; do
+	
+		#get the sampName variable
+		sampName1=${sampDir#$typeDir/}
+		sampName=${sampName1%/}
+		# echo "$type, $sampName"
+		
+		# now go to the bamfile in the sampDir
+		sampbam="$sampDir/$sampName.realn.recal.bam"
+		if [ -e $sampbam ]
+		then
+			# number mapped v. unmapped reads for each reference
+			samtools idxstats $sampbam > $sampDir/$sampName.realn.recal.idxstats.txt
+			
+			## and now the depth of coverage/percent of genome covered stuff
+			echo "Genome size (number of bps in the genome, denominator of coverage): " > $sampDir/$sampName.realn.recal.depth.txt
+			samtools view -H $sampbam | grep -P '^@SQ' | cut -f 3 -d ':' | awk '{sum+=$1} END {print sum}' >> $sampDir/$sampName.realn.recal.depth.txt
+			samtools view -H $sampbam | grep -P '^@SQ' | cut -f 3 -d ':' | awk '{sum+=$1} END {print sum}' >> $sampDir/$sampName.realn.recal.depth.txt
+			echo "" >> $sampDir/$sampName.realn.recal.depth.txt
+
+			echo "Number of total bases mapped (numerator of coverage, where genome size is denominator):" >> $sampDir/$sampName.realn.recal.depth.txt
+			samtools depth $sampbam |  awk '{sum+=$3} END {print sum}' >> $sampDir/$sampName.realn.recal.depth.txt
+			echo "" >> $sampDir/$sampName.realn.recal.depth.txt
+
[…]

### 2015-10-22 ANALYSIS/seqqual.py

@@ -1,3 +1,11 @@
 # runs on stampede
 # extracts information about how mapping quality
 # outputs table in form <sample ID> <patient ID> <> <> <>
+
+print "now I will analyze the sequences..."
+
+## % mapped and unmapped reads per sample
+
+## coverage (the good and the bad...)
+
+## % of genome covered...

### 2015-10-23 ANALYSIS/idxstats.sh
first pass at qual analysis

@@ -44,7 +44,7 @@ for typeDir in /scratch/01839/dakotaz/alignments/2015-05.TCGA-GBM-53techreps/*/;
 			echo "" >> $sampDir/$sampName.realn.recal.depth.txt
 
 			echo "Number of positions in the genome covered (numerator in % genome covered):" >> $sampDir/$sampName.realn.recal.depth.txt
-			samtools depth C484.TCGA-14-0790-10A-01D-1494-08.5.realn.recal.bam | wc -l >> $sampDir/$sampName.realn.recal.depth.txt
+			samtools depth C484.TCGA-14-0790-10A-01D-1494-08.5.realn.recal.bam | grep -o "chr" | wc -l >> $sampDir/$sampName.realn.recal.depth.txt
 			echo "" >> $sampDir/$sampName.realn.recal.depth.txt			
 		fi 
 	

### 2015-10-23 ANALYSIS/seqqual.py

@@ -1,11 +1,79 @@
-# runs on stampede
-# extracts information about how mapping quality
-# outputs table in form <sample ID> <patient ID> <> <> <>
+## runs on stampede
+## extracts information about how mapping quality
+## outputs table in form <sample ID> <patient ID> <%mapped> <%covered> <avg depth> <avg depth of covered>
+## outputs this data in /home1/01839/dakotaz/GBM_genomics/
 
+## environment
+import os
+
+## set up, variables, structures
 print "now I will analyze the sequences..."
+dataDict={}
 
-## % mapped and unmapped reads per sample
-
-## coverage (the good and the bad...)
-
-## % of genome covered...
+## commands
+dataDir="/scratch/01839/dakotaz/alignments/2015-05.TCGA-GBM-53techreps/"
+# for type in dataDir...
+types = os.listdir( dataDir )
+# print types
+for type in types:
+	# for sample in type (not whole genome)
+	typeDir=os.path.join( dataDir, type )
+	samples = os.listdir( typeDir )
+	# print type, samples
+	for sample in samples:
+		# print type, sample
+		sampDir=os.path.join( typeDir, sample )
+		depth=sample+".realn.recal.depth.txt"
+		idx=sample+".realn.recal.idxstats.txt"
+		depthFile=os.path.join( sampDir, depth )
+		idxFile=os.path.join( sampDir, idx )
+		if os.path.isfile( depthFile ) and os.path.isfile( idxFile ):
+			# print sample, " is a real file"
+			# put the sample names into the dictionary
+			
+			# read idxstats file to get %mapped
+			depthFh=open( depthFile, 'r' )
+			linenumber=1
+			coverage="zero"
+			for line in depthFh:
+				if linenumber==10:
+					coverage=line.strip('\n')
+					coverage=coverage.strip( 'Average: = ' )
+				linenumber=linenumber+1
+			if linenumber==16:
+			
+				dataDict[ sample ]={}
+				dataDict[ sample ][ "type" ] = type
+				patientID=sample[5:17]
+				# print patientID
+				dataDict[ sample ][ "patientID" ]= patientID
+			
+				dataDict[ sample ][ "cover" ]=coverage
+				# print sample, coverag3
+				# read idxstats file to get mapped/unmapped and calculate percent
[…]

### 2015-10-23 PAPER/Supplementary.tex
finished supplementary

@@
 
+\begin{figure}
+\centerline{
+\includegraphics[width=5in]{QualFigure1.pdf} }
+\caption{\textbf{Number of putative SNVs in a sample does not correlate with coverage.} The number of SNVs called in a sample does not correlate with the coverage of that sample  (Spearman $\rho=-0.1255027$, $S=671817.1$, $P=0.1222$).}
+\label{fig:coverage}
+\end{figure}
+
+\begin{figure}
+\centerline{
+\includegraphics[width=5in]{QualFigure2.pdf} }
+\caption{\textbf{Number of putative SNVs in a sample does not correlate with percentage of mapped reads.} The number of SNVs called in a sample does not correlate with the percentage of mapped reads in the alignment of that sample  (Spearman $\rho=-0.06771965$, $S=637326.1$, $P=0.4056$).}
+\label{fig:percent_mapped}
+\end{figure}
+
 \section{Tables}

### 2015-10-23 PAPER/paper.tex
finished revisions

@@
 We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
-The custom python code to connect the pipeline is available in a public
-git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
+The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
 Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
@@
 Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
+
+For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
+The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
+We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Supplementary Figures 2 and 3).
 

### 2015-11-01 PAPER/paper.tex
made first change to rebuttal from Matt's comments

@@
 Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation, defined as a mutation that is not fixed and therefor is not present in all tumor cells, may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation is defined as a mutation that is not fixed and therefore is not present in all tumor cells.
+Such mutations may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
 Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.

### 2015-11-12 PAPER/paper.tex
edited page 2

@@
 
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
-To identify somatic mutations in cancer genomics data, researchers typically independently sequence the blood and tumor of a single patient and align the blood and tumor sequences to each other to identify somatic variants that are specific to the tumor DNA, any of which could be cancer-causing or -sustaining.
-TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
-GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%.
+Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor with a median survival time of 14 months and a 5-year survival rate of ~5\%.
 Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
+Recently, research to improve treatments for patients with this devastating disease has focused on the idea of precision medicine, that is using large-scale genomics data to discover how the disease arises and progresses, and how to stop it.
+The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}, including over 500 GBMs.
+While these and other similar data \citep{Parsons}, researchers have discovered genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
+Despite widespread use of the data and adoption of the methods, however, efforts to benchmark the data--to assess the quality and repeatability of these and subsequent analyses--have lagged or been non-existent.
+Here, our goal is to begin to develop robust methods for benchmarking next-generation cancer sequencing data and analysis.
 
-However, these and other efforts have produced very few results for patients in terms of improved treatments.
-It may be that we simply haven't yet analyzed the data in the most informative manner.
-Another factor may be that the data are noisy.
 It is well known that sequencing and variant-calling pipelines are not error-free.

### 2015-11-13 PAPER/paper.tex
edited reviewer comments pages 2-4

@@
 The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-condance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
-Nevertheless, when less fine-tuned methods are appropriate, or are the only thing possible, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
+We acknowledge that the high rate of non-concordance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, validation by an orthogonal method (\textit{e.g.} PCR, Sanger sequencing) remains necessary.
+Nevertheless, when less fine-tuned methods are acceptable, or are the only option available, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
 We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest that when mutation validation by PCR is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
+We suggest that when orthogonal validation is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
 

### 2015-11-13 PAPER/paper.tex
finished editing reviewer comments

@@
 
-We next looked at the individual effects of six of the eight filters.
+We next looked at the individual effects of six of eight filters.
 (We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We ran each filter independently on the whole dataset to see which putative SNVs it caught.
@@
 We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
-(We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
+(We did not look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
 
@@
 \includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of putative SNVs in a given sample that are in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
 We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).

### 2015-11-13 FIGURES/FIGURE_SCRIPTS/Figures.R
finished figure edits

@@ -89,13 +89,13 @@ legend( 'topright', c( "overlap/WGA total", "overlap/WGS total" ), col=c( 'black
 ## FIGURE 4 (ORIG FIGURE 5 and 6): Number of putative SNPs removed by each of six filters ##
 ############################################################################################
 
-#pdf( "../FIGURE_PDFS/Figure4.pdf", width=6, height=5 )
+# pdf( "../FIGURE_PDFS/Figure4.pdf", width=6, height=5 )
 par(bty='n', mfrow=c(2,1), mar=c(5.1,4.1,1.1,0))
-boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', pch=20, ylab="Number putative SNVs filtered" )
+boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', cex.lab=0.7, pch=20, ylab="Number putative SNVs filtered" )
 axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', ylab="Percentage overlap filtered", pch=20 )
+boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', cex.lab=0.7, ylab="Percentage overlap filtered", pch=20 )
 axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-#dev.off()
+# dev.off()
 
 ##############################################################
 ## FIGURE 5 IS MADE IN THE SCRIPT Jacquard.R IN THIS FOLDER ##

### 2015-11-13 FIGURES/FIGURE_SCRIPTS/Jacquard.R

@@ -272,32 +272,32 @@ axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "
 ##############
 
 #pdf( "../FIGURE_PDFS/Jaccard1.pdf", width=6, height=5 )
-par(bty='n', mar=c(5.1,4.1,1.1,0) )
-boxplot( mergedData[,c(50,52)], ylim=c(0,450), xaxt='n', ylab=expression(paste( "Percentage ", Delta, "Jaccard( WGS, WGA )")), pch=20 )
-axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH" ) )
+#par(bty='n', mar=c(5.1,4.1,1.1,0) )
+#boxplot( mergedData[,c(50,52)], ylim=c(0,450), xaxt='n', ylab=expression(paste( "Percentage ", Delta, "Jaccard( WGS, WGA )")), pch=20 )
+#axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH" ) )
 #dev.off()
 
 #pdf( "../FIGURE_PDFS/Jaccard2.pdf", width=6, height=5 )
-par(bty='n', mar=c(5.1,4.1,1.1,0))
-boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab=expression(paste( "Percentage ", Delta, "Jaccard( WGS, WGA )")), pch=20 )
-axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
+#par(bty='n', mar=c(5.1,4.1,1.1,0))
+#boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab=expression(paste( "Percentage ", Delta, "Jaccard( WGS, WGA )")), pch=20 )
+#axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 #dev.off()
 
 #pdf( "../FIGURE_PDFS/Diffset_bypoints.pdf" )
-par(bty='n')
-plot( all$F7ratio, ylim=c(0,80), pch=20, col='black' )
-points( all$F5ratio, pch=20, col='navy' )
-points( all$F3ratio, pch=20, col='red' )
[…]

### 2015-11-13 FIGURES/FIGURE_SCRIPTS/Jacquard.R
Edits to Jaquard to move the legend away from the points: 	1.  increased ylim in the plot command 	2.  added cex parameter to the legend command.

@@ -246,12 +246,14 @@ median(all$F7ratio, na.rm=TRUE)
 
 par(bty='n', mfrow=c(2,2), mar=c(3,4.1,2,1) )
 
-plot( all$F7ratio, ylim=c(0,80), pch=20, col='black', ylab="Ratio of difference to overlap", xaxt='n', xlab="Samples" )
+# MC EDITS:  Changed ylim in the line below
+plot( all$F7ratio, ylim=c(0,90), pch=20, col='black', ylab="Ratio of difference to overlap", xaxt='n', xlab="Samples" )
 points( all$F5ratio, pch=20, col='navy' )
 points( all$F3ratio, pch=20, col='red' )
 points( all$F4ratio, pch=20, col='yellow' )
 points( all$F6ratio, pch=20, col='orange' )
-legend( 'topright', c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ), col=c( 'black', 'navy', 'yellow', 'red', 'orange' ), pch=20 )
+# MC EDITS: Added cex parameter to scale down the size of the legend.
+legend( 'topright', c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ), col=c( 'black', 'navy', 'yellow', 'red', 'orange' ), pch=20, cex=0.90 )
 
 boxplot( all[,c(26,28,25,24,27)], ylim=c(1,50), xaxt='n', pch=20, ylab="" )
 axis( 1, at=1:5, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ) )

### 2015-11-13 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure 6

@@ -114,10 +114,10 @@ names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT
 # note that this is pretty close to 1.00 every time, hence the inversion...
 LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
 
-pdf( "../FIGURE_PDFS/Figure6.pdf" )
-par(bty='n', mfrow=c(1,2) )
-plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by LOH", pch=20 )
-plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates filtered out by VAQ", pch=20 )
+pdf( "../FIGURE_PDFS/Figure6.pdf", width=10, height=5 )
+par(bty='n', mfrow=c(1,2), mar=c(5,7,4,2) )
+plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates\nfiltered out by LOH", pch=20 )
+plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates\nfiltered out by VAQ", pch=20 )
 dev.off()
 
 

### 2015-11-13 PAPER/paper_orig.tex
more changes for submission

--- 2015-11-12 PAPER/paper.tex
@@
-\documentclass[11 pt]{article} % use larger type; default would be 10pt
+\documentclass[11pt]{article} % use larger type; default would be 10pt
 \usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
@@
 \title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
-\author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
+\author[1]{Dakota Z. Derryberry\thanks{dakotaz@utexas.edu}}
 \author[2,3]{Matthew C. Cowperthwaite}
 \author[1,4]{Claus O. Wilke} 
-\affil[1]{Cell \& Molecular Biology, The University of Texas at Austin, Austin, TX USA}
-\affil[2]{St. David's NeuroTexas Institute Research Foundation, Austin, TX, USA}
-\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin, Austin TX, USA}
-\affil[4]{Integrative Biology, The University of Texas at Austin, Austin, TX, USA}
+\affil[1]{The University of Texas at Austin, Cell \& Molecular Biology}
+\affil[2]{St. David's NeuroTexas Institute Research Foundation}
+\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin}
+\affil[4]{The University of Texas at Austin, Integrative Biology}
 
@@
 
-\clearpage
-
 \section*{Introduction}
 
-Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor with a median survival time of 14 months and a 5-year survival rate of ~5\%.
+The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}.
+TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor.
+GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%.
 Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-Recently, research to improve treatments for patients with this devastating disease has focused on the idea of precision medicine, that is using large-scale genomics data to discover how the disease arises and progresses, and how to stop it.
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}, including over 500 GBMs.
-While these and other similar data \citep{Parsons}, researchers have discovered genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
-Despite widespread use of the data and adoption of the methods, however, efforts to benchmark the data--to assess the quality and repeatability of these and subsequent analyses--have lagged or been non-existent.
-Here, our goal is to begin to develop robust methods for benchmarking next-generation cancer sequencing data and analysis.
+To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
 
+However, these and other efforts have produced very few results for patients in terms of improved treatments.
+It may be that we simply haven't yet analyzed the data in the most informative manner.
+Another factor may be that the data are noisy.
 It is well known that sequencing and variant-calling pipelines are not error-free.
@@
 We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency between samples.
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
@@
 Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation is defined as a mutation that is not fixed and therefore is not present in all tumor cells.
-Such mutations may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
-Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
-In general, we would like to know how often these sorts of errors occur.
+Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a polymorphic mutation may appear fixed or completely absent.
+In general, we would like to know how often these errors occur.
 One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
@@
 One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would theoretically expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
+Despite the difference in sequencing protocol, we would expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
 Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
@@
 
-There were an average of 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:C282_v_C484}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_total_muts}).
-We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
+There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
+Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Pearson $r=0.51$, $df=53$, $P=8.35\times10^{-5}$, Figure~\ref{fig:C282_v_C484}).
+As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_overlap}).
+We further found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
+As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).
 
-% Experiment: number of mutations (figure 3 **ORIGINALLY FIGURE 4** )
+% Experiment: number of mutations (figure 4)
 %(a) a brief (one sentence to one paragraph) summary of what you did and why
 %(b) a description of what you found carrying out this experiment
-Although the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations in each sample varied by orders of magnitude (Table~\ref{tab:summary_stats}).
+Although the percent overlap in WGS and WGA samples was fairly constant across samples, and the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations varied by orders of magnitude (Table~\ref{tab:summary_stats}).
 It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
 GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample, significant alignment error, or otherwise bad data.
+One possible explanation is a degraded DNA sample or otherwise bad data.
 If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
+However, we found no significant correlation (Pearson $r=-0.04$, $df=53$, $P=0.76$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
 Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
@@
 
-% Experiment: which filters do what (figures 4 and 5) 
+% Experiment: which filters do what (figures 6 and 7) 
+% Experiment: those two filters are different (figures 8 and 9)
+%(a) a brief (one sentence to one paragraph) summary of what you did and why
+%(b) a description of what you found carrying out this experiment
 As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
@@
 The first three are based on the quality scores generated by GATK and SomaticSniper.
-We simply remove from the dataset anything that fails to meet these conventional quality-score thresholds.
+We simply remove from the dataset anything that fails to meet conventional quality-score thresholds.
 The other five are additional filters that we developed for this project.
@@
 We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We ran each filter independently on the whole dataset to see which putative SNVs it caught.
-We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}) on each sample.
+(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}).
 We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_filtered} and Table~\ref{tab:summary_stats}).
+Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}).
 The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
 
-%% start edits for Reviewer One comment 13 here%%
-We next looked at the individual effects the six of eight filters on the overlap between replicates.
-We asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
+We next asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
 Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
-To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_filtered}).
+To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}).
 We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
-By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_filtered}, Table~\ref{tab:summary_stats}).
+By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}).
 Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
 
-We next looked at the individual effects of the six of eight filters on the difference between replicates of each sample, those mutations found only in WGS \textit{or} WGA, \textit{but not} both.
-We looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$), where a higher ratio indicates that more mutations are removed from the difference than the overlap.
-We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}).
-To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}).
-We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
-(We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
-
-We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient ($WGS \cap WGA / WGS \cup WGA$) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
-By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}).
-We see this effect because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
-Removing so much of the data makes the intersection much, much smaller, and thus makes the Jaccard coefficient between samples much larger, mostly independently of the size of the overlap.
-
-%% this is figure 6
-Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most to all of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
+%% this is figures 8 and 9, which need to be added to the document and explained now...
+Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
+We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
 Either filter was removing between 0 and 100\% of the overlap for some samples.
@@
 
-We found an additional trend that was initially unexpected: The LOH and VAQ graphs (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
+We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH_all}) and the VAQ graph (Figure~\ref{fig:VAQ_all}) looked like exact inverses of each other.
 Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
@@
 First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
-Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood.
+This metric is potentially problematic for three reasons: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; and (iii) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
+Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely.
 Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
@@
 The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-PCR remains the gold standard for distinguishing true positives from false positives in sequencing, and the high rate of non-condance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, PCR remains the best method for validation.
-Nevertheless, when less fine-tuned methods are appropriate, or are the only thing possible, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
+While PCR remains the gold standard for distinguishing true positives from false positives in sequencing, we looked at the effects of six data filters that are commonly applied to validate SNVs.
 We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
@@
 
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
-The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
+We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table~\ref{tab:commands}.
+The custom python code to connect the pipeline is available in a public
+git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
 Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
@@
 
-For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
-The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
-We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Supplementary Figures 2 and 3).
-
 \subsection*{Filtering and data analysis}
@@
 We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
-Each of the eight filters was run independently on the whole data set.
 The eight filters were enacted as follows:
@@
 
+%% Data processing commands table
+\begin{landscape}
+\begin{table}
+\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
+The rows are in order of use.}
+\label{tab:commands}
+\begin{tabular}{ p{2.5cm} p{6cm} p{12cm} }
+	software & purpose & command\\
+	\hline
+	picard & regenerate fastq files from BAM file aligned to hg18 & \texttt{java -d64 -Xmx4g -jar SamToFastq.jar I=\$pfx.bam F=\$pfx.1.fastq F2=\$pfx.2.fastq 2$>$\&1} \\
+	bwa & align fastq files to hg19 & \texttt{bwa aln -q 30 -t 8 \$hgReference \$fastq $>$ \$fastq.aln.sai} \\
+	bwa, samtools & convert aligned fastq files into new BAM file & \texttt{bwa sampe -a 600 -P -r "\$RG" \$hgReference \$fastq1.aln.sai \$fastq2.aln.sai \$fastq1 \$fastq2 | samtools view -bSh -o \$outprefix.bam -} \\
+	samtools & sort and index new BAM file & \texttt{samtools sort -@ 16 \$outprefix.bam \$outprefix.sorted 2, samtools index \$outprefix.sorted.bam 2} \\
+	samtools & remove duplicate reads from BAM files & \texttt{samtools rmdup ../\$tumorpfx/\$tumorpfx.out.sorted.bam \$tumorpfx.dedup.bam} \\
+	GATK & indel realignment & \texttt{java -d64 -jar \$gatkJar -R \$hgReference -T IndelRealigner -rf BadCigar -I \$tumorpfx.dedup.bam -known \$G1000.Mills -known \$G1000.Phase1.Indels -targetIntervals \$tumorpfx.intervals -o \$tumorpfx.realn.bam} \\
+	GATK & base recalibration & \texttt{java -d64 -jar \$gatkJar -nct 8 -T BaseRecalibrator -rf BadCigar -I \$tumorpfx.realn.bam -R \$hgReference -knownSites \$dbSNP -o \$tumorpfx.recal.grp} \\
+	samtools & index recalibrated BAM file & \texttt{samtools index \$tumorpfx.realn.recal.bam} \\
+	SomaticSniper & call somatic mutations, generate VCF & \texttt{bam-somaticsniper -q 40 -Q 40 -J -s 0.001 -F vcf -f \$hgReference \$tumorbam \$normalbam \$tumorpfx.SS.vcf} \\
+\end{tabular}
+\end{table}
+\end{landscape}
+
 %% Figure 1
@@
 \centerline{
-\includegraphics[width=5in]{Figure1.pdf} }
+%\includegraphics[width=5in]{pipeline.pdf} 
+}
 \caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
@@
 \centerline{
-\includegraphics[width=5in]{Figure2.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
+%\includegraphics[width=5in]{C282_v_C484.pdf} 
+}
+\caption{\textbf{Number of putative SNVs in WGS v.\ WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
 The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
+The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Pearson $r=0.51$, $df = 53$, $P = 8.35\times 10^{-5}$).}
 \label{fig:C282_v_C484}
@@
 \centerline{
-\includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
-We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
-The correlation was calculated with respect to WGA.}
-\label{fig:unfiltered_total_muts}
+%\includegraphics[width=5in]{unfiltered_overlap_WGS_WGA_together_densities.pdf} 
+}
+\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
+The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.
+}
+\label{fig:unfiltered_overlap}
 \end{figure}
@@
 \centerline{
-\includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
-The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
-The <10\% filter removed very few putative SNVs and almost none of the overlap.
-The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
+%\includegraphics[width=5in]{unfiltered_total_muts_v_percent_overlap.pdf} 
 }
-\label{fig:boxplot_filtered}
+\caption{\textbf{Percent of overlap versus (log) number of putative SNVs per sample.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Pearson $r=-0.04$, $df=53$, $P=0.76$).}
+\label{fig:unfiltered_total_muts}
 \end{figure}
 
-%%Figure 5
+%% Figure 5
 \begin{figure}
 \centerline{
-\includegraphics[width=5in]{Figure5.pdf} }
-\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
-\textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis.
-\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
-\label{fig:diffset}
+%\includegraphics[width=6in]{boxplot_number_filtered.pdf} 
+}
+\caption{\textbf{Number of putative SNVs removed by each of six filters.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample.
+In all cases, the LOH and VAQ filters removed the majority of the putative SNVs for a given sample.}
+\label{fig:boxplot_number_filtered}
 \end{figure}
@@
 \begin{figure}
+% 379 samples, 311 patients
 \centerline{
-\includegraphics[width=5in]{Figure6.pdf} }
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
-\label{fig:LOH-VAQ_all}
+%\includegraphics[width=6in]{boxplot_percent_overlap_filtered.pdf} 
+}
+\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The $x$-axis gives the names of the filters, and the $y$-axis gives the percentage of the WGS--WGA overlap removed by the filter, per sample.
+Filters removing LOH and putative SNVs with low VAQ removed a significant percentage of the overlap; nothing else did.}
+\label{fig:boxplot_percent_overlap_filtered}
+\end{figure}
+
+%%Figure 7
+\begin{figure}
+\centerline{
+%\includegraphics[width=5in]{./LOH_VAQ/LOH_all.pdf} 
+}
+\caption{\textbf{Percentage of replicate SNVs filtered out by LOH, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by LOH increases.
+This is the almost (but not quite) perfect inverse of the pattern observed in VAQ mutations (Figure~\ref{fig:VAQ_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
+\label{fig:LOH_all}
+\end{figure}
+
+%%Figure 8
+\begin{figure}
+\centerline{
+%\includegraphics[width=5in]{./LOH_VAQ/VAQ_all.pdf} 
+}
+\caption{ \textbf{Percentage of replicate SNVs filtered out by VAQ, as a function of the total number of overlapping SNVs.} As the length of the overlap between WGS and WGA increases ($x$-axis), the percentage of the overlap filtered out by VAQ decreases.
+This is the almost (but not quite) perfect inverse of the pattern observed in LOH mutations (Figure~\ref{fig:LOH_all}), indicating that these two filters remove almost disjoint sets of putative SNVs.}
+\label{fig:VAQ_all}
 \end{figure}

### 2015-11-13 PAPER/Filters_Table2.tex
tables to upload for submission

--- 2015-11-13 PAPER/paper.tex
@@
 
-%%% Title and author are here
-\title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
-\author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
-\author[2,3]{Matthew C. Cowperthwaite}
-\author[1,4]{Claus O. Wilke} 
-\affil[1]{Cell \& Molecular Biology, The University of Texas at Austin, Austin, TX USA}
-\affil[2]{St. David's NeuroTexas Institute Research Foundation, Austin, TX, USA}
-\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin, Austin TX, USA}
-\affil[4]{Integrative Biology, The University of Texas at Austin, Austin, TX, USA}
-
 \begin{document}
-\maketitle
-
-\section*{Abstract}
-
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in calling single-nucleotide variants (SNVs).
-We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
-We further tested whether additional filtering increased or decreased the size of the overlap.
-We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966).
-We further found that using filtering after SNV-calling removed the overlap completely.
-We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
-
-\clearpage
-
-\section*{Introduction}
-
-Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor with a median survival time of 14 months and a 5-year survival rate of ~5\%.
-Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-Recently, research to improve treatments for patients with this devastating disease has focused on the idea of precision medicine, that is using large-scale genomics data to discover how the disease arises and progresses, and how to stop it.
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}, including over 500 GBMs.
-While these and other similar data \citep{Parsons}, researchers have discovered genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
-Despite widespread use of the data and adoption of the methods, however, efforts to benchmark the data--to assess the quality and repeatability of these and subsequent analyses--have lagged or been non-existent.
-Here, our goal is to begin to develop robust methods for benchmarking next-generation cancer sequencing data and analysis.
-
-It is well known that sequencing and variant-calling pipelines are not error-free.
-For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}.
-Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.
-Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls.
-One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}.
-Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
-
-We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
-For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches.
-We found significant overlap (around 50\%) between technical replicates, but also significant differences.
-As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
-This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
-
-\section*{Results}
-
-\subsection*{How similar are two replicate sequencing runs of the same tumor?}
-
-% Experiment: size of overlap (figures 3 and 4)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-
-DNA sequencing is not error-free.
-Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}.
-Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation is defined as a mutation that is not fixed and therefore is not present in all tumor cells.
-Such mutations may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
-Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
-In general, we would like to know how often these sorts of errors occur.
-One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
-However, this method is expensive and time consuming.
-A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates.
-Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
-Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
-
-TCGA's GBM data set includes 2 technical replicates for each of 55 tumors.
-In this case, the technical replicates are not identical.
-One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would theoretically expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
-Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
-We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNVs per sample, than the corresponding WGS samples.
-
-For each patient (n=55), we called mutations in both technical replicates and in the patient's blood sample, using the same computational pipeline (see Figure~\ref{fig:pipeline} and Methods): We downloaded TCGA BAM files with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the overlap, see Figure~\ref{fig:pipeline}).
-We further calculated the percentage of mutations in each replicate that occurred in the overlap between the two.
-
-There were an average of 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:C282_v_C484}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_total_muts}).
-We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
-
-% Experiment: number of mutations (figure 3 **ORIGINALLY FIGURE 4** )
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-Although the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations in each sample varied by orders of magnitude (Table~\ref{tab:summary_stats}).
-It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
-GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample, significant alignment error, or otherwise bad data.
-If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
-Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
-
-\subsection*{Does more sophisticated SNV filtering software increase or decrease the degree of similarity between replicates?}
-
-% Experiment: which filters do what (figures 4 and 5) 
-As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
-Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores.
-In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts.
-Table~\ref{tab:filters} lists eight distinct SNV filters we evaluated.
-The first three are based on the quality scores generated by GATK and SomaticSniper.
-We simply remove from the dataset anything that fails to meet these conventional quality-score thresholds.
-The other five are additional filters that we developed for this project.
-Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNV call is a false positive.
-
-We first asked whether filtering the data increases or decreases the percentage of the sample that is overlapping between the two technical replicates.
-We expect that this analysis informs us about whether filtering out putative somatic mutations with these features affects the proportion of true somatic mutations in the remaining dataset.
-We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14 (Table~\ref{tab:summary_stats}).
-The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, i.e., the overall overlap percentage also decreased.
-We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as all the noise).
-
-We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We ran each filter independently on the whole dataset to see which putative SNVs it caught.
-We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}) on each sample.
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_filtered} and Table~\ref{tab:summary_stats}).
-The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
-
-%% start edits for Reviewer One comment 13 here%%
-We next looked at the individual effects the six of eight filters on the overlap between replicates.
-We asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
-To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_filtered}).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
-By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_filtered}, Table~\ref{tab:summary_stats}).
-Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
-
-We next looked at the individual effects of the six of eight filters on the difference between replicates of each sample, those mutations found only in WGS \textit{or} WGA, \textit{but not} both.
-We looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$), where a higher ratio indicates that more mutations are removed from the difference than the overlap.
-We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}).
-To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}).
-We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
-(We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
-
-We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient ($WGS \cap WGA / WGS \cup WGA$) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
-By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}).
-We see this effect because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
-Removing so much of the data makes the intersection much, much smaller, and thus makes the Jaccard coefficient between samples much larger, mostly independently of the size of the overlap.
-
-%% this is figure 6
-Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most to all of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
-Either filter was removing between 0 and 100\% of the overlap for some samples.
-For samples with more overlap, however, LOH was the primary filter removing overlap.
-
-We found an additional trend that was initially unexpected: The LOH and VAQ graphs (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
-Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
-In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.
-
-\section*{Discussion}
-
-GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
-Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM.
-Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-These data are of enormous potential value to the research community, but their accuracy and reproducibility are unknown.
-Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
-
-We found that, on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
-The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.
-We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
-Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations.
-These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
-
-Filtering the raw computational data using both quality scores from GATK and SomaticSniper, as well as five additional custom filters, eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNVs.
-For example, LOH in this case is more likely to be an error than an actual mutation.
-When we do cancer genomics, however, our goal is the opposite, to highlight changes.
-Therefore, it is possible that we need entirely different filtering protocols.
-There are also more theoretical reasons to consider altering the filtering protocols for cancer genomes.
-For example, multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
-This hypothesis, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
-
-Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that we found in both technical replicates.
-In combination, they removed most or all of those mutations present in both replicates, since the two filters consistently removed nearly completely disjoint sets of putative mutations.
-Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative SNVs within a 10 bp window of putative indels or other putative SNVs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
-The third removes overlap with dbGaP.
-Our analysis suggests that these three filters do clean up the data in a meaningful way.
-By contrast, it may be more useful not to apply the two filters that removed principally data from the overlap of the two replicates.
-
-Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
-Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood.
-Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
-
-Second, we studied only one particular SNV caller, SomaticSniper \citep{SomaticSniper}.
-There are other, equally widely used SNV callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
-Also, we did not even consider every possible quality metric available in the SNV-caller we did choose.
-For example, we did not look at the effects of the SomaticScore or the GATK quality score individually.
-In the future, it would be worthwhile to evaluate other SNV-callers and other quality metrics.
-
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNVs in WGS samples and about 30\% in WGA samples.
-The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-We acknowledge that the high rate of non-concordance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, validation by an orthogonal method (\textit{e.g.} PCR, Sanger sequencing) remains necessary.
-Nevertheless, when less fine-tuned methods are acceptable, or are the only option available, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
-We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest that when orthogonal validation is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
-
-\section*{Methods}
-
-\subsection*{Data and back-end processing}
-
-All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
-We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
-For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
-In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
-
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
-The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
-We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
-Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
-
-For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
-The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
-We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Supplementary Figures 2 and 3).
-
-\subsection*{Filtering and data analysis}
-
-After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
-Each of the eight filters was run independently on the whole data set.
-The eight filters were enacted as follows:
-
-\begin{itemize}
-\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
-As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
-SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
-We excluded LOH variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-SNV variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
-We excluded dbSNP variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
-We excluded $<10\%$ variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\end{itemize}
-
-We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering.
-These scripts are also available in the public git repository.
-We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
-This code is also available in the public git repository.
-
-%\section*{}
-
-\bibliography{TCGA_bib}
-\bibliographystyle{peerj}
-
-\cleardoublepage
-\section*{Tables and Figures}
-
-%% Summary Statistics Table
-\begin{landscape}
-\begin{table}
-\caption{\textbf{Summary Statistics.} This table describes the results of our data processing pipeline across samples and pairs of replicates, in terms of the number of putative mutations found, length of the overlap between replicates, number of mutations removed (per sample) by each filter, and percent of the overlap removed (per pair of replicates) by each filter.}
-\label{tab:summary_stats}
-\begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
-	Quantity & average & median & min & max & stdev \\
-	\hline
-	No. mutations, WGS & 844 & 328 & 110 & 8,192 & 1,306 \\
-	No. mutations, WGA & 1,531 & 694 & 23 & 10,966 & 2,230 \\
-	\% mutations in overlap, WGS & 31\% & 31\% & 1\% & 74\% & 20\% \\
-	\% mutations in overlap, WGA & 44\% & 45\% & 3\% & 71\% & 13\% \\
-	No. putative mutations removed, VAQ & 309 & 164 & 17 & 11,439 & 1,372 \\
-	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
-	No. putative mutations removed, 10bp-SNV & 46 & 19 & 0 & 1,515 & 115 \\
-	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
-	No. putative mutations removed, dbSNP & 16 & 8 & 0 & 348 & 33 \\
-	No. putative mutations removed, $<$10\% & 1 & 1 & 1 & 36 & 3 \\
-	\% overlap removed, VAQ & 53\% & 57\% & 0\% & 100\% & 28\% \\
-	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
-	\% overlap removed, 10bp-SNV & 3\% & 2\% & 0\% & 20\% & 3\% \\
-	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
-	\% overlap removed, dbSNP & 3\% & 2\% & 0\% & 47\% & 7\% \\
-\end{tabular}
-\end{table}
-\end{landscape}
 
@@
 \begin{table}
-\caption{\textbf{SNV prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
+%\caption{\textbf{SNV prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
 \label{tab:filters}
@@
 
-%% Figure 1
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure1.pdf} }
-\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
-For each BAM file, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
-We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
-\label{fig:pipeline}
-\end{figure}
-
-%% Figure 2
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure2.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
-The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
-\label{fig:C282_v_C484}
-\end{figure}
-
-%% Figure 3
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
-We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
-The correlation was calculated with respect to WGA.}
-\label{fig:unfiltered_total_muts}
-\end{figure}
-
-%% Figure 4
-\begin{figure}
-\centerline{
-\includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
-The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
-The <10\% filter removed very few putative SNVs and almost none of the overlap.
-The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
-}
-\label{fig:boxplot_filtered}
-\end{figure}
-
-%%Figure 5
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure5.pdf} }
-\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
-\textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis.
-\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
-\label{fig:diffset}
-\end{figure}
-
-%%Figure 6
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure6.pdf} }
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
-\label{fig:LOH-VAQ_all}
-\end{figure}
-
 \end{document}

### 2015-11-13 PAPER/Summary_Stats_Table1.tex
tables to upload for submission

--- 2015-11-13 PAPER/paper.tex
@@
 
-%%% Title and author are here
-\title{Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
-\author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
-\author[2,3]{Matthew C. Cowperthwaite}
-\author[1,4]{Claus O. Wilke} 
-\affil[1]{Cell \& Molecular Biology, The University of Texas at Austin, Austin, TX USA}
-\affil[2]{St. David's NeuroTexas Institute Research Foundation, Austin, TX, USA}
-\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin, Austin TX, USA}
-\affil[4]{Integrative Biology, The University of Texas at Austin, Austin, TX, USA}
-
 \begin{document}
-\maketitle
-
-\section*{Abstract}
-
-We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in calling single-nucleotide variants (SNVs).
-We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates.
-We further tested whether additional filtering increased or decreased the size of the overlap.
-We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966).
-We further found that using filtering after SNV-calling removed the overlap completely.
-We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.
-
-\clearpage
-
-\section*{Introduction}
-
-Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor with a median survival time of 14 months and a 5-year survival rate of ~5\%.
-Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}.
-Recently, research to improve treatments for patients with this devastating disease has focused on the idea of precision medicine, that is using large-scale genomics data to discover how the disease arises and progresses, and how to stop it.
-The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}, including over 500 GBMs.
-While these and other similar data \citep{Parsons}, researchers have discovered genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.
-Despite widespread use of the data and adoption of the methods, however, efforts to benchmark the data--to assess the quality and repeatability of these and subsequent analyses--have lagged or been non-existent.
-Here, our goal is to begin to develop robust methods for benchmarking next-generation cancer sequencing data and analysis.
-
-It is well known that sequencing and variant-calling pipelines are not error-free.
-For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}.
-Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low.
-Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls.
-One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}.
-Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue.
-And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
-
-We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol.
-For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using various SNV-filtering approaches.
-We found significant overlap (around 50\%) between technical replicates, but also significant differences.
-As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
-This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
-Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
-
-\section*{Results}
-
-\subsection*{How similar are two replicate sequencing runs of the same tumor?}
-
-% Experiment: size of overlap (figures 3 and 4)
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-
-DNA sequencing is not error-free.
-Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}.
-Loss of heterozygosity may be a real feature of the data, or an amplification artifact that occurs when only one allele of a polymorphic site is amplified.
-Cancer DNA is highly heterogenous, which makes for an additional source of error: A polymorphic mutation is defined as a mutation that is not fixed and therefore is not present in all tumor cells.
-Such mutations may or may not be represented at high enough frequency in a particular tumor specimen to be identified with next-gen sequencing.
-Alternatively, a polymorphic mutation may appear to be fixed when present at at very high frequency in a tumor specimen.
-In general, we would like to know how often these sorts of errors occur.
-One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation.
-However, this method is expensive and time consuming.
-A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates.
-Theoretically, any fixed mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not.
-(Polymorphic mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to unavailability of additional pathology specimens and the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research.
-Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.
-
-TCGA's GBM data set includes 2 technical replicates for each of 55 tumors.
-In this case, the technical replicates are not identical.
-One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate.
-Despite the difference in sequencing protocol, we would theoretically expect any fixed mutation to appear in both replicates, while polymorphic mutations and sequencing error might appear in only one.
-Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate.
-We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNVs per sample, than the corresponding WGS samples.
-
-For each patient (n=55), we called mutations in both technical replicates and in the patient's blood sample, using the same computational pipeline (see Figure~\ref{fig:pipeline} and Methods): We downloaded TCGA BAM files with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}.
-We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the overlap, see Figure~\ref{fig:pipeline}).
-We further calculated the percentage of mutations in each replicate that occurred in the overlap between the two.
-
-There were an average of 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (Table~\ref{tab:summary_stats}).
-Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in its corresponding WGA replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$, Figure~\ref{fig:C282_v_C484}).
-As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:C282_v_C484} and~\ref{fig:unfiltered_total_muts}).
-We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
-
-% Experiment: number of mutations (figure 3 **ORIGINALLY FIGURE 4** )
-%(a) a brief (one sentence to one paragraph) summary of what you did and why
-%(b) a description of what you found carrying out this experiment
-Although the numbers of putative mutations in WGS and WGA samples were correlated, the exact number of putative mutations in each sample varied by orders of magnitude (Table~\ref{tab:summary_stats}).
-It is known that different cancers mutate at different rates: Some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly increased numbers of mutations, usually resulting from errors in DNA repair pathways \citep{mutator}.
-GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome).
-One possible explanation is a degraded DNA sample, significant alignment error, or otherwise bad data.
-If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations.
-However, we found no significant correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}).
-Thus, our data suggest that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.
-
-\subsection*{Does more sophisticated SNV filtering software increase or decrease the degree of similarity between replicates?}
-
-% Experiment: which filters do what (figures 4 and 5) 
-As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}.
-Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations.
-The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores.
-In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts.
-Table~\ref{tab:filters} lists eight distinct SNV filters we evaluated.
-The first three are based on the quality scores generated by GATK and SomaticSniper.
-We simply remove from the dataset anything that fails to meet these conventional quality-score thresholds.
-The other five are additional filters that we developed for this project.
-Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNV call is a false positive.
-
-We first asked whether filtering the data increases or decreases the percentage of the sample that is overlapping between the two technical replicates.
-We expect that this analysis informs us about whether filtering out putative somatic mutations with these features affects the proportion of true somatic mutations in the remaining dataset.
-We found that, after removing putative mutations tagged by any one of the eight filters, the number of putative mutations per replicate decreased from 23--10,966 to 0--14 (Table~\ref{tab:summary_stats}).
-The size of the overlap between technical replicates decreased to 0--2 per sample, with 0 as the mode, i.e., the overall overlap percentage also decreased.
-We concluded that running all the filters on the data, in the absence of any other verification method, was counterproductive, because it removed all of the signal (as well as all the noise).
-
-We next looked at the individual effects of six of the eight filters.
-(We did not specifically study the GATK and SS filters, since both are directly linked to read quality.) We ran each filter independently on the whole dataset to see which putative SNVs it caught.
-We first considered the total number of putative SNVs removed by each filter (Figure~\ref{fig:boxplot_filtered}) on each sample.
-We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ (Variant Allele Quality) and LOH (Loss of Heterozygosity) filters, which removed on average 309 and 539 putative SNVs, respectively (Table~\ref{tab:summary_stats}).
-Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNVs, removed 16, 28, and 46 putative mutations per sample, respectively (Figure~\ref{fig:boxplot_filtered} and Table~\ref{tab:summary_stats}).
-The final filter, which removed putative SNVs with less than 10\% coverage of the alternate allele, removed 1 putative SNV on average.
-
-%% start edits for Reviewer One comment 13 here%%
-We next looked at the individual effects the six of eight filters on the overlap between replicates.
-We asked how many of the SNVs removed by each filter were SNVs present in the overlap between technical replicates, and how many were in just one sample?
-Put differently, what percentage of putative SNVs removed by a given filter was in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only present in one replicate)?
-To answer this question, we plotted the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_filtered}).
-We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNV removed, on average, only 3\%--4\% of the overlap (Table~\ref{tab:summary_stats}).
-By contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, respectively (Figure~\ref{fig:boxplot_filtered}, Table~\ref{tab:summary_stats}).
-Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, while the filters removing low VAQ and LOH are less discriminatory and may be removing a large proportion of the true positives.
-
-We next looked at the individual effects of the six of eight filters on the difference between replicates of each sample, those mutations found only in WGS \textit{or} WGA, \textit{but not} both.
-We looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$), where a higher ratio indicates that more mutations are removed from the difference than the overlap.
-We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}).
-To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}).
-We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order.
-(We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.
-
-We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient ($WGS \cap WGA / WGS \cup WGA$) before and after filtering (Jaccard after-Jaccard before /Jaccard before).
-By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}).
-We see this effect because these two filters remove so much data: half or more of the overlap and as much or more of the difference.
-Removing so much of the data makes the intersection much, much smaller, and thus makes the Jaccard coefficient between samples much larger, mostly independently of the size of the overlap.
-
-%% this is figure 6
-Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most to all of the overlap, was more likely to remove overlap for samples that showed a lot of overlap.
-We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend.
-Either filter was removing between 0 and 100\% of the overlap for some samples.
-For samples with more overlap, however, LOH was the primary filter removing overlap.
-
-We found an additional trend that was initially unexpected: The LOH and VAQ graphs (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other.
-Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases.
-In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.
-
-\section*{Discussion}
-
-GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole.
-Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM.
-Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers.
-These data are of enormous potential value to the research community, but their accuracy and reproducibility are unknown.
-Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.
-
-We found that, on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates.
-The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations.
-We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates.
-Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations.
-These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
-
-Filtering the raw computational data using both quality scores from GATK and SomaticSniper, as well as five additional custom filters, eliminated more than half of the total number of putative mutations in all 110 samples, including most or all of those present in both replicates.
-To some extent, this result was unsurprising: These filters were designed to be used on wild type genomes, where it is generally assumed that any observed differences are more likely due to error than to the presence of true SNVs.
-For example, LOH in this case is more likely to be an error than an actual mutation.
-When we do cancer genomics, however, our goal is the opposite, to highlight changes.
-Therefore, it is possible that we need entirely different filtering protocols.
-There are also more theoretical reasons to consider altering the filtering protocols for cancer genomes.
-For example, multiple sources suggest that LOH mutations may be essential to cancer \citep{LOH}.
-This hypothesis, along with the data presented here, makes a strong argument for retaining these mutations in functional analyses rather than excluding them.
-
-Of the six filters whose individual effects we examined, only two, those that removed Loss of Heterozygosity (LOH) mutations and putative mutations with low VAQ (calculated by SomaticSniper), removed primarily mutations that we found in both technical replicates.
-In combination, they removed most or all of those mutations present in both replicates, since the two filters consistently removed nearly completely disjoint sets of putative mutations.
-Of the remaining four filters, only three removed any appreciable number of putative mutations from the sample, and each of these preferentially removed mutations present in only one technical replicate.
-Two of these three filters, those removing putative SNVs within a 10 bp window of putative indels or other putative SNVs, recognize a feature (clustered mutations) that suggests a local problem with the reads or alignment.
-The third removes overlap with dbGaP.
-Our analysis suggests that these three filters do clean up the data in a meaningful way.
-By contrast, it may be more useful not to apply the two filters that removed principally data from the overlap of the two replicates.
-
-Several factors limit the conclusions we may draw from this analysis.
-First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV.
-This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}.
-Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood.
-Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
-
-Second, we studied only one particular SNV caller, SomaticSniper \citep{SomaticSniper}.
-There are other, equally widely used SNV callers available \citep{MuTect, VarScan, Strelka}, which might do better or worse than the one we have chosen here.
-Also, we did not even consider every possible quality metric available in the SNV-caller we did choose.
-For example, we did not look at the effects of the SomaticScore or the GATK quality score individually.
-In the future, it would be worthwhile to evaluate other SNV-callers and other quality metrics.
-
-We have shown that there is significant overlap between technical replicates of whole exome sequencing in the TCGA GBM dataset, comprising about 50\% of putative SNVs in WGS samples and about 30\% in WGA samples.
-The overlap exists even for samples with a high number of putative SNVs, suggesting that some GBMs may have significantly more somatic mutation than others.
-We acknowledge that the high rate of non-concordance between replicates in our analyses indicates that even the best computational analyses are insufficient to validate any single SNV; when the identity of a single SNV in a single sample is important, validation by an orthogonal method (\textit{e.g.} PCR, Sanger sequencing) remains necessary.
-Nevertheless, when less fine-tuned methods are acceptable, or are the only option available, one may wish to employ other methods of validation rather than nothing, such as the six data filters that are commonly applied to validate SNVs that we examined.
-We found that some of these filters remove principally those mutations found in one sample or the other, while other filers remove primarily those in the overlap.
-We suggest that when orthogonal validation is not an option, only the filters that removed little overlap between samples should be used for computational SNV validation.
-
-\section*{Methods}
-
-\subsection*{Data and back-end processing}
-
-All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
-We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
-For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
-In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
-
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
-The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
-We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
-Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
-
-For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
-The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
-We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Supplementary Figures 2 and 3).
-
-\subsection*{Filtering and data analysis}
-
-After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
-Each of the eight filters was run independently on the whole data set.
-The eight filters were enacted as follows:
-
-\begin{itemize}
-\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
-As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
-SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
-We excluded LOH variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-SNV variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
-We excluded dbSNP variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
-We excluded $<10\%$ variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\end{itemize}
-
-We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering.
-These scripts are also available in the public git repository.
-We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
-This code is also available in the public git repository.
-
-%\section*{}
-
-\bibliography{TCGA_bib}
-\bibliographystyle{peerj}
-
-\cleardoublepage
-\section*{Tables and Figures}
 
@@
 \begin{table}
-\caption{\textbf{Summary Statistics.} This table describes the results of our data processing pipeline across samples and pairs of replicates, in terms of the number of putative mutations found, length of the overlap between replicates, number of mutations removed (per sample) by each filter, and percent of the overlap removed (per pair of replicates) by each filter.}
+%\caption{\textbf{Summary Statistics.} This table describes the results of our data processing pipeline across samples and pairs of replicates, in terms of the number of putative mutations found, length of the overlap between replicates, number of mutations removed (per sample) by each filter, and percent of the overlap removed (per pair of replicates) by each filter.}
 \label{tab:summary_stats}
@@
 
-%% Filters table
-\begin{landscape}
-\begin{table}
-\caption{\textbf{SNV prediction filters.} This table shows the various methods (filters) used to predict which differences found in tumor alignments relative to blood alignments are real somatic variants, as opposed to sequencing errors or other variants.}
-\label{tab:filters}
-\begin{tabular}{ p{2.25cm} p{4.25cm} p{14cm} }
-	filter & software & purpose \\
-	\hline
-	GATK & GATK & Removes putative SNVs with GATK quality scores less than 40 (as part of the GATK processing, with indel realignment and base recalibration) \\
-	SS & SomaticSniper & Removes putative SNVs with a SomaticScore less than 40 \\
-	VAQ & SomaticSniper & Removes putative SNVs with SomaticSniper Varaint Allele Quality scores less than 20 \\
-	LOH & SomaticSniper, python & Removes putative SNVs that are identified as loss of heterozygosity \\
-	10bp-SNV & python & Removes putative SNVs located within a 10 bp window of any other putative SNV \\
-	10bp-INDEL & python & Removes putative SNVs located within a 10 bp window of indels \\
-	dbSNP & python & Removes putative SNVs that overlap with dbSNP coverage \\
-	$<$10\% & python & Removes putative SNVs if, in the tumor data, the percentage of reads covering the site with the alternate allele is less than 10\% \\
-\end{tabular}
-\end{table}
-\end{landscape}
-
-%% Figure 1
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure1.pdf} }
-\caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
-For each BAM file, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
-We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
-\label{fig:pipeline}
-\end{figure}
-
-%% Figure 2
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure2.pdf} }
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
-The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
-\label{fig:C282_v_C484}
-\end{figure}
-
-%% Figure 3
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure3.pdf} }
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of a given WGS sample that is in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
-We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
-The correlation was calculated with respect to WGA.}
-\label{fig:unfiltered_total_muts}
-\end{figure}
-
-%% Figure 4
-\begin{figure}
-\centerline{
-\includegraphics[width=6in]{Figure4.pdf} }
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
-The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
-The <10\% filter removed very few putative SNVs and almost none of the overlap.
-The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
-}
-\label{fig:boxplot_filtered}
-\end{figure}
-
-%%Figure 5
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure5.pdf} }
-\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
-\textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis.
-\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
-\label{fig:diffset}
-\end{figure}
-
-%%Figure 6
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure6.pdf} }
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
-\label{fig:LOH-VAQ_all}
-\end{figure}
-
 \end{document}

### 2015-11-13 PAPER/TableS1.tex
tableS1

--- 2015-10-01 PAPER/Supplementary.tex
@@
 
-%%% Title and author are here
-\title{Supplementary materials: Reproducibility of SNV-calling in multiple sequencing runs from single tumors}
-\author[1]{Dakota Z. Derryberry\thanks{Dakota Z. Derryberry: 2500 Speedway, Austin, TX, 78712; (512)232-2459; dakotaz@utexas.edu}}
-\author[2,3]{Matthew C. Cowperthwaite}
-\author[1,4]{Claus O. Wilke} 
-\affil[1]{Cell \& Molecular Biology, The University of Texas at Austin, Austin, TX USA}
-\affil[2]{St. David's NeuroTexas Institute Research Foundation, Austin, TX, USA}
-\affil[3]{Center for Systems and Synthetic Biology, The University of Texas at Austin, Austin TX, USA}
-\affil[4]{Integrative Biology, The University of Texas at Austin, Austin, TX, USA}
-
 \begin{document}
-\maketitle
-
-\section{Figures}
-
-%% Figure 2
-\begin{figure}
-\centerline{
-\includegraphics[width=5in]{orig_Figure3.pdf} }
-\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
-The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.}
-\label{fig:density}
-\end{figure}
-
-\section{Tables}
 
@@
 \begin{table}
-\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
+%\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
 The rows are in order of use.}

### 2015-11-13 PAPER/paper.tex
more changes for submission
manuscript with changes highlighted.
final submission ready

@@
 We found that the percent overlap between the two samples, calculated as $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, varied widely, from 1\%-74\%, but was fairly evenly distributed around the average of 31\% in WGA replicates and 44\% in WGS replicates (Table~\ref{tab:summary_stats}).
-As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Supplementary Figure 1).
+As expected, the distribution of the percentage overlap was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure S1).
 
@@
 
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Supplementary Table 1.
+We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table S1.
 The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
@@
 The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
-We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Supplementary Figures 2 and 3).
+We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Figures S2 and S3).
 
@@
 \centerline{
-\includegraphics[width=5in]{Figure1.pdf} }
+\includegraphics[width=5in]{Figure1.pdf} 
+}
 \caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
@@
 \centerline{
-\includegraphics[width=5in]{Figure2.pdf} }
+\includegraphics[width=5in]{Figure2.pdf} 
+}
 \caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
@@
 \centerline{
-\includegraphics[width=5in]{Figure3.pdf} }
+\includegraphics[width=5in]{Figure3.pdf} 
+}
 \caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of putative SNVs in a given sample that are in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
@@
 \centerline{
-\includegraphics[width=6in]{Figure4.pdf} }
+\includegraphics[width=6in]{Figure4.pdf} 
+}
 \caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
@@
 \centerline{
-\includegraphics[width=5in]{Figure5.pdf} }
+\includegraphics[width=5in]{Figure5.pdf} 
+}
 \caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
@@
 \centerline{
-\includegraphics[width=5in]{Figure6.pdf} }
+\includegraphics[width=5in]{Figure6.pdf} 
+}
 \caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.

### 2015-11-23 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure 3

@@ -62,11 +62,11 @@ abline(0,1)
 #pdf( "../FIGURE_PDFS/Figure3.pdf" )
 # when necessary, use the following title: Number of putative SNPs does not correlate with percentage WGS/WGA overlap
 par(bty = 'n')
-plot( data34$TOTAL, data34$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, main="", xlab="Number of putative SNVs in unfiltered data (log)", ylab="Percentage overlap between WGS and WGA in unfiltered data" )
-points ( data34$TOTAL, data34$PERCENT_OVERLAP_by_WGS.484, pch=20, col="red" )
+plot( data34$TOTAL, data34$PERCENT_OVERLAP_by_WGA.282, log='x', pch=20, col="orange", main="", xlab="Number of putative SNVs in unfiltered data (log)", ylab="Percentage overlap between WGS and WGA in unfiltered data" )
+points ( data34$TOTAL, data34$PERCENT_OVERLAP_by_WGS.484, pch="+", col="green" )
 abline( 0,1 )
-legend( 'topright', c( "overlap/WGA total", "overlap/WGS total" ), col=c( 'black', 'red' ) )
-legend( 'topright', c( "overlap/WGA total", "overlap/WGS total" ), col=c( 'black', 'red' ), pch=20 )
+legend( 'topleft', c( "overlap/WGS total", "overlap/WGA total" ), col=c( 'green', 'orange' ) )
+legend( 'topleft', c( "overlap/WGS total", "overlap/WGA total" ), col=c( 'green', 'orange' ), pch=c(3,20) )
 #dev.off()
 
 ############################

### 2015-11-23 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure S1

@@ -43,17 +43,18 @@ abline(0,1)
 
 ##################################################################################
 ## FIGURE 3 ORIGINAL: Density of percentage overlap between WGS and WGA samples ##
+## THIS IS NOW FIGURE S1														##
 ##################################################################################
 
-# pdf( "../FIGURE_PDFS/Figure3.pdf" )
+#pdf( "../FIGURE_PDFS/FigureS1.pdf" )
 # figure title (should it ever be needed): Greater % of WGS sample found in overlap than WGA
-# amp_frac = density( data34$OVERLAP/data34$TOTAL_WGA.C282., from=0 )
-# seq_frac = density( data34$OVERLAP/data34$TOTAL_WGS.C484., from=0 )
-# par(bty='n')
-# plot( seq_frac$x, seq_frac$y, type='l', col='blue', main="", xlab="Percentage of putative SNVs in a sample also recovered in the samples's technical replicate", ylab="Density")
-# legend( "topright", c( "WGS", "WGA" ), col=c( "blue", "red" ), lty=1 )
-# lines( amp_frac$x, amp_frac$y, col='red' )
-# dev.off()
+amp_frac = density( data34$OVERLAP/data34$TOTAL_WGA.C282., from=0 )
+seq_frac = density( data34$OVERLAP/data34$TOTAL_WGS.C484., from=0 )
+par(bty='n')
+plot( seq_frac$x, seq_frac$y, type='l', col='orange', main="", xlab="Percentage of putative SNVs in a sample also recovered in its technical replicate", ylab="Density")
+legend( "topright", c( "WGS", "WGA" ), col=c( "green", "orange" ), lty=1 )
+lines( amp_frac$x, amp_frac$y, col='green' )
+#dev.off()
 
 #############################################################################################################
 ## FIGURE 3 MOD (ORIGINAL FIGURE 4): Percentage of overlap versus (log) number of putative SNVs per sample ##

### 2015-11-23 PAPER/Supplementary.tex
finished edits to Supplementary (captions)

--- 2015-10-01 PAPER/Supplementary.tex
@@
 
-\section{Figures}
+%\section{Figures}
 
@@
 \centerline{
-\includegraphics[width=5in]{orig_Figure3.pdf} }
-\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (blue) and WGA (red) sample that is present in the overlap between replicates for each patient.
+\includegraphics[width=5in]{FigureS1.pdf} }
+\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (green) and WGA (orange) sample that is present in the overlap between replicates for each patient.
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.}
@@
 
-\section{Tables}
+\begin{figure}
+\centerline{
+\includegraphics[width=5in]{FigureS2.pdf} }
+\caption{\textbf{Number of putative SNVs in a sample does not correlate with coverage.} The number of SNVs called in a sample does not correlate with the coverage of that sample  (Spearman $\rho=-0.13$, $S=671817$, $P=0.12$).
+While at first glance, it appears that there may be significant separation between WGS and WGA samples along the $y$-axis (WGA samples lie above and below WGS samples), this is not the case because the $y$-axis value (coverage) is an experimental and not a test parameter.
+It does not impact our results.}
+\label{fig:coverage}
+\end{figure}
+
+\begin{figure}
+\centerline{
+\includegraphics[width=5in]{FigureS3.pdf} }
+\caption{\textbf{Number of putative SNVs in a sample does not correlate with percentage of mapped reads.} The number of SNVs called in a sample does not correlate with the percentage of mapped reads in the alignment of that sample  (Spearman $\rho=-0.068$, $S=637326$, $P=0.41$).}
+\label{fig:percent_mapped}
+\end{figure}
+
+%\section{Tables}
 

### 2015-11-23 FIGURES/FIGURE_SCRIPTS/Figures.R
fixed figure5

@@ -115,11 +115,11 @@ names( LOH_VAQ_data )[ names( LOH_VAQ_data )=="FILT_7_PERCENT" ] <- "LOH_PERCENT
 # note that this is pretty close to 1.00 every time, hence the inversion...
 LOH_VAQ_data$SUM=LOH_VAQ_data$LOH_PERCENT + LOH_VAQ_data$VAQ_PERCENT
 
-pdf( "../FIGURE_PDFS/Figure6.pdf", width=10, height=5 )
+#pdf( "../FIGURE_PDFS/Figure6.pdf", width=10, height=5 )
 par(bty='n', mfrow=c(1,2), mar=c(5,7,4,2) )
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$LOH_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates\nfiltered out by LOH", pch=20 )
 plot( LOH_VAQ_data$OVERLAP, LOH_VAQ_data$VAQ_PERCENT, xlab="Number of SNVs in both replicates", ylab="Percentage of SNVs in both replicates\nfiltered out by VAQ", pch=20 )
-dev.off()
+#dev.off()
 
 
 ##############

### 2015-11-23 FIGURES/FIGURE_SCRIPTS/Jacquard.R

@@ -242,7 +242,7 @@ median(all$F7ratio, na.rm=TRUE)
 ## THEN MAKE FIGURE 6 (COMPOUND FIGURE) FOR THE PAPER ##
 ########################################################
 
-#pdf( "../FIGURE_PDFS/Figure5.pdf", width=6, height=5 )
+#pdf( "../FIGURE_PDFS/Figure5.pdf", width=9, height=7 )
 
 par(bty='n', mfrow=c(2,2), mar=c(3,4.1,2,1) )
 
@@ -255,13 +255,13 @@ points( all$F6ratio, pch=20, col='orange' )
 # MC EDITS: Added cex parameter to scale down the size of the legend.
 legend( 'topright', c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ), col=c( 'black', 'navy', 'yellow', 'red', 'orange' ), pch=20, cex=0.90 )
 
-boxplot( all[,c(26,28,25,24,27)], ylim=c(1,50), xaxt='n', pch=20, ylab="" )
+boxplot( all[,c(26,28,25,24,27)], ylim=c(1,90), xaxt='n', pch=20, ylab="Ratio of difference to overlap" )
 axis( 1, at=1:5, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP" ) )
 
 boxplot( mergedData[,c(52,50)], ylim=c(0,450), xaxt='n', ylab=expression(paste( Delta, "Jaccard( WGS, WGA )")), pch=20 )
 axis( 1, at=1:2, padj = 0.8, cex.axis=0.7, labels=c( "LOH", "VAQ" ) )
 
-boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab="", pch=20 )
+boxplot( mergedData[,c(49,48,51,53)], ylim=c(0,15), xaxt='n', ylab=expression(paste( Delta, "Jaccard( WGS, WGA )")), pch=20 )
 axis( 1, at=1:4, padj = 0.8, cex.axis=0.7, labels=c( "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 
 #dev.off()

### 2015-11-23 PAPER/Supplementary.tex
finished editing legends in Supplementary.tex

@@
 
-%% Figure 2
+%% Figure S1
 \begin{figure}
@@
 \includegraphics[width=5in]{FigureS1.pdf} }
-\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} Density of the percentage of each WGS (green) and WGA (orange) sample that is present in the overlap between replicates for each patient.
+\caption{\textbf{One third (WGA) to one half (WGS) of putative SNVs were recovered in technical replicates.} For each pair of replicates (WGS and WGA), we looked at the percentage of WGS SNVs that were recovered in the WGA sample (about one half, in green), and the percentage of the WGA SNVs that were recovered in the WGS sample (about one third, in orange).
 The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter.}
@@
 
+%% Figure S2
 \begin{figure}
@@
 \caption{\textbf{Number of putative SNVs in a sample does not correlate with coverage.} The number of SNVs called in a sample does not correlate with the coverage of that sample  (Spearman $\rho=-0.13$, $S=671817$, $P=0.12$).
-While at first glance, it appears that there may be significant separation between WGS and WGA samples along the $y$-axis (WGA samples lie above and below WGS samples), this is not the case because the $y$-axis value (coverage) is an experimental and not a test parameter.
-It does not impact our results.}
+This is shown by the consistent variability along the $x$-axis at each level of overage (shown on the $y$-axis).
+The separation of the two experimental condition on the $y$-axis is not relevant to this measure.}
 \label{fig:coverage}
@@
 
+%% Figure S3
 \begin{figure}
@@
 \begin{table}
-\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with it.
+\caption{\textbf{Back-end Processing.} This table shows the software packages we used in data processing, what we used each piece of software for, and the command associated with the use of it.
 The rows are in order of use.}

### 2015-11-23 PAPER/paper.tex
removed some internal comments
fixed figure legends 1, 2, 3

@@
 Our attempt to use SNV filtering, either via SomaticSniper \citep{SomaticSniper} or via custom-built filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.
+
+\section*{Methods}
+
+\subsection*{Data and back-end processing}
+
+All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
+We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
+For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
+In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
+
+We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table S1.
+The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
+Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
+Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
+We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
+Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
+
+For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
+The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
+We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Figures S2 and S3).
+
+\subsection*{Filtering and data analysis}
+
+After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
+We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
+Each of the eight filters was run independently on the whole data set.
+The eight filters were enacted as follows:
+
+\begin{itemize}
+\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
+We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
+\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
+As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
+We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
+\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
+SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
+We excluded LOH variants using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+We excluded 10bp-SNV variants using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
+We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
+We excluded dbSNP variants using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
+We excluded $<10\%$ variants using a custom python script available in the public git repository.
+This recommendation is discussed in the Results and Discussion.
+\end{itemize}
+
+We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering.
+These scripts are also available in the public git repository.
+We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
+This code is also available in the public git repository.
 
@@
 
-\section*{Methods}
-
-\subsection*{Data and back-end processing}
-
-All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}.
-We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}.
-For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate.
-In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.
-
-We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table S1.
-The custom python code to connect the pipeline is available in a public git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}).
-Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}.
-Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file.
-We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}.
-Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.
-
-For all alignments, average read coverage was 30X, with a low of 5X and a high of 60X (stdev=15.95898).
-The percentage of mapped reads in our alignments was universally high (mean=98.03545, stdev=3.257872), with only two samples below 90\% (68\% and 74\%), and only 15 samples below 98\%.
-We found no correlation between the read coverage or the percentage of mapped reads and the total number of SNVs called (Figures S2 and S3).
-
-\subsection*{Filtering and data analysis}
-
-After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
-We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
-Each of the eight filters was run independently on the whole data set.
-The eight filters were enacted as follows:
-
-\begin{itemize}
-\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper.
-As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2).
-We did not at any point consider putative mutations removed by this filter, and did not consider its individual action.
-\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type.
-SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors.
-We excluded LOH variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-SNV:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-SNV variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment.
-We excluded 10bp-INDEL variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant.
-We excluded dbSNP variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\item[$<10\%$:] It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error.
-We excluded $<10\%$ variants using a custom python script available in the public git repository.
-This recommendation is discussed in the Results and Discussion.
-\end{itemize}
-
-We used the same or substantially similar python scripts to calculate and compare the overlap between technical replicates before and after filtering.
-These scripts are also available in the public git repository.
-We plotted all data and did all statistics with standard plotting and statistics functions in R \citep{Rsoftware}.
-This code is also available in the public git repository.
-
 %\section*{}
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure1.pdf} 
-}
+%\centerline{
+%\includegraphics[width=5in]{Figure1.pdf} 
+%}
 \caption{\textbf{Data processing pipeline.} For each of 55 patients, we began with a C484 tumor BAM file (WGS), a C282 tumor BAM file (WGA), and a normal BAM file, all aligned to hg18.
-For each BAM file, we used picard to regenerate fastq files, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
+For each BAM file, we used picard to regenerate fastq reads, bwa to realign the fastq files to hg19, and GATK to recalibrate bases and indels.
 We used SomaticSniper to call somatic mutations (differences between the tumor and normal sequences) for each replicate.
-When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations which appeared in both replicates.}
+When we had a VCF for each replicate, we calculated the overlap between the two lists as the number of individual mutations that appeared in both replicates.}
 \label{fig:pipeline}
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure2.pdf} 
+%\centerline{
+%\includegraphics[width=5in]{Figure2.pdf} 
+%}
+\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} For each patient (represented by one point), we compared the number of SNVs called in the WGS sample ($y$-axis) versus the number of SNVs called in the WGA sample ($x$-axis).
+As expected, the number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).
+The line shown is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
 }
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} Each point represents data for a single patient.
-The line is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
-The number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).}
 \label{fig:C282_v_C484}
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure3.pdf} 
-}
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} The percentage of putative SNVs in a given sample that are in the overlap between replicates is not correlated to the number of mutations in that sample (Spearman $\rho=0.05$, $S=29268$, $P=0.68$).
+%\centerline{
+%\includegraphics[width=5in]{Figure3.pdf} 
+%}
+\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} To test the hypothesis that samples with a greater number of SNVs also have a greater number of errors, and consequently a smaller percentage of SNVs recovered in the overlap between replicates, we compared the percentage of putative SNVs recovered in the overlap between replicates for a sample ($y$-axis) versus the number of mutations in that sample ($x$-axis).
+We found no correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$), suggesting that higher numbers of SNVs in a sample may not be only due to errors.
 We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=6in]{Figure4.pdf} 
-}
+%\centerline{
+%\includegraphics[width=6in]{Figure4.pdf} 
+%}
 \caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure5.pdf} 
-}
+%\centerline{
+%\includegraphics[width=5in]{Figure5.pdf} 
+%}
 \caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=5in]{Figure6.pdf} 
-}
+%\centerline{
+%\includegraphics[width=5in]{Figure6.pdf} 
+%}
 \caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.

### 2015-11-23 PAPER/paper.tex
tracking rebuttal2 letter

@@
 %}
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample.
+\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} For each filter (names given on the $x$-axis, with detailed description in Table 1), we looked at the number of mutations removed by a given filter in a given sample on a log scale \textit{(above)}, and the percentage of the WGS--WGA overlap removed by the filter \textit{(below)}, per sample.
 The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
@@
 %}
-\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
+\caption{\textbf{Independent performance of individual filters of the difference between replicates.} For each sample, we compared the size of the difference between replicates (the number of SNVs found in only one of WGS or WGA) to the size of the overlap between replicates (the number of SNVs recovered from both the WGA and WGS replicates).
+\textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
 \textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis.
-\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
+\textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA), with Jaccard=$|WGS \cap WGA|/|WGS \cup WGA|$, before and after the action of each filter (each filter was run on the whole data set independently).}
 \label{fig:diffset}
@@
 %}
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} As the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses; putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
+\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} Across samples, as the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
+The two are almost (but not quite) perfect inverses: putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
 \label{fig:LOH-VAQ_all}

### 2015-11-23 FIGURES/FIGURE_SCRIPTS/Figures.R

@@ -90,13 +90,13 @@ legend( 'topleft', c( "overlap/WGS total", "overlap/WGA total" ), col=c( 'green'
 ## FIGURE 4 (ORIG FIGURE 5 and 6): Number of putative SNPs removed by each of six filters ##
 ############################################################################################
 
-# pdf( "../FIGURE_PDFS/Figure4.pdf", width=6, height=5 )
+#pdf( "../FIGURE_PDFS/Figure4.pdf", width=6, height=6 )
 par(bty='n', mfrow=c(2,1), mar=c(5.1,4.1,1.1,0))
 boxplot( data5[,c(5,7,4,3,6,8)], log='y', ylim=c(1,100000), xaxt='n', cex.lab=0.7, pch=20, ylab="Number putative SNVs filtered" )
 axis( 1, at=1:6, padj=0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
 boxplot( data6[,c(9,13,7,5,11,15)], ylim=c(0,1), xaxt='n', cex.lab=0.7, ylab="Percentage overlap filtered", pch=20 )
 axis( 1, at=1:6, padj = 0.8, cex.axis=0.7, labels=c( "VAQ", "LOH", "10bp-SNV", "10bp-INDEL", "dbSNP", "<10%" ) )
-# dev.off()
+#dev.off()
 
 ##############################################################
 ## FIGURE 5 IS MADE IN THE SCRIPT Jacquard.R IN THIS FOLDER ##

### 2015-11-24 PAPER/paper_mc.tex
Matt's edits to the figure legends.

--- 2015-11-23 PAPER/paper.tex
@@
 %}
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} For each patient (represented by one point), we compared the number of SNVs called in the WGS sample ($y$-axis) versus the number of SNVs called in the WGA sample ($x$-axis).
+\caption{\textbf{Number of putative SNVs in WGS versus WGA before filtering.} For each patient (represented by one point), we compared the number of SNVs called in the WGS sample ($y$-axis) versus the number of SNVs called in the WGA sample ($x$-axis).
 As expected, the number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).
-The line shown is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
+The line shown is $y=x$ (WGS = WGA); points falling below the line support the hypothesis that WGA introduces additional spurious variants in a sample.
 }
@@
 %}
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} To test the hypothesis that samples with a greater number of SNVs also have a greater number of errors, and consequently a smaller percentage of SNVs recovered in the overlap between replicates, we compared the percentage of putative SNVs recovered in the overlap between replicates for a sample ($y$-axis) versus the number of mutations in that sample ($x$-axis).
-We found no correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$), suggesting that higher numbers of SNVs in a sample may not be only due to errors.
+\caption{\textbf{Total number of SNVs per sample does not correlate with the number of putative SNVs recovered after filtering.} To test the hypothesis that samples with a greater number of SNVs also have a greater number of errors that impede recovery of putative SNVs, we compared the percentage of putative SNVs recovered in the overlap between replicates for a sample ($y$-axis) to the number of total number of mutations in that sample ($x$-axis).
+We found no correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$), suggesting that higher total SNV counts in a sample may due to additional factors other than errors.
 We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
@@
 %}
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} For each filter (names given on the $x$-axis, with detailed description in Table 1), we looked at the number of mutations removed by a given filter in a given sample on a log scale \textit{(above)}, and the percentage of the WGS--WGA overlap removed by the filter \textit{(below)}, per sample.
+\caption{\textbf{Effect of individual filters on the overlap between replicates.} For each filter (names given on the $x$-axis, with detailed description in Table 1), we looked at the number of mutations removed by a given filter in a given sample \textit{(above)}, and the percentage of the WGS--WGA overlap removed by the filter \textit{(below)}, per sample.
 The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
@@
 The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data.
+SNV counts are plotted on a log scale.
 }
@@
 %}
-\caption{\textbf{Independent performance of individual filters of the difference between replicates.} For each sample, we compared the size of the difference between replicates (the number of SNVs found in only one of WGS or WGA) to the size of the overlap between replicates (the number of SNVs recovered from both the WGA and WGS replicates).
+\caption{\textbf{Effect of individual filters on the difference between replicates.} For each sample, we compared the size of the difference between replicates (the number of SNVs found in only one of WGS or WGA) to the size of the overlap between replicates (the number of SNVs recovered from both the WGA and WGS replicates).
 \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$).
@@
 %}
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} Across samples, as the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
+\caption{ \textbf{LOH and VAQ filters remove almost completely different sets of mutations and cover most of the sample.} Across samples, as the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
 The two are almost (but not quite) perfect inverses: putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}

### 2015-11-24 PAPER/paper.tex
made Matt's changes to figure titles and legends

@@
 We found significant overlap (around 50\%) between technical replicates, but also significant differences.
-As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates.
-We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
-Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
+As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative SNVs to the sample, so that on average these replicates had (i) more putative SNVs, and (ii) a smaller percentage overlap between replicates.
+We found that the number of SNVs in the WGS replicates varied by orders of magnitude, from 110 to 8,192.
+Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative SNVs, suggesting real variation in mutation frequency (the absolute number of SNVs found in a tumor) between samples from the same tumor.
 This result may be in part due to known mutational hotspots in some tumors \citep{Karen}.
@@
 
-After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates.
+After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative SNVs in each VCF, and to calculate the overlap between technical replicates.
 We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository).
@@
 %}
-\caption{\textbf{Number of putative SNVs in WGS versus WGA, as called by SomaticSniper before filtering.} For each patient (represented by one point), we compared the number of SNVs called in the WGS sample ($y$-axis) versus the number of SNVs called in the WGA sample ($x$-axis).
+\caption{\textbf{Number of putative SNVs in WGS versus WGA before filtering.} For each patient (represented by one point), we compared the number of SNVs called in the WGS sample ($y$-axis) versus the number of SNVs called in the WGA sample ($x$-axis).
 As expected, the number of mutations found in one replicate correlates with the number of mutations found in the other replicate (Spearman $\rho=0.42$, $S=16142$, $P=0.002$).
-The line shown is $y=x$, so points falling below the line agree with the hypothesis that an additional amplification step produces more sequencing errors in a sample.
+The line shown is $y=x$ (WGS=WGA); points falling below the line support the hypothesis that WGA introduces additional spurious variants in a sample.
 }
@@
 %}
-\caption{\textbf{Number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} To test the hypothesis that samples with a greater number of SNVs also have a greater number of errors, and consequently a smaller percentage of SNVs recovered in the overlap between replicates, we compared the percentage of putative SNVs recovered in the overlap between replicates for a sample ($y$-axis) versus the number of mutations in that sample ($x$-axis).
-We found no correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$), suggesting that higher numbers of SNVs in a sample may not be only due to errors.
+\caption{\textbf{Total number of putative SNVs per sample does not correlate with the number of putative SNVs recoverable in both replicates.} To test the hypothesis that samples with a greater number of SNVs also have a greater number of errors, which impede accurate recovery of real variation, we compared the percentage of putative SNVs recovered in the overlap between replicates for a sample ($y$-axis) to the total number of SNVs called in that sample ($x$-axis).
+We found no correlation (Spearman $\rho=0.05$, $S=29268$, $P=0.68$), suggesting that higher total SNV counts in a sample may be due to additional factors other than errors.
 We calculated the percent overlap in two ways: with reference to the total number of putative SNVs in the WGS sample (red) and with reference to the total number of putative SNVs in the WGA sample (black).
@@
 %}
-\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} For each filter (names given on the $x$-axis, with detailed description in Table 1), we looked at the number of mutations removed by a given filter in a given sample on a log scale \textit{(above)}, and the percentage of the WGS--WGA overlap removed by the filter \textit{(below)}, per sample.
+\caption{\textbf{Effect of individual filters on the overlap between replicates.} For each filter (names given on the $x$-axis, with detailed description in Table 1), we looked at the number of mutations removed by a given filter in a given sample on a log scale \textit{(above)}, and the percentage of the WGS--WGA overlap removed by the filter \textit{(below)}, per sample.
 The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap.
@@
 %}
-\caption{ \textbf{Replicate SNVs filtered out by LOH and VAQ are almost completely non-overlapping and cover most of the sample.} Across samples, as the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
-The two are almost (but not quite) perfect inverses: putative SNPs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
+\caption{ \textbf{LOH and VAQ filters remove almost completely different sets of mutations and cover most of the sample.} Across samples, as the length of the overlap between WGS and WGA increases ($x$-axis),  \textit{(above)} the percentage of the overlap filtered out by LOH increases, and \textit{(below)} the percentage of the overlap filtered out by VAQ decreases.
+The two are almost (but not quite) perfect inverses: putative SNVs filtered by LOH and VAQ cover almost the entire overlap, and together sum to nearly 100\% of the overlap.}
 \label{fig:LOH-VAQ_all}
